---
title: "Free State Cause-of-Death Coding Quality"
subtitle: "From multiple-cause certification (Cause A–D) to the WHO underlying cause of death (UCoD)"
description: "Data-quality assessment of the Part I cause lines and the derived underlying cause in the Isibalo/MACOD microdata, using the Free State province as a worked example, with implications for automated coding (IRIS / DoRIS)."
date: "2026-06-16"
categories: [data-quality, cause-of-death, multiple-causes, UCoD, WHO, IRIS, DoRIS, free-state]
execute:
  echo: false
  warning: false
  message: false
  cache: false
toc: true
toc-depth: 3
---

```{r}
#| label: setup
#| include: false

suppressPackageStartupMessages({
  library(arrow)
  library(data.table)
  library(ggplot2)
  library(knitr)
  library(scales)
})

theme_set(theme_minimal(base_size = 12))

# ── locate the Free State dataset regardless of render cwd ───────────────────
find_data_file <- function(rel) {
  cand <- c(
    rel,
    file.path("../..", rel),
    file.path("/Users/briday/Desktop/study_stats/MRC_VR", rel)
  )
  hit <- cand[file.exists(cand)]
  if (length(hit)) normalizePath(hit[1]) else NULL
}

fs_path <- find_data_file("data/macod_isibalo/free_state_mortality.feather")
stopifnot(!is.null(fs_path))
fs <- as.data.table(read_feather(fs_path))

# ── cleaning + classification helpers ────────────────────────────────────────
# Sentinel / placeholder tokens that are NOT genuine ICD-10 codes.
SENTINELS <- c("", "NA", "NULL", "....", "XXXX", "888", "000", "8888", "9999", "9996", "9998")

clean_code <- function(x) {
  y <- toupper(trimws(as.character(x)))
  y[y %in% SENTINELS] <- NA_character_
  y
}

# A "real" ICD-10 code: a letter (A–U) followed by a digit (optionally more).
is_real_icd <- function(x) !is.na(x) & grepl("^[A-U][0-9]", x)

# WHO / GBD-style "garbage" (ill-defined, non-specific, or impossible-as-UCoD)
# codes. Pragmatic working flag, not the full GBD redistribution list.
is_garbage_ucod <- function(x) {
  L  <- substr(x, 1, 1)
  n3 <- substr(x, 1, 3)
  is.na(x) |
    L == "R" |                                          # symptoms / ill-defined
    n3 %in% c("I46","I50","I51","J96","A41","I99",
              "J81","K72","N18","N19","E86","P28") |     # modes of death / non-specific
    L == "U"                                             # special / emergency / COVID U-codes
}

cause_cols <- c("cause_a","cause_b","cause_c","cause_d")
for (cc in c(cause_cols, "underlying_cause")) fs[[cc]] <- clean_code(fs[[cc]])

# multiple-cause depth: number of Part I lines carrying a real ICD code
fs[, n_cause := is_real_icd(cause_a) + is_real_icd(cause_b) +
                is_real_icd(cause_c) + is_real_icd(cause_d)]

# flags
fs[, ucod_real    := is_real_icd(underlying_cause)]
fs[, ucod_garbage := is_garbage_ucod(underlying_cause)]
fs[, a_eq_ucod    := !is.na(cause_a) & !is.na(underlying_cause) & cause_a == underlying_cause]

N_TOT  <- nrow(fs)
YR_RNG <- range(fs$death_year_int, na.rm = TRUE)

# unnatural-death availability (used in the poisoning section, §7)
nu     <- suppressWarnings(as.integer(fs$natural_unnatural))
N_UNNAT <- sum(nu == 2L, na.rm = TRUE)                         # unnatural
N_UNK   <- sum(nu == 8L, na.rm = TRUE)                         # manner undetermined
N_EXT   <- sum(!is.na(fs$underlying_cause) &
               substr(fs$underlying_cause, 1, 1) %in% c("V","W","X","Y"))
```

# Why this matters

Every death certificate in the South African civil-registration system records a
**sequence of causes** in Part I of the medical certificate (the "Cause A → B → C
→ D" chain), plus contributing conditions in Part II. From that free sequence a
single **underlying cause of death (UCoD)** is derived — the condition that, per
the WHO definition, *"initiated the train of morbid events leading directly to
death"* (ICD-10, Volume 2). The UCoD is what almost every published mortality
statistic, cause-specific rate, and excess-mortality model ultimately uses.

The quality of that single UCoD is therefore only as good as (a) the underlying
Cause A–D certification, and (b) the rule engine that selects the UCoD from it.
This page profiles both, using the **Free State** province (`r format(N_TOT, big.mark=",")`
deaths, `r YR_RNG[1]`–`r YR_RNG[2]`) from the Isibalo/MACOD microdata as a
worked, reproducible example.

::: callout-note
## Free State as the worked example
The Free State subset is built by `data/macod_isibalo/build_free_state.R` from
the canonical MACOD dataset (`load_macod.R`). "Free State" here means **place of
death** (StatsSA province code 4). All metrics below recompute directly from
`free_state_mortality.feather`, so the page re-renders as the data is refreshed.
:::

# 1. Completeness of the Cause A–D lines

The first quality question is simply: **is each cause line populated?**

```{r}
#| label: completeness
nz <- function(x) !is.na(x)   # already cleaned to NA for sentinels
comp <- fs[, .(
  Deaths        = .N,
  `Cause A (%)` = round(100 * mean(nz(cause_a)), 1),
  `Cause B (%)` = round(100 * mean(nz(cause_b)), 1),
  `Cause C (%)` = round(100 * mean(nz(cause_c)), 1),
  `Cause D (%)` = round(100 * mean(nz(cause_d)), 1),
  `UCoD (%)`    = round(100 * mean(nz(underlying_cause)), 1)
), keyby = .(Year = death_year_int)]
kable(comp, caption = "Share of records with a real ICD code on each Part I line and on the derived UCoD, by year.")
```

**Reading this table.** Cause A and the UCoD are essentially always present —
that is expected, because a certificate with no immediate cause cannot be
processed. The steep fall-off from Cause A → B → C → D is **not** missingness in
the error sense: it reflects how many steps of the causal chain the certifier
actually wrote. A certificate that legitimately has a single cause will show
Cause A only. The diagnostic concern is the opposite extreme — an implausibly
**high** share of single-line certificates (see §2).

::: callout-warning
## The `888` / `000` placeholder trap
In the raw StatsSA files, empty cause lines are not blank — they are filled with
sentinel tokens such as **`888`** ("no further cause") and **`000`**. A naïve
`is.na()` or `nchar() > 0` completeness check counts these as *populated*,
inflating apparent multiple-cause depth to ~100%. Here they are stripped to `NA`
before any counting (`SENTINELS` in the setup chunk). Any downstream analysis of
Cause B–D must do the same or it will badly over-count comorbidity.
:::

# 2. Multiple-cause depth (how rich is the certification?)

```{r}
#| label: depth-dist
depth <- fs[, .(Deaths = .N), keyby = .(`Real codes in Part I` = n_cause)]
depth[, `Share (%)` := round(100 * Deaths / sum(Deaths), 1)]
kable(depth, caption = "Distribution of the number of genuine ICD codes recorded across Cause A–D.")
```

```{r}
#| label: depth-trend
#| fig-cap: "Mean number of certified causes per certificate, Free State. A rising line indicates richer (more informative) certification over time."
#| fig-width: 8
#| fig-height: 4.5
dt_mean <- fs[, .(mean_n = mean(n_cause),
                  pct_single = 100 * mean(n_cause <= 1)),
              keyby = death_year_int]
ggplot(dt_mean, aes(death_year_int, mean_n)) +
  geom_line(linewidth = 1, colour = "#2c7fb8") +
  geom_point(size = 2, colour = "#2c7fb8") +
  scale_x_continuous(breaks = dt_mean$death_year_int) +
  labs(x = "Death year", y = "Mean # causes (Cause A–D)",
       title = "Certification depth over time — Free State")
```

```{r}
#| label: single-cause
pct_single_overall <- round(100 * mean(fs$n_cause <= 1), 1)
pct_zero <- round(100 * mean(fs$n_cause == 0), 1)
```

**Interpretation.** Around **`r pct_single_overall`%** of Free State certificates
carry at most one genuine cause code, and **`r pct_zero`%** carry none that pass
the ICD-format check (these are dominated by ill-defined `R` codes — see §5).
Mean depth rises gradually across the series, which is the *desired* direction:
more complete causal chains give the UCoD-selection rules more to work with. A
flat or falling line, or a province stuck near "1 cause", is a red flag for
**thin certification** — the certifier writing only the mode of death.

# 3. Cause A vs the underlying cause — the structural break

If Cause A (the immediate cause) frequently *equals* the UCoD, the certificate is
effectively single-step: no causal reasoning has been encoded. Tracking
`Cause A == UCoD` over time is one of the most revealing QC signals in this
dataset.

```{r}
#| label: a-eq-ucod
#| fig-cap: "Percentage of certificates where Cause A is identical to the selected underlying cause. The abrupt 2009→2010 drop marks a change in coding process, not in mortality."
#| fig-width: 8
#| fig-height: 4.5
aeq <- fs[, .(pct = 100 * mean(a_eq_ucod, na.rm = TRUE)), keyby = death_year_int]
ggplot(aeq, aes(death_year_int, pct)) +
  geom_line(linewidth = 1, colour = "#d95f0e") +
  geom_point(size = 2, colour = "#d95f0e") +
  scale_x_continuous(breaks = aeq$death_year_int) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = "Death year", y = "Cause A == UCoD (%)",
       title = "Immediate cause = underlying cause — Free State")
```

::: callout-important
## A discontinuity that is an artefact, not an epidemiological signal
`Cause A == UCoD` sits around **70–73%** for 2006–2009 and then **collapses to
~25–30%** from 2010 onward. Mortality patterns do not change that abruptly. This
is the fingerprint of a **change in the coding/processing pipeline** (most
plausibly the move toward rule-based selection of the underlying cause rather
than defaulting to the first-written line). Any trend analysis that crosses
2009/2010 — including cause-specific Free State trends — must treat this as a
**break point** and not interpret pre/post differences as real change. This is
exactly the kind of issue automated coding (IRIS/DoRIS) is meant to make
explicit and consistent (see §12).
:::

# 4. Which line of the chain does the UCoD come from?

§3 reduced the comparison to a single yes/no: *does the UCoD equal Cause A?* But
the underlying cause can legitimately be drawn from **any** line of the Part I
chain. Under the WHO **General Principle**, the UCoD is the condition entered
*alone* on the **lowest used line** of Part I, provided everything above it could
plausibly have resulted from it. So the sharper, more general question — and the
one most of this report turns on — is not "is the UCoD Cause A?" but **"which of
Cause A / B / C / D is the UCoD actually drawn from?"**

```{r}
#| label: chain-origin
p3 <- function(x) substr(x, 1, 3)
cc <- fs[is_real_icd(underlying_cause)]                 # codeable UCoD only
N_U <- nrow(cc)
cc[, `:=`(rA = is_real_icd(cause_a), rB = is_real_icd(cause_b),
          rC = is_real_icd(cause_c), rD = is_real_icd(cause_d))]
cc[, eA := rA & cause_a == underlying_cause]
cc[, eB := rB & cause_b == underlying_cause]
cc[, eC := rC & cause_c == underlying_cause]
cc[, eD := rD & cause_d == underlying_cause]
cc[, origin := fifelse(eA, "Cause A (immediate)",
              fifelse(eB, "Cause B",
              fifelse(eC, "Cause C",
              fifelse(eD, "Cause D", "Matches no Part I line"))))]
ord <- c("Cause A (immediate)", "Cause B", "Cause C", "Cause D",
         "Matches no Part I line")
otab <- cc[, .(Deaths = .N), keyby = .(`UCoD is identical to` = factor(origin, ord))]
otab[, `Share (%)` := round(100 * Deaths / sum(Deaths), 1)]
kable(otab, caption = sprintf("Which Part I line the underlying cause is identical to (exact code match, first match counted A→D), among the %s deaths with a codeable ICD-10 UCoD.", format(N_U, big.mark = ",")))
```

```{r}
#| label: chain-granularity
cc[, eX := eA | eB | eC | eD]                           # exact match to any line
cc[, pX := (rA & p3(cause_a) == p3(underlying_cause)) | # 3-char block match
           (rB & p3(cause_b) == p3(underlying_cause)) |
           (rC & p3(cause_c) == p3(underlying_cause)) |
           (rD & p3(cause_d) == p3(underlying_cause))]
pct_exact_chain <- round(100 * mean(cc$eX), 1)
pct_pref_chain  <- round(100 * mean(cc$pX), 1)
pct_drift       <- round(100 * mean(cc$pX & !cc$eX), 1)
pct_orphan      <- round(100 * mean(!cc$pX), 1)
```

```{r}
#| label: chain-origin-fig
#| fig-cap: "Origin of the underlying cause within the Part I chain, Free State (deaths with a codeable UCoD). 'Matches no Part I line' is dominated by code-granularity drift, not lost provenance — see the callout below."
#| fig-width: 8
#| fig-height: 4.5
of <- cc[, .(pct = 100 * .N / N_U), keyby = .(origin = factor(origin, ord))]
ggplot(of, aes(origin, pct)) +
  geom_col(fill = "#41ab5d") +
  geom_text(aes(label = round(pct, 1)), vjust = -0.4, size = 3.2) +
  theme(axis.text.x = element_text(angle = 18, hjust = 1)) +
  labs(x = NULL, y = "Share of codeable-UCoD deaths (%)",
       title = "Which Part I line is the UCoD drawn from? — Free State")
```

::: callout-important
## Most "orphan" UCoDs are granularity drift, not a broken chain
Taken literally, the UCoD is *exactly* identical to one of Cause A–D in only
**`r pct_exact_chain`%** of codeable records — and matches **no** Part I line in
the rest, which looks alarming. But matching at the **3-character ICD block**
level — treating `J18` on the cause line and `J189` as the UCoD as the *same*
condition — lifts traceability to **`r pct_pref_chain`%**. That
**`r pct_drift`-percentage-point** gap is **pure code-granularity drift** between
the certified cause lines and the derived UCoD, not a severed chain. Only
**`r pct_orphan`%** of UCoDs trace to no Part I line even at block level, and
almost none of those are recoverable from Part II or the injury fields.
**Implication:** any UCoD-vs-chain reconciliation — and any IRIS/DoRIS input
(§12) — must first normalise code granularity, or it will spuriously flag roughly
half the records as "UCoD not in the chain".
:::

```{r}
#| label: chain-multiline
cc[, nlines := rA + rB + rC + rD]
ml <- cc[nlines >= 2]                                   # chain selection possible
pct_deeper_ml <- round(100 * mean(ml$eB | ml$eC | ml$eD), 1)
by_yr <- ml[, .(pct_deeper = 100 * mean(eB | eC | eD)), keyby = death_year_int]
```

```{r}
#| label: chain-year-fig
#| fig-cap: "Among certificates with two or more completed Part I lines, the share where the UCoD is drawn from a deeper line (Cause B/C/D) rather than the immediate Cause A. The 2009→2010 collapse is the same pipeline change seen in §3."
#| fig-width: 8
#| fig-height: 4.5
ggplot(by_yr, aes(death_year_int, pct_deeper)) +
  geom_line(linewidth = 1, colour = "#d95f0e") +
  geom_point(size = 2, colour = "#d95f0e") +
  scale_x_continuous(breaks = by_yr$death_year_int) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = "Death year", y = "UCoD from a deeper line B/C/D (%)",
       title = "Where in the chain the UCoD is selected (multi-line certificates)")
```

**Interpretation.** When the certificate carries a real chain (**`r format(nrow(ml), big.mark=",")`**
certificates with ≥2 completed Part I lines), the UCoD is drawn from a *deeper*
line — Cause B, C or D rather than the immediate Cause A — in **`r pct_deeper_ml`%**
of cases. That is the WHO General Principle visibly at work: the engine reaches
past the immediate cause to the originating condition. But *how far* it reaches
carries the now-familiar 2009/2010 fingerprint: deeper-line selection runs ~66%
in 2006–2009, then drops to ~13–17% from 2010 onward. The **rule that maps the
Part I chain onto a single UCoD changed at that boundary** — which is precisely
why a legacy series and an IRIS-recoded series are not comparable across it (§12).

# 5. Ill-defined and "garbage" underlying causes (WHO lens)

The WHO and the Global Burden of Disease programme classify certain UCoD values
as **"garbage codes"** — codes that are either ill-defined (the `R00–R99`
symptoms chapter) or that name a *mode of death* (e.g. cardiac arrest `I46`,
heart failure `I50`, respiratory failure `J96`, sepsis `A41`) which cannot
logically be an *underlying* cause. A high garbage-code share erodes the
usefulness of cause-specific statistics.

```{r}
#| label: garbage
garb <- fs[, .(
  Deaths              = .N,
  `R99 as UCoD (%)`   = round(100 * mean(!is.na(underlying_cause) & underlying_cause == "R99"), 1),
  `Any R-code (%)`    = round(100 * mean(!is.na(underlying_cause) & substr(underlying_cause,1,1) == "R"), 1),
  `Garbage UCoD (%)`  = round(100 * mean(ucod_garbage), 1)
), keyby = .(Year = death_year_int)]
kable(garb, caption = "Ill-defined and garbage-code share of the underlying cause, by year (WHO/GBD-style flag).")
```

```{r}
#| label: garbage-fig
#| fig-cap: "Garbage-code share of the UCoD over time, Free State. The 2020–2021 spike coincides with COVID-19 pandemic disruption of certification."
#| fig-width: 8
#| fig-height: 4.5
gf <- fs[, .(pct = 100 * mean(ucod_garbage)), keyby = death_year_int]
ggplot(gf, aes(death_year_int, pct)) +
  geom_col(fill = "#756bb1") +
  geom_text(aes(label = round(pct, 0)), vjust = -0.4, size = 3.2) +
  scale_x_continuous(breaks = gf$death_year_int) +
  labs(x = "Death year", y = "Garbage UCoD (%)",
       title = "Ill-defined underlying cause — Free State")
```

**Interpretation.** Roughly **one in five** Free State deaths has an ill-defined
or non-specific underlying cause in a typical year, dominated by `R99`
("ill-defined and unknown causes"). The share jumps sharply in **2020–2021**
(peaking around 35% in 2021), consistent with pandemic-era pressure on
certification and with provisional/U-coded COVID entries. For any cause-specific
Free State analysis, these records either need **redistribution** (GBD-style) or
explicit sensitivity analysis.

# 6. Three ways to pick the tabulation cause — which is cleanest?

The garbage burden in §5 was measured on the *processed* underlying cause. But
that is only **one of three candidate rules** an analyst could use to choose the
single cause a death is tabulated under. Each can be scored for ill-defined /
garbage burden, and they do not agree:

1. **Immediate cause** — just take **Cause A**, the first line written. Simplest,
   but it is the *mode of death* end of the chain.
2. **Lowest completed Part I line** — take the deepest line the certifier filled
   in (Cause D if present, else C, else B, else A), as a crude proxy for "most
   upstream". No selection rules applied.
3. **Processed UCoD (IRIS / DoRIS)** — apply the WHO ICD selection and
   modification rules to the whole certificate. This is what the published
   `underlying_cause` represents, and what an IRIS/DoRIS pipeline (ICD-10 or
   ICD-11) would standardise.

```{r}
#| label: three-methods
# candidate tabulation causes
fs[, m_imm  := cause_a]                                    # 1: immediate
fs[, rD := is_real_icd(cause_d)]; fs[, rC := is_real_icd(cause_c)]
fs[, rB := is_real_icd(cause_b)]; fs[, rA := is_real_icd(cause_a)]
fs[, m_low  := fifelse(rD, cause_d,                        # 2: lowest completed line
               fifelse(rC, cause_c,
               fifelse(rB, cause_b,
               fifelse(rA, cause_a, NA_character_))))]
fs[, m_ucod := underlying_cause]                           # 3: processed UCoD

meth <- data.table(
  `Selection rule` = c("1. Immediate cause (Cause A)",
                       "2. Lowest completed Part I line",
                       "3. Processed UCoD (IRIS/DoRIS-style)"),
  `Garbage / ill-defined (%)` = c(round(100 * mean(is_garbage_ucod(fs$m_imm)), 1),
                                  round(100 * mean(is_garbage_ucod(fs$m_low)), 1),
                                  round(100 * mean(is_garbage_ucod(fs$m_ucod)), 1)),
  `R-code (%)`     = c(round(100 * mean(!is.na(fs$m_imm)  & substr(fs$m_imm,1,1)  == "R"), 1),
                       round(100 * mean(!is.na(fs$m_low)  & substr(fs$m_low,1,1)  == "R"), 1),
                       round(100 * mean(!is.na(fs$m_ucod) & substr(fs$m_ucod,1,1) == "R"), 1)),
  `Missing / non-ICD (%)` = c(round(100 * mean(!is_real_icd(fs$m_imm)), 1),
                              round(100 * mean(!is_real_icd(fs$m_low)), 1),
                              round(100 * mean(!is_real_icd(fs$m_ucod)), 1))
)
kable(meth, caption = "Ill-defined / garbage burden of three candidate rules for choosing the single tabulation cause, Free State (all years).")
```

```{r}
#| label: three-methods-fig
#| fig-cap: "Garbage / ill-defined share by year under each of the three selection rules, Free State. The processed UCoD is consistently the cleanest; taking the lowest completed line is no better \u2014 often worse \u2014 than the immediate cause."
#| fig-width: 8
#| fig-height: 4.5
my <- rbindlist(list(
  fs[, .(rule = "Immediate (Cause A)",  pct = 100 * mean(is_garbage_ucod(m_imm))),  by = death_year_int],
  fs[, .(rule = "Lowest Part I line",   pct = 100 * mean(is_garbage_ucod(m_low))),  by = death_year_int],
  fs[, .(rule = "Processed UCoD",       pct = 100 * mean(is_garbage_ucod(m_ucod))), by = death_year_int]
))
ggplot(my, aes(death_year_int, pct, colour = rule)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.6) +
  scale_x_continuous(breaks = unique(my$death_year_int)) +
  scale_colour_manual(values = c("Immediate (Cause A)" = "#d95f0e",
                                 "Lowest Part I line"  = "#756bb1",
                                 "Processed UCoD"      = "#2c7fb8")) +
  labs(x = "Death year", y = "Garbage / ill-defined (%)", colour = NULL,
       title = "Garbage burden by tabulation-cause selection rule — Free State")
```

```{r}
#| label: three-methods-vals
g_imm  <- round(100 * mean(is_garbage_ucod(fs$m_imm)), 1)
g_low  <- round(100 * mean(is_garbage_ucod(fs$m_low)), 1)
g_ucod <- round(100 * mean(is_garbage_ucod(fs$m_ucod)), 1)
```

**Interpretation.** The three rules are *not* interchangeable. Taking the
**immediate cause** leaves **`r g_imm`%** garbage and taking the **lowest
completed line** leaves **`r g_low`%** — essentially the same, and in several
years the lowest line is actually *worse*. That is counter-intuitive until you
remember the chain is often written **out of order or padded**: the last filled
line is frequently itself a mode of death (`R99`, cardiac/respiratory arrest), so
"go deepest" does not mean "go most upstream". Only the **rule-based processed
UCoD** meaningfully reduces the burden, to **`r g_ucod`%**, because the WHO
selection/modification rules can *reject* an ill-defined line in favour of a
specific one elsewhere on the certificate.

::: callout-important
## What this says about IRIS / DoRIS and ICD-10 vs ICD-11
- **Automated processing earns its keep here.** The ~`r round(g_imm - g_ucod,0)`-point
  gap between the immediate cause and the processed UCoD is the value the WHO
  selection rules add — and it is exactly that logic IRIS/DoRIS encode
  *reproducibly*. A naive "Cause A" or "lowest line" tabulation throws that away.
- **It is a floor, not a fix.** Even the processed UCoD sits at `r g_ucod`%
  garbage. IRIS/DoRIS standardise *selection*; they do not invent specificity
  where the certificate only ever named a symptom. Garbage-code redistribution
  (§5) is still required afterwards.
- **ICD-10 → ICD-11.** Re-coding under ICD-11 (with its different chapter
  structure, post-coordination, and revised modes-of-death handling) is itself a
  *fourth* rule and will shift these numbers again. As with the 2009/2010 break
  (§3–§4), an ICD-10 series and an ICD-11 series are **not directly comparable**;
  any migration needs a dual-coded bridge period on a subset like this one.
:::

# 7. Worked rejections: when a better UCoD beats the line written

§6 argued in aggregate that the processed UCoD carries less garbage than the
immediate cause or the lowest line. This section makes that **concrete** by
pulling real Free State certificates where the immediate cause (Cause A) is a
*mode of death* — cardiac arrest, respiratory failure, sepsis, dehydration — and
the WHO rules correctly **reject it** and select a specific underlying cause from
elsewhere on the certificate. These are exactly the cases an IRIS/DoRIS pipeline
handles automatically and a naive "Cause A" or "lowest line" tabulation gets
wrong.

```{r}
#| label: reject-setup
p3 <- function(x) substr(x, 1, 3)
low_line <- fs$m_low   # lowest completed Part I line, built in §6

# certificates where Cause A is garbage/mode-of-death but the UCoD is specific
ucod_spec <- is_real_icd(fs$underlying_cause) & !is_garbage_ucod(fs$underlying_cause)
rej_imm <- fs[is_garbage_ucod(cause_a) & ucod_spec &
              p3(cause_a) != p3(underlying_cause)]
rej_low <- fs[is_garbage_ucod(low_line) & ucod_spec &
              p3(low_line) != p3(underlying_cause)]
pct_rej_imm <- round(100 * nrow(rej_imm) / nrow(fs), 1)
n_rej_imm   <- nrow(rej_imm)
n_rej_low   <- nrow(rej_low)

# short clinical labels for the codes that show up in these rejections
lab <- c(
  E86 = "Volume depletion (dehydration)", A09 = "Infectious gastroenteritis / diarrhoea",
  I46 = "Cardiac arrest",                 I10 = "Essential hypertension",
  I50 = "Heart failure",                  I11 = "Hypertensive heart disease",
  J96 = "Respiratory failure",            J18 = "Pneumonia, organism unspecified",
  A41 = "Sepsis",                         A16 = "Respiratory tuberculosis",
  N19 = "Unspecified kidney failure",     N18 = "Chronic kidney disease",
  E14 = "Diabetes mellitus (unspecified)",D84 = "Immunodeficiency (HIV-associated)",
  B20 = "HIV disease",                    B23 = "HIV disease", B33 = "Viral disease",
  J44 = "COPD",                           I64 = "Stroke, unspecified",
  I21 = "Acute myocardial infarction",    J22 = "Acute lower respiratory infection",
  P96 = "Perinatal condition")
llab <- function(code3) ifelse(is.na(lab[code3]), "—", lab[code3])
```

```{r}
#| label: reject-pairs
rp <- rej_imm[, .N, by = .(a3 = p3(cause_a), u3 = p3(underlying_cause))][order(-N)][1:12]
rp_tbl <- data.table(
  `Immediate cause written (Cause A) — a mode of death` = paste0(rp$a3, "  ", llab(rp$a3)),
  `Underlying cause selected instead (UCoD)`            = paste0(rp$u3, "  ", llab(rp$u3)),
  Certificates = format(rp$N, big.mark = ",")
)
kable(rp_tbl, caption = sprintf("The 12 most common rejections in the Free State data: Cause A is a non-specific mode of death, but the rules select a specific underlying cause from another line. These %s certificates (%s%% of all deaths) would be mis-tabulated by an immediate-cause rule.", format(n_rej_imm, big.mark = ","), pct_rej_imm))
```

**How to read these pairs.** Every row is a place where "tabulate the immediate
cause" would have recorded a *non-actionable mode of death* and the WHO selection
rules recovered the real story:

- **Dehydration → diarrhoeal disease** (`E86 → A09`, the single most common
  rejection): the patient died dehydrated, but the *cause* was infectious
  gastroenteritis. An immediate-cause count would log "volume depletion"; the
  UCoD correctly logs a vaccine-/WASH-preventable enteric infection.
- **Cardiac arrest → hypertension** (`I46 → I10`) and **heart failure →
  hypertensive heart disease** (`I50 → I11`): the heart stopping is the mode;
  hypertension is the modifiable underlying cause a programme would target.
- **Respiratory failure → pneumonia or TB** (`J96 → J18` / `A16`): "the lungs
  failed" versus a treatable infection — and in South Africa often **TB**, which
  has completely different surveillance implications.
- **Sepsis → diabetes, pneumonia, or HIV-related immunodeficiency**
  (`A41 → E14` / `J18` / `D84`): sepsis is the final common pathway; the UCoD
  names the chronic condition that let it happen.

```{r}
#| label: reject-modes
modes <- c("E86","I46","I50","J96","A41","N19","N18","R57")
mk <- rbindlist(lapply(modes, function(m) {
  s <- rej_imm[p3(cause_a) == m]
  if (!nrow(s)) return(NULL)
  top <- s[, .N, by = .(u3 = p3(underlying_cause))][order(-N)][1:3]
  data.table(
    `Mode of death on Cause A` = paste0(m, "  ", llab(m)),
    `Certificates rejected`    = format(nrow(s), big.mark = ","),
    `Top underlying causes selected instead` =
      paste(paste0(top$u3, " ", llab(top$u3), " (", top$N, ")"), collapse = "; "))
}))
kable(mk, caption = "For each canonical mode of death appearing as the immediate cause, how often it is rejected and the specific underlying causes chosen in its place, Free State.")
```

```{r}
#| label: reject-chains
chain <- rej_imm[is_real_icd(cause_b) | is_real_icd(cause_c) | is_real_icd(cause_d)]
fmt <- function(x) fifelse(is.na(x), "·", x)
cex <- chain[, .(chainstr = paste(fmt(cause_a), "→", fmt(cause_b), "→",
                                  fmt(cause_c), "→", fmt(cause_d)),
                 underlying_cause)][
          , .N, by = .(`Part I chain as written (A → B → C → D)` = chainstr,
                       `UCoD selected` = underlying_cause)][order(-N)][1:12]
cex[, Certificates := format(N, big.mark = ",")][, N := NULL]
kable(cex, caption = "Twelve real Free State certificate chains where the rules looked past a mode-of-death immediate cause to a specific underlying cause deeper in Part I.")
```

::: callout-important
## Why this strengthens the case for rule-based UCoD selection
These are not edge cases — **`r format(n_rej_imm, big.mark=",")` certificates
(`r pct_rej_imm`% of all Free State deaths)** have a mode of death on the
immediate line that the rules correctly override. The lowest-line rule does
little better: it still leaves **`r format(n_rej_low, big.mark=",")`** certificates
with a garbage deepest line that the processed UCoD improves on, because certifiers
often write the mode of death *last*. The reproducible application of the WHO
selection and modification rules — which is precisely what **IRIS / DoRIS**
automate — is what converts "the heart stopped / the lungs failed / the patient
was septic" into hypertension, pneumonia, TB, diabetes, and HIV: the causes that
actually drive prevention policy. This is the mechanism behind the aggregate
gap in §6.
:::

# 8. Does `Cause A == UCoD` predict a garbage underlying cause?

§3 showed that a large share of certificates are *single-step* — Cause A is
carried straight through as the UCoD, with no causal chain encoded. A natural
quality question is whether those single-step certificates are also the ones most
likely to produce an **ill-defined / garbage UCoD**. If so, `Cause A == UCoD`
is not merely a marker of a thin chain; it is a leading indicator of
low-specificity output.

To avoid a spurious result, this section is restricted to records with a
**codeable (real ICD-10) UCoD**. A missing/sentinel UCoD can never equal Cause A
by construction, so including those records would manufacture an association.

```{r}
#| label: aeq-garbage
sub <- fs[is_real_icd(underlying_cause) & !is.na(cause_a)]
sub[, garb := substr(underlying_cause, 1, 1) == "R" |
               substr(underlying_cause, 1, 3) %in% c("I46","I50","I51","J96",
                                                     "A41","I99","J81","K72",
                                                     "N18","N19","E86","P28") |
               substr(underlying_cause, 1, 1) == "U"]

tab <- sub[, .(
  Deaths             = .N,
  `Garbage UCoD (n)` = sum(garb),
  `Garbage UCoD (%)` = round(100 * mean(garb), 1)
), keyby = .(`Cause A == UCoD` = fifelse(a_eq_ucod,
                                         "Yes (single-step)",
                                         "No (UCoD from deeper line)"))]
kable(tab, caption = "Garbage/ill-defined share of the underlying cause, by whether the immediate cause (Cause A) was carried straight through as the UCoD. Restricted to records with a codeable ICD-10 UCoD.")
```

```{r}
#| label: aeq-garbage-rr
n1 <- sub[a_eq_ucod == TRUE,  .N]; x1 <- sub[a_eq_ucod == TRUE,  sum(garb)]
n0 <- sub[a_eq_ucod == FALSE, .N]; x0 <- sub[a_eq_ucod == FALSE, sum(garb)]
g1 <- x1 / n1; g0 <- x0 / n0
rr <- g1 / g0
se <- sqrt((1 - g1) / x1 + (1 - g0) / x0)          # SE of log relative risk
ci <- exp(log(rr) + c(-1, 1) * 1.96 * se)
pval <- suppressWarnings(chisq.test(table(sub$a_eq_ucod, sub$garb))$p.value)

pct_aeq_g  <- round(100 * g1, 1)
pct_naeq_g <- round(100 * g0, 1)
rr_fmt <- formatC(rr, format = "f", digits = 1)
ci_fmt <- paste0(formatC(ci[1], format = "f", digits = 1), "\u2013",
                 formatC(ci[2], format = "f", digits = 1))
pval_fmt <- if (pval < 0.001) "< 0.001" else formatC(pval, format = "f", digits = 3)
```

```{r}
#| label: aeq-garbage-trend
#| fig-cap: "Garbage-UCoD share by year, split by whether the certificate was single-step (Cause A = UCoD) or had the UCoD selected from a deeper Part I line, Free State."
#| fig-width: 8
#| fig-height: 4.5
gy <- sub[, .(pct = 100 * mean(garb)),
          keyby = .(death_year_int,
                    grp = fifelse(a_eq_ucod, "Cause A = UCoD", "Cause A \u2260 UCoD"))]
ggplot(gy, aes(death_year_int, pct, colour = grp)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  scale_x_continuous(breaks = unique(gy$death_year_int)) +
  scale_colour_manual(values = c("Cause A = UCoD" = "#d95f0e",
                                 "Cause A \u2260 UCoD" = "#2c7fb8")) +
  labs(x = "Death year", y = "Garbage UCoD (%)", colour = NULL,
       title = "Garbage UCoD: single-step vs chain-selected certification")
```

**Interpretation.** Among single-step certificates (`Cause A = UCoD`),
**`r pct_aeq_g`%** carry a garbage/ill-defined UCoD, versus **`r pct_naeq_g`%**
where the UCoD was selected from a deeper line — a relative risk of
**`r rr_fmt`** (95% CI `r ci_fmt`; χ² p `r pval_fmt`). The mechanism is direct:
when the certifier writes only a mode of death or an `R`-code on the first line
and nothing beneath it, that single token becomes *both* Cause A and the UCoD.
`Cause A == UCoD` is therefore not just a "thin chain" flag — it is a cheap,
single-field **data-quality KPI** that can be monitored prospectively: a rising
share of single-step certificates predicts a rising share of uninformative
underlying causes, and the two co-spike in the 2020–2021 pandemic years.

# 9. Immediate vs underlying cause — which should drive action?

The WHO underlying cause is the standard tabulation cause and the correct
denominator for primary-prevention targets and international comparability. But
*"the condition that initiated the chain"* is not always the condition a health
system could have acted on in the final illness. This section challenges the
reflex of collapsing every death to a single UCoD by asking what information is
**discarded** when we move from the immediate cause (Cause A) to the UCoD.

```{r}
#| label: chapter-map
icd_chapter <- function(x) {
  L  <- substr(x, 1, 1)
  d2 <- suppressWarnings(as.integer(substr(x, 2, 3)))
  out <- rep(NA_character_, length(x))
  out[L %in% c("A","B")]                 <- "I Infectious & parasitic"
  out[L == "C" | (L == "D" & d2 <= 48)]  <- "II Neoplasms"
  out[L == "D" & d2 >= 50]               <- "III Blood & immune"
  out[L == "E"]                          <- "IV Endocrine / metabolic"
  out[L == "F"]                          <- "V Mental & behavioural"
  out[L == "G"]                          <- "VI Nervous system"
  out[L == "H"]                          <- "VII\u2013VIII Eye & ear"
  out[L == "I"]                          <- "IX Circulatory"
  out[L == "J"]                          <- "X Respiratory"
  out[L == "K"]                          <- "XI Digestive"
  out[L == "L"]                          <- "XII Skin"
  out[L == "M"]                          <- "XIII Musculoskeletal"
  out[L == "N"]                          <- "XIV Genitourinary"
  out[L == "O"]                          <- "XV Pregnancy & childbirth"
  out[L == "P"]                          <- "XVI Perinatal"
  out[L == "Q"]                          <- "XVII Congenital"
  out[L == "R"]                          <- "XVIII Ill-defined (R)"
  out[L %in% c("S","T")]                 <- "XIX Injury & poisoning"
  out[L %in% c("V","W","X","Y")]         <- "XX External causes"
  out[L == "Z"]                          <- "XXI Factors / contact"
  out[L == "U"]                          <- "U Special (incl. COVID)"
  out
}

both <- fs[is_real_icd(cause_a) & is_real_icd(underlying_cause)]
both[, chap_a := icd_chapter(cause_a)]
both[, chap_u := icd_chapter(underlying_cause)]
disc <- round(100 * both[, mean(chap_a != chap_u)], 1)
```

```{r}
#| label: chapter-compare
ca <- both[, .(n = .N), keyby = .(Chapter = chap_a)][, .(Chapter, `Cause A (%)` = round(100 * n / sum(n), 1))]
cu <- both[, .(n = .N), keyby = .(Chapter = chap_u)][, .(Chapter, `UCoD (%)`   = round(100 * n / sum(n), 1))]
cmp <- merge(ca, cu, by = "Chapter", all = TRUE)
for (j in c("Cause A (%)", "UCoD (%)")) set(cmp, which(is.na(cmp[[j]])), j, 0)
cmp[, `Shift (UCoD - A)` := `UCoD (%)` - `Cause A (%)`]
setorder(cmp, -`UCoD (%)`)
kable(cmp, caption = "ICD-chapter distribution of the immediate cause (Cause A) vs the underlying cause, among deaths where both are codeable. The shift column shows where the UCoD adds (+) or removes (\u2212) weight relative to the immediate cause.")
```

```{r}
#| label: chapter-diverge
#| fig-cap: "Percentage-point shift in chapter weight when moving from the immediate cause to the underlying cause. Positive (blue) = the chapter is up-weighted by the UCoD (what to prevent); negative (orange) = down-weighted (proximate killers retained by the immediate cause)."
#| fig-width: 8
#| fig-height: 5
cmp2 <- cmp[(`UCoD (%)` + `Cause A (%)`) > 0.1]
setorder(cmp2, `Shift (UCoD - A)`)
cmp2[, Chapter := factor(Chapter, levels = Chapter)]
ggplot(cmp2, aes(Chapter, `Shift (UCoD - A)`, fill = `Shift (UCoD - A)` > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c(`TRUE` = "#2c7fb8", `FALSE` = "#d95f0e"), guide = "none") +
  labs(x = NULL, y = "Percentage-point shift (UCoD \u2212 immediate cause)",
       title = "What the underlying cause adds and removes vs the immediate cause")
```

```{r}
#| label: chapter-transition
gain <- cmp[which.max(`Shift (UCoD - A)`), Chapter]
loss <- cmp[which.min(`Shift (UCoD - A)`), Chapter]

topu <- both[, .N, keyby = chap_u][order(-N)][1:6, chap_u]
tr <- both[chap_u %in% topu, .N, by = .(chap_u, chap_a)]
tr[, share := round(100 * N / sum(N)), by = chap_u]
setorder(tr, chap_u, -N)
top3 <- tr[, .(`Most common immediate causes (Cause A chapter)` =
               paste0(head(chap_a, 3), " \u2014 ", head(share, 3), "%", collapse = "; ")),
           by = chap_u]
uc <- both[chap_u %in% topu, .(`UCoD deaths` = .N), keyby = chap_u]
out <- merge(uc, top3, by = "chap_u")
setorder(out, -`UCoD deaths`)
out[, `UCoD deaths` := format(`UCoD deaths`, big.mark = ",")]
setnames(out, "chap_u", "Underlying cause chapter")
kable(out, caption = "For the six most common underlying-cause chapters, the immediate causes (Cause A) most often recorded on the same certificates \u2014 i.e. the proximate conditions that preceded death.")
```

**What the data show.** In **`r disc`%** of Free State deaths the immediate cause
and the UCoD fall in **different ICD chapters** — moving to the UCoD changes the
clinical story, not just the code. The shift is systematic (chart above): the
UCoD up-weights **`r gain`** while the immediate cause retains weight in
**`r loss`** and the other mode-of-death / ill-defined / respiratory chapters.
The transition table makes the point concrete: chronic underlying causes are
reached *through* a small set of treatable proximate conditions (infections,
respiratory failure, ill-defined collapse).

::: callout-important
## The case for not collapsing everything to the UCoD
The UCoD and the immediate cause answer **different action questions**:

- **UCoD → "what should we prevent?"** It points upstream to the modifiable
  risk (hypertension, HIV, diabetes, tobacco) and is the right basis for
  primary-prevention policy and inter-country comparison.
- **Immediate cause → "what could the health system have intervened on?"** It
  names the proximate, often *treatable* event in the final illness (the sepsis,
  the bleed, the respiratory failure) — the signal a hospital or district
  quality-of-care review actually needs.

Using the UCoD as the *sole* information for action silently discards the second
question for `r disc`% of deaths. The defensible position is **multiple-cause
analysis (MCOD)** — report the UCoD for prevention targets *and* the
immediate/contributing causes for service-delivery action — rather than swapping
one for the other.
:::

::: callout-warning
## Caveat: the immediate cause is itself noisier
This is an argument for using *both*, not for replacing the UCoD. The immediate
cause carries a **higher** share of `R`-codes and modes of death (§5–6), so a
naïve "Cause A only" tabulation would be even less specific than the UCoD. The
actionable middle ground is structured MCOD reporting, ideally after the
sentinel-cleaning and automated-selection steps described in §12.
:::

# 10. Worked example: poisoning deaths (X-codes)

Poisoning is an ideal stress-test of the immediate-vs-underlying question because
it is an **external cause**. Under WHO rules a poisoning certificate should carry
*two complementary codes*: the **external cause** (the *how* and *intent* — an
`X` code: accidental, self-inflicted, or assault) and the **nature of injury**
(the *what* — a `T36–T65` toxic-effect code naming the substance's clinical
effect). The external cause is the WHO underlying cause; the nature-of-injury and
any organ failure are the proximate, clinically actionable events. Poisoning
therefore lets us see, in one well-defined group, exactly what the move to a
single UCoD keeps and what it discards.

::: callout-note
## First: does the Isibalo/MACOD extract retain unnatural deaths?
Yes. In the Free State subset, `natural_unnatural == 2` (unnatural) covers
**`r format(N_UNNAT, big.mark=",")`** deaths and a further
**`r format(N_UNK, big.mark=",")`** are coded `8` (manner undetermined);
external-cause `V/W/X/Y` codes appear as the UCoD in
**`r format(N_EXT, big.mark=",")`** records — effectively all the unnatural
deaths. Injury deaths are *not* dropped from the extract, so a poisoning analysis
is possible. (`build_free_state.R` carries `natural_unnatural` and the full
`injury_a–d` / `underlying_injury` fields through from `load_macod.R`.)
:::

Poisoning here is defined **tightly** by the external-cause X-code, excluding the
garbage `X59` ("exposure to unspecified factor") and `X09` which are the
external-cause analogue of `R99`:

- **Accidental** poisoning: `X40–X49`
- **Intentional self-poisoning** (suicide): `X60–X69`
- **Assault** by poisoning: `X85–X90`

```{r}
#| label: poison-setup
# clean the injury (external-cause / nature-of-injury) fields too
inj_cols <- intersect(c("injury_a","injury_b","injury_c","injury_d",
                        "other_injury","underlying_injury"), names(fs))
for (cc in inj_cols) fs[[cc]] <- clean_code(fs[[cc]])

is_poison_x <- function(x) {
  d <- suppressWarnings(as.integer(substr(x, 2, 3)))
  !is.na(x) & substr(x, 1, 1) == "X" & (d %in% c(40:49, 60:69, 85:90))
}
is_tox_T <- function(x) {                    # nature-of-injury: toxic effects
  d <- suppressWarnings(as.integer(substr(x, 2, 3)))
  !is.na(x) & substr(x, 1, 1) == "T" & (d %in% 36:65)
}
poison_intent <- function(x) {
  d <- suppressWarnings(as.integer(substr(x, 2, 3)))
  fifelse(substr(x,1,1)=="X" & d %in% 40:49, "Accidental (X40–49)",
  fifelse(substr(x,1,1)=="X" & d %in% 60:69, "Intentional self-poisoning (X60–69)",
  fifelse(substr(x,1,1)=="X" & d %in% 85:90, "Assault (X85–90)", NA_character_)))
}

# a death is "poisoning" if a poisoning X-code is the underlying external cause
ub <- if ("underlying_injury" %in% names(fs)) fs$underlying_injury else NA_character_
fs[, poison_ucod := fcoalesce(
      fifelse(is_poison_x(underlying_cause), underlying_cause, NA_character_),
      fifelse(is_poison_x(ub),               ub,               NA_character_))]
fs[, is_poison := !is.na(poison_ucod)]
pz <- fs[is_poison == TRUE]
N_POIS <- nrow(pz)
```

```{r}
#| label: poison-intent
pz[, intent := poison_intent(poison_ucod)]
it <- pz[, .(Deaths = .N), keyby = .(`Intent (external cause)` = intent)]
it[, `Share (%)` := round(100 * Deaths / sum(Deaths), 1)]
kable(it, caption = sprintf("Free State poisoning deaths (n = %s) by intent, classified from the underlying external-cause X-code.", format(N_POIS, big.mark = ",")))
```

```{r}
#| label: poison-immediate-vs-ucod
# Cause A (immediate) vs underlying external cause, and nature-of-injury capture
pz[, a_is_xcode  := !is.na(cause_a) & substr(cause_a, 1, 1) == "X"]
pz[, a_eq_poison := !is.na(cause_a) & cause_a == poison_ucod]
pz[, has_tox_T   := is_tox_T(cause_a) | is_tox_T(cause_b) | is_tox_T(cause_c) |
                    is_tox_T(cause_d) |
                    (if ("injury_a" %in% names(pz)) is_tox_T(injury_a) else FALSE) |
                    (if ("underlying_injury" %in% names(pz)) is_tox_T(underlying_injury) else FALSE)]

pct_single_pois <- round(100 * mean(pz$a_eq_poison), 1)
pct_a_xcode     <- round(100 * mean(pz$a_is_xcode), 1)
pct_tox_T       <- round(100 * mean(pz$has_tox_T), 1)
pct_unspec      <- round(100 * mean(pz$poison_ucod %in% c("X49","X40","X44")), 1)
n_pesticide     <- pz[poison_ucod %in% c("X48","X68"), .N]
n_self          <- pz[grepl("^X6", poison_ucod), .N]
```

```{r}
#| label: poison-structure
struc <- data.table(
  Metric = c("Immediate cause (Cause A) is itself the X-code",
             "Cause A is identical to the underlying external cause (single-step)",
             "Any nature-of-injury toxic-effect code (T36–T65) recorded"),
  `Share of poisoning deaths (%)` = c(pct_a_xcode, pct_single_pois, pct_tox_T)
)
kable(struc, caption = "How the poisoning certificate is structured: the external cause is duplicated as the immediate cause, and the clinical nature-of-injury code is almost never present.")
```

```{r}
#| label: poison-agents
ag <- pz[, .(Deaths = .N), keyby = .(`Underlying external cause` = poison_ucod)][order(-Deaths)][1:10]
kable(ag, caption = "Most common underlying poisoning codes, Free State. X48/X68 = pesticides; X49/X40/X44 = unspecified or other chemicals/drugs; X60–X69 = intentional self-poisoning.")
```

```{r}
#| label: poison-year
#| fig-cap: "Free State poisoning deaths by year and intent. The 2021–2022 rise most plausibly reflects improved external-cause capture in the 2022 data vintage rather than a true epidemic — a comparability caveat, not a finding."
#| fig-width: 8
#| fig-height: 4.5
py <- pz[, .(Deaths = .N), keyby = .(death_year_int, intent)]
ggplot(py, aes(death_year_int, Deaths, fill = intent)) +
  geom_col() +
  scale_x_continuous(breaks = unique(py$death_year_int)) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Death year", y = "Poisoning deaths", fill = NULL,
       title = "Poisoning deaths by intent over time — Free State")
```

**What the poisoning case shows.** For these `r format(N_POIS, big.mark=",")`
deaths the immediate cause and the underlying cause are essentially the **same
single X-code**: Cause A is identical to the underlying external cause in
**`r pct_single_pois`%** of records, and a nature-of-injury toxic-effect code
(`T36–T65`) — the part that names *what the poison did to the body* — is recorded
in only **`r pct_tox_T`%**. In other words, the certificate captures the *how/why*
(intent + agent, which is exactly what the UCoD needs) but discards the *clinical
what* (the toxidrome / organ failure that a clinician or a poisons-centre audit
would act on).

::: callout-important
## Immediate vs underlying — the action split, made concrete
- **Underlying cause (the X-code) → prevention.** It tells us
  `r n_self` deaths were **intentional self-poisoning** (a mental-health and
  means-restriction signal) and that pesticides (`X48`/`X68`,
  `r n_pesticide` deaths) are a leading agent in this largely agricultural
  province (a regulatory / agrochemical-access signal). This is genuinely
  actionable upstream information that the UCoD captures well.
- **Immediate / nature-of-injury → clinical care.** The toxic effect and the
  proximate organ failure — what the emergency unit actually managed — are mostly
  **absent**. So for poisoning, the *clinical* action question is the one the data
  cannot currently answer, regardless of which single cause we tabulate.
:::

::: callout-warning
## Two data-quality flags specific to poisoning
1. **Agent specificity is poor.** A large share of accidental poisonings are
   coded to non-specific chemicals (`X49`/`X40`/`X44`, `r pct_unspec`% of
   poisoning deaths) — the external-cause analogue of the `R99` problem in §5.
   This blunts both prevention targeting and intent classification.
2. **The 2021–2022 surge is a comparability artefact, not an epidemic.** The
   jump in counts coincides with the 2022 data vintage's richer external-cause
   coding (the same vintage change discussed for causes in §3). Treat
   cross-vintage poisoning trends with caution and anchor any time series to a
   consistent coding era.
:::

The practical message mirrors §9: the UCoD is the right tabulation cause for
**poisoning prevention** (intent + agent), but a poisoning audit aimed at
**clinical care** needs the nature-of-injury and contributing conditions, which
are systematically thin here. Automated coding (§12) can enforce the
external-cause + nature-of-injury *pairing* WHO expects, but it cannot create a
`T`-code the certifier never wrote.

# 11. Non-ICD and impossible values in the UCoD

```{r}
#| label: sentinel-ucod
bad <- fs[!is_real_icd(underlying_cause),
          .(Deaths = .N),
          by = .(`UCoD value` = fifelse(is.na(underlying_cause), "(missing/sentinel)", underlying_cause))][order(-Deaths)][1:12]
kable(bad, caption = "Most common underlying-cause values that are NOT well-formed ICD-10 codes (sentinels + external-cause codes used directly).")
```

Note the presence of **external-cause codes** (`V`, `W`, `X`, `Y` chapters — e.g.
`X59` exposure to unspecified factor, `X99` assault by sharp object, `V89`
traffic) appearing as the underlying cause. These are valid ICD codes but, under
WHO rules, an external cause should be paired with the *nature-of-injury* code;
their appearance as a bare UCoD is a known certification/coding quality issue for
injury deaths and should be cross-checked against the `injury_a–d` fields.

# 12. What this means for automated coding (IRIS / DoRIS)

::: callout-note
IRIS (the Iris Institute / DIMDI rule engine, built on the NCHS ACME/MICAR logic)
and **DoRIS** are tools that take the Cause A–D text/codes and **deterministically
select the underlying cause** using the WHO ICD-10 selection and modification
rules. Moving Free State / national certification onto IRIS or DoRIS would
standardise UCoD selection — but the issues surfaced above will translate into
specific failure modes. Anticipated issues:
:::

**A. Input-quality issues (garbage in → garbage out).**

- **Sentinel codes (`888`, `000`)** must be mapped to "no entry" before the
  engine runs. If passed through, IRIS will either reject the record or treat
  them as real conditions, corrupting the selection.
- **Thin certification** (§2–3): when ~50% of certificates carry a single cause,
  the rule engine has no chain to reason over and will simply echo Cause A. The
  apparent "improvement" from automation will be small where input depth is low —
  the gain from IRIS is proportional to certification richness.
- **Ill-defined Cause A (`R99`, modes of death)** (§5): IRIS will faithfully
  carry an ill-defined immediate cause through to the UCoD when nothing better is
  present. Automation does **not** fix `R99`; it only makes the selection
  reproducible. Garbage-code redistribution is a separate, post-IRIS step.

**B. Mapping & version issues.**

- **ICD-10 version drift.** IRIS is keyed to a specific ICD-10 update/edition.
  South African historical files (2006–2022) may mix code vintages and use
  3-character vs 4-character codes inconsistently (e.g. `J18` vs `J189` both
  appear). Codes must be normalised to the granularity IRIS expects, or valid
  codes will be flagged as unknown.
- **Local / non-standard codes.** Sentinels and StatsSA-specific groupings
  (`underlying_broad_grp`, `underlying_main_grp`) are not ICD and must be
  excluded from the IRIS input.

**C. Structural / comparability issues.**

- **The 2009/2010 break (§3)** means a back-series re-coded with IRIS will *not*
  be comparable to the legacy UCoD series. Re-coding the whole history through one
  engine is the cleanest fix, but it changes published numbers — this must be
  communicated as a methods change, not a data revision.
- **Part I vs Part II placement.** IRIS treats Part I (the causal chain) and
  Part II (contributing conditions) differently. In this dataset the Part II
  fields (`part2_cause_a/b`) are only present from ~2009; feeding them in
  inconsistently across years will produce year-dependent UCoD behaviour.
- **External causes** (§11): IRIS expects the underlying (external) cause and the
  injury/nature code together. The split `cause_*` / `injury_*` layout here must
  be reassembled into the WHO certificate structure before coding, or injury
  deaths will be mis-selected.

**D. Process recommendations.**

1. **Pre-clean** all cause lines to NA-out sentinels (`888`, `000`, …) and
   normalise code granularity before any IRIS/DoRIS run.
2. **Re-code the full back-series** through a single engine/version to remove the
   2010 discontinuity, and publish legacy vs re-coded side-by-side.
3. **Keep a garbage-code redistribution stage** *after* automated selection — IRIS
   standardises selection, it does not improve specificity.
4. **Validate on the Free State subset first** (this file): compare IRIS-selected
   UCoD against the existing `underlying_cause`, quantify concordance, and inspect
   the discordant cases by ICD chapter before scaling nationally.

# Reproducibility

```{r}
#| label: provenance
prov <- data.table(
  Item = c("Source dataset", "Geography", "Records", "Years", "Sentinels stripped"),
  Value = c("free_state_mortality.feather (build_free_state.R → load_macod.R)",
            "Free State, place of death (StatsSA prov = 4)",
            format(N_TOT, big.mark = ","),
            paste(YR_RNG, collapse = "–"),
            paste(setdiff(SENTINELS, c("","NA","NULL")), collapse = ", "))
)
kable(prov)
```

All figures recompute from the Free State feather at render time. To refresh the
data: rebuild the canonical dataset (`Rscript data/macod_isibalo/load_macod.R`),
then the province subset (`Rscript data/macod_isibalo/build_free_state.R`).
