---
title: "DHA Access: Comparing Analytical Approaches"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
execute:
  warning: false
  message: false
  echo: false
embed-resources: true
editor: source
---

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(plotly)
library(flextable)
library(dplyr)
library(scales)
library(patchwork)

# All outputs precomputed by ward_level_dha_access_script.R
load("outputs/ward_level_dha_access.RData")
```

# The Problem

Birth and death registration in South Africa requires physical access to a Department of Home Affairs (DHA) office. Death registration happens **only** at DHA offices; birth registration can also happen at a subset of hospitals with **DHA satellite desks**. Long travel distances are a plausible barrier to completeness for both. We compare **four analytical approaches** of increasing rigour to estimate what proportion of the population lives within reasonable reach of a registration node, then apply the most decision-relevant of them to quantify access for **both birth and death registration**.

::: callout-tip
## How to read this document

1.  **Methods** — four approaches to estimating distance/coverage, simplest to most rigorous
2.  **Comparison** — a qualitative feature comparison, and one table of completeness proportions across all four methods, split by birth vs death registration where available
3.  **Chosen Method** — the approach used for reporting, with full results (birth vs death split) and limitations
:::

# Data Sources

-   **Registration nodes**: 2024 DHA offices (all methods); a curated set of hospital-based DHA satellite desks is added for the birth-registration analysis (Method 3 only)
-   **Population estimates**: Stats SA mid-year population estimates (2025, ward/district level); WorldPop 100m satellite-derived grid (2020, Method 4 only)
-   **Spatial boundaries**: District boundaries (Stats SA); Ward boundaries (MDB 2020)

::: panel-tabset
## Population by District

```{r}
#| label: SA-pop-map
SA_pop
```

## Number of Offices by District

```{r}
#| label: SA-offices-map
SA_N_off
```

## Office Locations (Interactive)

```{r}
#| label: office-locations-interactive
plotly::ggplotly(p_with_buffers, tooltip = "text")
```
:::

------------------------------------------------------------------------

# Methods

```{r}
#| label: method4-data
#| include: false

# Read pre-computed outputs from the redistribute_facility_deaths pipeline
m4_data_dir <- "/Users/briday/Desktop/study_stats/redistribute_facility_deaths/outputs/tables"

m4_summary <- read.csv(file.path(m4_data_dir, "coverage_office_distance_summary_western_cape_2026-06-09.csv"))

# Extract key metrics from the summary CSV (metric/value format)
m4_metrics <- setNames(m4_summary$value, m4_summary$metric)

m4_pct_5km_road <- m4_metrics["pct_population_within_5km_road"]
m4_pct_10km_road <- m4_metrics["pct_population_within_10km_road"]
m4_pct_20km_road <- m4_metrics["pct_population_within_20km_road"]

m4_pct_10min <- m4_metrics["pct_population_within_10min_drive"]
m4_pct_30min <- m4_metrics["pct_population_within_30min_drive"]
m4_pct_60min <- m4_metrics["pct_population_within_60min_drive"]

m4_median_road <- m4_metrics["median_km_road_popweighted"]
m4_total_pop <- m4_metrics["total_population"]
m4_n_offices <- m4_metrics["n_offices"]
```

## Method 1: District-Level Office-to-Office Distance {#method1}

::: callout-note
## Approach

Half the median nearest-neighbour distance between DHA offices in a district is used as a single distance proxy for the district's whole population.
:::

::: callout-important
## Key finding

**`r m1_national$pct_5km`%** of the population lives within 5km, **`r m1_national$pct_10km`%** within 10km, **`r m1_national$pct_20km`%** within 20km (death registration, DHA offices only).
:::

::: callout-warning
## Limitations

Ignores within-district variation and actual population distribution; conservative bias in rural districts with sparse but well-placed offices.
:::

## Method 2: District-Level Modelled Population Grid {#method2}

::: callout-note
## Approach

A 1km grid is laid over each district; population is redistributed onto grid points using an **inverse power decay model** ($\alpha = 1.5$) that clusters population near offices, and distance to the nearest office is measured per grid point.
:::

::: callout-important
## Key finding

**`r m2_pct_5km`%** within 5km, **`r m2_pct_10km`%** within 10km, **`r m2_pct_20km`%** within 20km (death registration, DHA offices only).
:::

::: callout-warning
## Limitations

Population is modelled, not observed; the decay parameter is not empirically calibrated; distances are still straight-line.
:::

## Method 3: Ward-Level Centroid Distance {#method3}

::: callout-note
## Approach

Actual ward-level population (Stats SA, 4,468 wards) is joined to ward polygons (MDB 2020); straight-line distance is measured from each **ward centroid** to the nearest registration node. Run twice: once against DHA offices only (death registration) and once adding hospital DHA satellite desks (birth registration).
:::

::: callout-important
## Key finding

Death registration: **`r m3_pct_5km`%** within 5km, **`r m3_pct_10km`%** within 10km, **`r m3_pct_20km`%** within 20km. Birth registration (with hospital satellites): **`r birth_pct_5km`%**, **`r birth_pct_10km`%**, **`r birth_pct_20km`%** respectively.
:::

::: callout-warning
## Limitations

Ward centroids are a proxy for where people live — large rural wards may have centroids far from population clusters; `r missing_pop` wards are missing population data; distances remain straight-line, not road-network.
:::

## Method 4: WorldPop Grid + OSM Road Network (Western Cape) {#method4}

::: callout-note
## Approach

100m-resolution **observed** (satellite-derived) population from WorldPop is routed to the nearest DHA office over the actual **OpenStreetMap road network**, giving road-distance and driving-time coverage rather than straight-line distance. Implemented for the Western Cape only, in the companion [redistribute_facility_deaths](https://github.com/bridaybrummer/redistribute_facility_deaths) pipeline.
:::

::: callout-important
## Key finding

Road distance: **`r m4_pct_5km_road`%** within 5km, **`r m4_pct_10km_road`%** within 10km, **`r m4_pct_20km_road`%** within 20km. Driving time: **`r m4_pct_30min`%** within 30 minutes, **`r m4_pct_60min`%** within 60 minutes (death registration, DHA offices only, Western Cape).
:::

::: callout-warning
## Limitations

Western Cape only — not yet run nationally; assumes free-flow road speeds; excludes residential/unclassified streets; no birth-registration (hospital satellite) analysis has been run against this method.
:::

------------------------------------------------------------------------

# Comparison of Methods {#comparison}

## Qualitative Comparison

```{r}
#| label: comparison-table-full

comparison_df_v2 <- tibble::tribble(
  ~Feature, ~`Method 1`, ~`Method 2`, ~`Method 3`, ~`Method 4`,
  "Geographic unit", "District", "1km grid within district", "Ward (~4,400 units)", "100m population cell",
  "Population data", "District totals", "District totals (modelled to grid)", "Ward-level counts (observed)", "WorldPop satellite grid (observed)",
  "Distance measure", "Half-median inter-office distance", "Grid point to nearest office", "Ward centroid to nearest office", "Road-network distance & driving time",
  "Birth vs death split", "No", "No", "Yes (hospital satellites added)", "No",
  "Geographic coverage", "National", "National", "National", "Western Cape only",
  "Key weakness", "Single estimate per district", "Modelled population may not reflect reality", "Centroid may not represent large rural wards", "Not yet run nationally"
)

flextable(comparison_df_v2) |>
  bold(part = "header") |>
  bg(j = "Method 3", bg = "#E8F5E9") |>
  bg(j = "Method 4", bg = "#FFF3E0") |>
  theme_zebra() |>
  autofit() |>
  set_caption("Qualitative comparison of the four DHA access methodologies")
```

## Completeness Proportions Across Methods

The table below is the core comparison: estimated population **completeness** (% within reach of a registration node) for each method, split by registration type where the analysis was run.

```{r}
#| label: completeness-comparison-table

completeness_df <- tibble::tribble(
  ~Method, ~Scope, ~Registration, ~`% < 5km`, ~`% < 10km`, ~`% < 20km`,
  "Method 1", "National", "Death (DHA only)", paste0(m1_national$pct_5km, "%"), paste0(m1_national$pct_10km, "%"), paste0(m1_national$pct_20km, "%"),
  "Method 1", "National", "Birth", "N/A", "N/A", "N/A",
  "Method 2", "National", "Death (DHA only)", paste0(m2_pct_5km, "%"), paste0(m2_pct_10km, "%"), paste0(m2_pct_20km, "%"),
  "Method 2", "National", "Birth", "N/A", "N/A", "N/A",
  "Method 3", "National", "Death (DHA only)", paste0(m3_pct_5km, "%"), paste0(m3_pct_10km, "%"), paste0(m3_pct_20km, "%"),
  "Method 3", "National", "Birth (DHA + hospital)", paste0(birth_pct_5km, "%"), paste0(birth_pct_10km, "%"), paste0(birth_pct_20km, "%"),
  "Method 4", "Western Cape", "Death, road distance (DHA only)", paste0(m4_pct_5km_road, "%"), paste0(m4_pct_10km_road, "%"), paste0(m4_pct_20km_road, "%"),
  "Method 4", "Western Cape", "Birth", "N/A", "N/A", "N/A"
)

flextable(completeness_df) |>
  bold(part = "header") |>
  bg(i = ~ Registration %in% "Birth (DHA + hospital)", bg = "#E8F5E9") |>
  merge_v(j = c("Method", "Scope")) |>
  theme_zebra() |>
  autofit() |>
  set_caption("Population completeness (% within distance threshold) across all four methods, split by birth vs death registration where available")
```

::: callout-note
Birth registration is only quantified separately for Method 3, because hospital-based DHA satellite desks (valid for births, not deaths) have so far only been linked into the ward-centroid pipeline. Extending the birth/death split to Methods 1, 2 and 4 is future work.
:::

------------------------------------------------------------------------

# Chosen Method: Method 3 — Ward-Level Centroid Distance {#chosen-method}

## Why this method

Method 3 is used for reporting because it is the only approach that is simultaneously **national in scope**, uses **actual observed population** (not modelled), and has been extended to **both birth and death registration** by including hospital-based DHA satellite desks. Method 4 is more physically realistic (real road distances) but is currently limited to the Western Cape; Method 3 remains the best available national estimate until Method 4 is extended to all provinces.

## Results: Death Registration (DHA Offices Only)

```{r}
#| label: chosen-death-summary
m3_dist_summary
```

```{r}
#| label: chosen-death-table
ward_access_ft
```

```{r}
#| label: chosen-death-map
#| fig-width: 10
#| fig-height: 8
ward_distance_map
```

```{r}
#| label: chosen-death-curve
#| fig-width: 10
#| fig-height: 6
coverage_curve_plot
```

## Results: Birth vs Death Registration

Adding hospital-based DHA satellite desks (valid for **birth** registration only) increases coverage relative to death registration, which is restricted to standalone DHA offices.

```{r}
#| label: chosen-birth-death-table
if (exists("birth_vs_death_ft") && !is.null(birth_vs_death_ft)) {
  birth_vs_death_ft
} else {
  cat("Birth registration comparison not available — run DHA_access/link_DHA_to_MFL.r first.")
}
```

```{r}
#| label: chosen-birth-death-map
#| fig-width: 10
#| fig-height: 8
if (exists("birth_distance_map") && !is.null(birth_distance_map)) {
  birth_distance_map
}
```

```{r}
#| label: chosen-birth-death-curve
#| fig-width: 10
#| fig-height: 6
if (exists("birth_vs_death_curve_plot") && !is.null(birth_vs_death_curve_plot)) {
  birth_vs_death_curve_plot
}
```

::: callout-important
## Key Finding

Death registration (DHA offices only): **`r m3_pct_5km`%** of the population is within 5km, **`r m3_pct_10km`%** within 10km, **`r m3_pct_20km`%** within 20km. Birth registration (DHA + hospital satellites): **`r birth_pct_5km`%**, **`r birth_pct_10km`%**, **`r birth_pct_20km`%** respectively — a gain of **`r round(birth_pct_5km - m3_pct_5km, 1)`, `r round(birth_pct_10km - m3_pct_10km, 1)` and `r round(birth_pct_20km - m3_pct_20km, 1)` percentage points** at the 5/10/20km thresholds.
:::

::: callout-warning
## Limitations

-   Ward **centroids** are a proxy for where people live — large rural wards may have centroids far from actual population clusters
-   `r missing_pop` wards are missing population data
-   Distances are **straight-line**, not road-network or travel-time (Method 4 shows this matters — road distance coverage is consistently lower than straight-line at the same threshold)
-   Does not account for office/hospital capacity, opening hours or queueing
-   Hospital DHA satellite list is a curated snapshot and may not reflect the current set of facilities offering birth registration
:::

# Next Steps

1.  **Extend Method 4 (road-network routing) nationally** and re-run the birth-vs-death split against it
2.  **Add hospital DHA satellites to Methods 1, 2 and 4** so the birth/death completeness table has no "N/A" cells
3.  **Obtain enumeration area data** from Stats SA to validate ward-centroid and modelled-grid population placement
