---
title: "MRC VR National Overview"
description: "National observed vs expected deaths analysis with provincial breakdowns"
date: "2026-01-20"
categories: [excess-mortality, national, provincial, modelling]
execute:
  echo: false
  warning: false
  message: false
  cache: false
page-navigation: true
toc: true
---

```{r}
#| label: setup
#| include: false

# library(NMCleaner)
library(arrow)
library(knitr)
pacman::p_load(
    ggplot2, plotly, gtsummary, tidyverse, magrittr, conflicted,
    htmlwidgets, knitr, webshot, rmapshaper, ggpubr, scales, here, tibble, glue,
    data.table
)

knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE, cache.compress = TRUE)
# If you rely on these, load once here (keeps later chunks lean)
conflicted::conflicts_prefer(
    dplyr::filter,
    dplyr::select,
    flextable::compose
)

source(here::here("dashboard_functions.r"))
# source("dashboard_functions.R")
# Data load goes here (kept commented if you load upstream)
# dt_analysis <- read_feather("../LGH_AnalysisData_2022.feather") %>% as.data.table()

# Expect icd_lookup and dt_analysis to exist; fail friendly:
stopifnot(exists("icd_lookup"), exists("dt_analysis"))
```

```{css}
/* Gentle spacing for headings inside tabsets */
.panel-tabset h3, .panel-tabset h4 { margin-top: .6rem; }
/* Make the pill tabs a touch bolder and tighter */
.nav-pills > li > a, .nav-pills .nav-link { font-weight: 600; }
/* Compact the inner tab-strip a bit */
.panel-tabset > .nav-tabs { margin-bottom: .5rem; }
/* Slightly narrower body for readability on wide screens */
.quarto-dashboard .content { max-width: 1600px; margin-left: auto; margin-right: auto; }
/* Tables: keep them from stretching edge-to-edge */
.table, .gt_table, .gts_table { margin: .25rem auto .75rem auto; }
/* Small card-ish feel for content blocks */
.section > :where(h2,h3,h4) + * { background: rgba(0,0,0,.015); padding: .35rem .5rem; border-radius: .4rem; }
```


```{r}
load("../figures/model_comparison_prov_nat_plots.rda")
#load("figures/model_comparison_prov_nat_plots.rda")
```

# National Picture
Use the nav bar at the top of the page to navigate to cause-specific tabs below.

```{r}
#| label: fig-national_model_plot_SA
#| fig-cap:  National observed vs expected deaths, 2015-2022 (TM's National Negative Binomial Model converted from stats to R code)
#| fig-width: 12
#| fig-height: 8
#| fig-align: "center"
#| out-width: "75%"
#| fig-dpi: 200

national_model_plot_SA

```

```{r}
#| label: fig-national_model_icd_O_vs_E_plot_SA
#| fig-cap:  National observed vs expected deaths difference 2019-2022 by ICD code
#| fig-width: 12
#| fig-height: 8
#| fig-align: "center"
#| out-width: "75%"
#| fig-dpi: 200

national_model_icd_O_vs_E_plot_SA

```

```{r}
#| label: fig-national_model_icd_difference_plot_SA
#| fig-cap:  National observed vs expected deaths difference 2019-2022 by ICD code
#| fig-width: 12
#| fig-height: 8
#| fig-align: "center"
#| out-width: "75%"
#| fig-dpi: 200

national_model_icd_difference_plot_SA
```



```{r}
#| label: fig-national_model_cause_heatmap_SA
#| fig-cap:  National observed vs expected deaths difference heatmap 2019-2022 by ICD code and year
#| fig-width: 12
#| fig-height: 8
#| fig-align: "center"
#| out-width: "75%"
#| fig-dpi: 200

national_model_cause_heatmap_SA

```



# Modelling provincial level 

TH's NB model with interaction terms takes a relatively longer time and similar results can be achieved with no interaction terms but adding provinicial terms.  The models appear to do well at the national level compared to Toms model.  I did not find this provincial level model to perform any better when adding interaction terms, population offsets, or fourier terms. 

```{r}
#| label: fig-national_vs_provincial_model_diff_plot_SA
#| fig-cap:  Difference between provincial simple model and national model
#| fig-width: 12
#| fig-height: 8
#| fig-align: "center"
#| out-width: "75%"
#| fig-dpi: 200
national_vs_provincial_model_diff_plot_SA

```

```{r}
#| label: fig-provincial_model_prov_O_vs_E_plot_SA
#| fig-cap:  Provincial observed vs expected deaths 2015-2022 (Provincial Negative Binomial Model without interaction terms)
#| fig-width: 12
#| fig-height: 8
#| fig-align: "center"
#| out-width: "75%"
#| fig-dpi: 200

provincial_model_prov_O_vs_E_plot_SA

```
The provincial model doesn't appear to do well with ICD groupings at the provincial level; this can be viewed when inspecting the provincial model ICD observed vs expected plots below in the tab.

[Go to Cause codes](#cause-codes){.btn .btn-primary}

# Cause codes {.tabset .tabset-pills}
```{r}
#| results: asis

# --- helpers ---
get_desc <- function(code) {
    d <- icd_lookup$description[match(code, icd_lookup$LGH_Cause)]
    ifelse(is.na(d) || identical(d, ""), "Missing / Unknown", d)
}

# Pull codes as a clean vector and drop obvious empties once
icd_codes <- unique(icd_lookup$LGH_Cause)
icd_codes <- icd_codes[!is.na(icd_codes) & icd_codes != ""]

# --- open the outer containers ONCE ---
cat("::: {.nav-pills}\n")
cat("::: {.panel-tabset}\n")

# provice explanation of navigation
cat("Use the tabs to navigate between cause-specific analyses. Each cause has sub-tabs for Summary, By Age Group, By Province, and Multiple Cause Codes which may be at the bottom of the page if your screen is small.\n\n")

for (icd_code in icd_codes) {
    desc <- get_desc(icd_code)
    code_label <- ifelse(is.na(icd_code) || identical(icd_code, ""), "NA", icd_code)
    tabs <- make_sequence_tables(dt, icd_code = icd_code)


    title <- glue("{code_label} — {desc}")

    # H2 tab for this ICD code
    cat(glue("## {title}\n\n"))

    # inner tabset for Summary / Observed vs Expected / Multiple cause codes
    cat("::: {.panel-tabset}\n")

    # --- Summary ---
    cat("### Summary Table\n\n")
    # Expect your function returns a flextable; convert for Rmd
    cat(flextable_to_rmd(tabulate_excess_factor(dt_analysis, icd_codes = icd_code)))

    cat("\n\n")
    # By Age group
    cat("### By Age Group\n\n")

    cat("::: {.columns}\n")

    # Actual

    cat("::: {.column width=\"100%\"}\n")
    cat("#### Actual\n\n")
    print(
        icd_code_by_agegroup(
            dt_analysis,
            icd_code = icd_code,
            actual_vs_difference = "actual"
        )
    )
    cat("\n:::\n")

    # Difference

    cat("::: {.column width=\"100%\"}\n")
    cat("#### Difference\n\n")
    print(
        icd_code_by_agegroup(
            dt_analysis,
            icd_code = icd_code,
            actual_vs_difference = "difference"
        )
    )
    cat("\n:::\n")

    cat(":::\n\n") # close columns

    # BY province
    cat("### By Province\n\n")
    cat("::: {.columns}\n")
    cat("::: {.column width=\"100%\"}\n")
    cat("#### Actual\n\n")
    print(
        icd_code_by_province(
            dt_collapse_prov,
            icd_code = icd_code,
            actual_vs_difference = "actual"
        )
    )
    cat("\n:::\n")
    # difference
    cat("::: {.column width=\"100%\"}\n")
    cat("#### Difference\n\n")
    print(
        icd_code_by_province(
            dt_collapse_prov,
            icd_code = icd_code,
            actual_vs_difference = "difference"
        )
    )
    cat("\n:::\n")

    cat(":::\n\n") # close columns

    # --- Multiple cause codes ---

    cat("### Multiple cause codes\n\n")
    cat("::: {.panel-tabset}\n")

    cat("\n\n")
    cat("#### Cause A Singlets  \n\n") # replace with your table output when ready
    cat(flextable_to_rmd(tabs$ft1))

    cat("\n\n")
    cat("#### Cause A \u2192 B doubles  \n\n") # replace with your table output when ready
    cat(flextable_to_rmd(tabs$ft2_global))

    cat("\n\n")
    cat("#### Cause A \u2192 B (B within As)  \n\n") # replace with your table output when ready
    cat(flextable_to_rmd(tabs$ft2_withinA))

    cat("\n\n")
    cat("#### Cause A \u2192 B \u2192 C triplets \n\n") # replace with your table output when ready
    cat(flextable_to_rmd(tabs$ft3_global))

    cat("\n\n")
    cat("#### Cause A \u2192 B \u2192 C \u2192 quadruplets \n\n") # replace with your table output when ready
    cat(flextable_to_rmd(tabs$ft4_global))

    cat("\n\n")
    cat(":::\n\n") # close Multiple cause codes .panel-tabset




    cat(":::\n\n") # close inner .panel-tabset
}

# --- close the outer containers ---
cat(":::\n") # close outer .panel-tabset
cat(":::\n") # close .nav-pills
```
