---
title: "ICD-10 ↔ NMC / IDSR Condition Explorer"
description: "Searchable reference linking Notifiable Medical Conditions (NMC) to their verified ICD-10 / ICD-11 codes, with coding pitfalls and MACOD data-quality notes"
date: "2026-07-23"
categories: [NMC, IDSR, ICD-10, ICD-11, reference, data-quality]
execute:
  echo: false
  warning: false
  message: false
  cache: false
toc: true
resources:
  - "../../outputs/nmc_icd_full_reference.csv"
---

```{r}
#| label: setup
#| include: false

suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(DT)
  library(here)
})

knitr::opts_knit$set(root.dir = here::here())

ref <- read_csv(here("outputs/nmc_icd_full_reference.csv"), show_col_types = FALSE)
```

## What this is

A single reference table answering, for **each Notifiable Medical Condition (NMC)** — South Africa's IDSR-aligned notification list — the question: *"if this condition were the underlying cause of death, what ICD-10 and ICD-11 code should be assigned?"* Alongside the verified **UCoD (underlying cause of death) ICD-10 and ICD-11 codes**, the table lists known **coding pitfalls** and how the condition tends to actually appear (or not) in **MACOD** vital-registration data.

Use the search box or per-column filters below to find a condition, category, or ICD code. Columns wrap long text — click a cell to expand, or use the **CSV / Excel** buttons to export the full table.

::: callout-note
This table is generated from the SA NMC ICD Reference (v3) and rebuilt by `nmc_icd10_verifier/build_verified_nmc_table.R`. It is also used as the source map for the [NMC & MACOD Comparison](../nmc_macod_comparison/index.qmd) project, where the counts and MACOD sensitivity analysis derived from this map are presented.
:::

## Definitions & cause-of-death coding

Medical certification of death (the WHO/MCCD form) records causes in a structured sequence, and only one cause per death is coded as the statistic used in mortality reporting.

| Term | Definition |
|---|---|
| **Underlying cause of death (UCoD)** | "The disease or injury which initiated the train of morbid events leading directly to death." This is the single cause selected for national/international mortality statistics (WHO ICD-10/ICD-11 rules), and is the code shown in the **UCoD ICD-10 / UCoD ICD-11** columns below. |
| **Immediate cause of death** | The final disease or condition directly causing death (death certificate **Part I(a)**) — e.g. "septic shock" — which is usually a *consequence* of the underlying cause, not the underlying cause itself. |
| **Antecedent / intervening causes** | Conditions reported in Part I(b)/(c) that link the immediate cause back to the underlying cause (the causal chain). |
| **Multiple cause of death (MCOD) coding** | All conditions listed on the certificate (Part I chain **and** Part II contributory conditions) are coded, not just the UCoD. MCOD analysis can reveal a notifiable condition even when it was not selected as the UCoD (e.g. HIV listed as a contributory cause under a UCoD of TB). |
| **Part II — contributory conditions** | Conditions that contributed to death but were not part of the direct causal chain to the immediate cause (e.g. diabetes, hypertension). These are coded but never selected as UCoD. |

**Why this matters for NMC ↔ MACOD linkage:** the NMC/MACOD comparison and sensitivity tables in this site match on the **UCoD only** — the single cause South Africa's vital statistics report. A notifiable condition that is frequently recorded as an immediate or contributory cause (rather than the UCoD) will therefore appear **under-captured** in MACOD, even if it was mentioned on the death certificate. This is a key reason capture ratios in the [NMC & MACOD Comparison](../nmc_macod_comparison/index.qmd) sensitivity analysis fall below 100% for some conditions — the coding pitfall columns below flag where this is a known issue.

For each row: **UCoD ICD-10** / **UCoD ICD-11** are the codes to assign *when this NMC condition is the underlying cause of death*.

```{r}
#| label: tbl-icd-nmc-explorer
#| tbl-cap: "Underlying cause of death (UCoD) ICD-10 / ICD-11 codes for all NMC / IDSR conditions"

ref_display <- ref |>
  arrange(nmc_category, nmc_condition) |>
  rename(
    `Category`                       = nmc_category,
    `NMC Condition`                  = nmc_condition,
    `UCoD ICD-10`                    = icd10_code,
    `UCoD ICD-11`                    = icd11_code,
    `Pathogen / specificity notes`   = pathogen_notes,
    `Surveillance artefacts & coding pitfalls` = surveillance_artefacts,
    `SA MACOD data quality note`     = macod_note,
    `Reporting timeframe`            = reporting_timeframe
  )

datatable(
  ref_display,
  filter     = "top",
  extensions = "Buttons",
  rownames   = FALSE,
  options    = list(
    dom        = "Bfrtip",
    buttons    = c("csv", "excel"),
    pageLength = 15,
    scrollX    = TRUE,
    autoWidth  = FALSE,
    columnDefs = list(
      list(width = "80px",  targets = 0),  # Category
      list(width = "220px", targets = 1),  # NMC Condition
      list(width = "80px",  targets = 2),  # ICD-10
      list(width = "80px",  targets = 3),  # ICD-11
      list(width = "320px", targets = 4),  # Pathogen notes
      list(width = "320px", targets = 5),  # Surveillance artefacts
      list(width = "320px", targets = 6),  # MACOD note
      list(width = "120px", targets = 7)   # Reporting timeframe
    )
  )
)
```

## Download the raw map

The underlying two-column verified map (NMC condition → primary UCoD ICD-10 code) used for joining to MACOD/VR data is available here:

- [`nmc_icd_full_reference.csv`](../../outputs/nmc_icd_full_reference.csv) — full reference with categories, UCoD ICD-10/ICD-11 codes, and coding notes
- Use the **CSV** / **Excel** export buttons above to download a filtered subset of the table

## Coverage note

```{r}
#| label: coverage-summary

n_conditions <- nrow(ref)
n_cat <- ref |> distinct(nmc_category) |> nrow()
```

This reference covers **`r n_conditions` NMC/IDSR conditions** across **`r n_cat` reporting categories**. One NMC condition — *Waterborne illness outbreak – undefined* — has no assigned UCoD ICD-10 code in the source map and is therefore excluded from ICD-based matching (e.g. in the MACOD comparison).
