---
title: "ICD-10 ↔ NMC / IDSR Condition Explorer"
description: "Searchable reference linking Notifiable Medical Conditions (NMC) to their verified ICD-10 / ICD-11 codes, with coding pitfalls and MACOD data-quality notes"
date: "2026-07-23"
categories: [NMC, IDSR, ICD-10, ICD-11, reference, data-quality]
execute:
  echo: false
  warning: false
  message: false
  cache: false
toc: true
resources:
  - "../../outputs/nmc_icd_full_reference.csv"
---

```{r}
#| label: setup
#| include: false

suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(DT)
  library(here)
})

knitr::opts_knit$set(root.dir = here::here())

ref <- read_csv(here("outputs/nmc_icd_full_reference.csv"), show_col_types = FALSE)
```

## What this is

A single reference table answering, for **each Notifiable Medical Condition (NMC)** — South Africa's IDSR-aligned notification list — the question: *"if this condition were the underlying cause of death, what ICD-10 and ICD-11 code should be assigned?"* Alongside the verified **UCoD (underlying cause of death) ICD-10 and ICD-11 codes**, the table lists known **coding pitfalls** and how the condition tends to actually appear (or not) in **MACOD** vital-registration data.

Use the search box or per-column filters below to find a condition, category, or ICD code. Columns wrap long text — click a cell to expand, or use the **CSV / Excel** buttons to export the full table.

::: callout-note
This table is generated from the SA NMC ICD Reference (v3) and rebuilt by `nmc_icd10_verifier/build_verified_nmc_table.R`. It is also used as the source map for the [NMC & MACOD Comparison](../nmc_macod_comparison/index.qmd) project, where the counts and MACOD sensitivity analysis derived from this map are presented.
:::

## Definitions & cause-of-death coding

Medical certification of death (the WHO/MCCD form) records causes in a structured sequence, and only one cause per death is coded as the statistic used in mortality reporting.

| Term | Definition |
|---|---|
| **Underlying cause of death (UCoD)** | "The disease or injury which initiated the train of morbid events leading directly to death." This is the single cause selected for national/international mortality statistics (WHO ICD-10/ICD-11 rules), and is the code shown in the **UCoD ICD-10 / UCoD ICD-11** columns below. |
| **Immediate cause of death** | The final disease or condition directly causing death (death certificate **Part I(a)**) — e.g. "septic shock" — which is usually a *consequence* of the underlying cause, not the underlying cause itself. |
| **Antecedent / intervening causes** | Conditions reported in Part I(b)/(c) that link the immediate cause back to the underlying cause (the causal chain). |
| **Multiple cause of death (MCOD) coding** | All conditions listed on the certificate (Part I chain **and** Part II contributory conditions) are coded, not just the UCoD. MCOD analysis can reveal a notifiable condition even when it was not selected as the UCoD (e.g. HIV listed as a contributory cause under a UCoD of TB). |
| **Part II — contributory conditions** | Conditions that contributed to death but were not part of the direct causal chain to the immediate cause (e.g. diabetes, hypertension). These are coded but never selected as UCoD. |

**Why this matters for NMC ↔ MACOD linkage:** the NMC/MACOD comparison and sensitivity tables in this site match on the **UCoD only** — the single cause South Africa's vital statistics report. A notifiable condition that is frequently recorded as an immediate or contributory cause (rather than the UCoD) will therefore appear **under-captured** in MACOD, even if it was mentioned on the death certificate. This is a key reason capture ratios in the [NMC & MACOD Comparison](../nmc_macod_comparison/index.qmd) sensitivity analysis fall below 100% for some conditions — the coding pitfall columns below flag where this is a known issue.

For each row: **UCoD ICD-10** / **UCoD ICD-11** are the codes to assign *when this NMC condition is the underlying cause of death*. The **Likely immediate-cause / miscode ICD-10s** column lists specific proxy codes that tend to appear in MACOD instead — i.e. what the death is actually coded as when the notifiable condition is not correctly identified/documented as the UCoD (e.g. Measles deaths often carry an immediate cause of `J18.9` "pneumonia, unspecified" rather than the measles-specific code).

```{r}
#| label: tbl-icd-nmc-explorer
#| tbl-cap: "Underlying cause of death (UCoD) ICD-10 / ICD-11 codes for all NMC / IDSR conditions"

# ── Curated lookup: likely immediate-cause / miscode ICD-10 codes ───────────
# For each NMC condition, the specific proxy/terminal-event ICD-10 codes that
# tend to appear in MACOD in place of the correct UCoD, when the notifiable
# condition itself is not documented on the death certificate (e.g. Measles
# deaths often carry an immediate cause of J18.9 "pneumonia, unspecified"
# rather than B05.2 "measles with pneumonia"). Curated from the surveillance
# artefact / MACOD notes above.
miscode_lookup <- tibble::tribble(
  ~nmc_condition, ~likely_miscodes,
  "Acute flaccid paralysis", "G83.9 retained even when polio (A80) confirmed",
  "Acute rheumatic fever", "I05\u2013I09 (chronic valvular/RHD); BA80 (ICD-11 terminal RHD)",
  "Anthrax", "A41.9 (sepsis, unspecified)",
  "Botulism", "J96 (respiratory failure); G83.9 (AFP); P28.5 (newborn resp. failure)",
  "Cholera", "A09 (gastroenteritis NOS); E86 (dehydration)",
  "Diphtheria", "I51.4 (myocarditis, unspecified)",
  "Enteric fever (typhoid / paratyphoid)", "R50.9 (fever NOS); A09 (gastroenteritis NOS); K63.1 (intestinal perforation as UCoD)",
  "Food-borne disease outbreak", "A09; A05.9 (no outbreak linkage flag)",
  "Haemolytic uraemic syndrome (HUS)", "D59.3 used as UCoD instead of A04.3 (STEC) \u2014 sequencing reversed",
  "Listeriosis", "P36.9 (neonatal sepsis, unspecified)",
  "Malaria", "R50.9 (fever NOS); G93.6 (cerebral oedema)",
  "Measles", "J18.9 (pneumonia, unspecified); G04.9 (encephalitis, unspecified)",
  "Meningococcal disease", "A41.9 (sepsis NOS); G03.9 (meningitis NOS); D65 (DIC, terminal event)",
  "Pertussis", "J18.9; J21.9; J96.0 (respiratory failure, infants)",
  "Plague", "A41.9 (sepsis NOS); J15.9 (pneumonia NOS, pneumonic form)",
  "Poliomyelitis", "J96 (respiratory failure)",
  "Rabies (human)", "G04.9 (encephalitis NOS); A86 (viral encephalitis NOS); G61.0 (Guillain-Barr\u00e9, dumb rabies mistaken)",
  "Respiratory disease \u2014 novel pathogen", "J18.9 (pneumonia NOS); B34.2 (coronavirus NOS); R99 (ill-defined)",
  "Rift Valley fever (human)", "A99 (haemorrhagic fever NOS); A86 (viral encephalitis NOS)",
  "Smallpox", "Miscoded as B04 (mpox) or B01 (varicella)",
  "Viral haemorrhagic fever diseases", "A41.9 (sepsis NOS); D65 (DIC); R68.89 (multi-organ failure)",
  "Yellow fever", "K72.0 (liver failure); A99 (haemorrhagic fever NOS)",
  "Agricultural/stock remedy poisoning", "J96 (respiratory failure); I46.9 (cardiac arrest)",
  "Bilharzia (schistosomiasis)", "K74.6 (cirrhosis); K76.6 (portal hypertension); K92.2 (UGI bleed)",
  "Brucellosis", "R50.9 (fever NOS); A41.9 (sepsis NOS); I38 (endocarditis)",
  "Congenital rubella syndrome", "Q20\u2013Q24 (structural cardiac defects); Q12 (ophthalmic defects)",
  "Congenital syphilis", "P36.9 (neonatal sepsis NOS); P07 (preterm)",
  "Coronavirus disease (COVID-19)", "J18.9 (pneumonia NOS); B34.2 (coronavirus NOS); R99 (ill-defined)",
  "Haemophilus influenzae type B", "G03.9 (meningitis NOS); J18.9 (pneumonia NOS); A41.9 (sepsis NOS)",
  "Hepatitis A", "K72.0 (liver failure); K72.9 (hepatic encephalopathy)",
  "Hepatitis B", "K74.6 (cirrhosis); K72.1 (chronic liver failure); C22.0 (HCC) without B18.1 linked",
  "Hepatitis C", "K74.6 (cirrhosis); C22.0 (HCC) without B18.2 linked",
  "Hepatitis E", "O26.6 (liver disorder in pregnancy); O95 (maternal death NOS)",
  "Lead poisoning", "N18.5 (ESRD); G93.4 (encephalopathy NOS); D64.9 (anaemia)",
  "Legionellosis", "J18.9; J15.9 (pneumonia, unspecified)",
  "Leprosy", "N18.5 (ESRD); A41.9 (sepsis NOS); terminal complication codes",
  "Maternal death", "A15/B20 (TB/HIV coded outside O-chapter when indirect cause); conversely O95 over-coded for non-obstetric deaths",
  "Mercury poisoning", "N18.9 (CKD NOS); G62.2 (toxic polyneuropathy); R99 (ill-defined)",
  "Soil-transmitted helminths", "D50 (anaemia); E44 (malnutrition); K56.6 (bowel obstruction)",
  "Tetanus", "J96 (respiratory failure); J69.0 (aspiration pneumonia); P36.9 (neonatal sepsis NOS)",
  "Tuberculosis: MDR-TB", "Identical to DS-TB (A15.0) \u2014 resistance invisible without NHLS linkage",
  "Tuberculosis: XDR-TB", "Identical to DS-TB (A15.0) \u2014 resistance invisible without NHLS linkage",
  "Tuberculosis: extra-pulmonary", "I31 (pericardial disease NOS); G03.9 (meningitis NOS); A41.9/J80 (ARDS, miliary TB)",
  "Tuberculosis: pulmonary", "B20/B22.1 (HIV disease) coded instead of A15.0 in TB/HIV coinfection",
  "Endemic arboviral diseases (West Nile/Sindbis/Chikungunya)", "G04.9 (encephalitis NOS); A86 (viral encephalitis NOS); I51.4 (myocarditis NOS)",
  "Gonorrhoea (ceftriaxone-resistant N.gonorrhoeae)", "A41.89 (other sepsis NOS, no AMR flag)",
  "Non-endemic arboviral diseases (Dengue/other imported)", "D69.6 (thrombocytopaenia); K72.0 (liver failure); A99 (haemorrhagic fever NOS)",
  "Non-typhoidal Salmonellosis (NTS)", "B20/B22 (HIV disease); A41.89 (sepsis NOS)",
  "Rubella", "A86 (viral encephalitis NOS); CRS (P35.0) more common entry point",
  "Shiga toxin-producing E.coli (STEC)", "D59.3 (HUS); N17.9 (AKI) used as UCoD instead of A04.3",
  "Shigellosis", "A09 (gastroenteritis NOS); E86 (dehydration); K52.9 (haemorrhagic colitis)",
  "HCAI/MDRO \u2014 CPE (Carbapenemase-producing Enterobacteriaceae)", "A41.5 (Gram-negative sepsis NOS, WHO ICD-10 collapses to this)",
  "HCAI/MDRO \u2014 Clostridioides difficile (CDI)", "K52.1 (pseudomembranous colitis NOS); K57.2 (toxic megacolon); sepsis codes",
  "HCAI/MDRO \u2014 Colistin-resistant Acinetobacter baumannii", "A41.5 (Gram-negative sepsis NOS); J15.9 (bacterial pneumonia NOS)",
  "HCAI/MDRO \u2014 Colistin-resistant Pseudomonas aeruginosa", "A41.5 (Gram-negative sepsis NOS, no organism ID)",
  "HCAI/MDRO \u2014 S.aureus (MRSA/hGISA/GISA)", "A41.0 (S.aureus sepsis) \u2014 MRSA indistinguishable from MSSA"
)

ref <- ref |>
  left_join(miscode_lookup, by = "nmc_condition")

ref_display <- ref |>
  arrange(nmc_category, nmc_condition) |>
  mutate(likely_miscodes = tidyr::replace_na(likely_miscodes, "Not documented")) |>
  rename(
    `Category`                       = nmc_category,
    `NMC Condition`                  = nmc_condition,
    `UCoD ICD-10`                    = icd10_code,
    `UCoD ICD-11`                    = icd11_code,
    `Pathogen / specificity notes`   = pathogen_notes,
    `Likely immediate-cause / miscode ICD-10s` = likely_miscodes,
    `Surveillance artefacts & coding pitfalls` = surveillance_artefacts,
    `SA MACOD data quality note`     = macod_note,
    `Reporting timeframe`            = reporting_timeframe
  )

datatable(
  ref_display,
  filter     = "top",
  extensions = "Buttons",
  rownames   = FALSE,
  options    = list(
    dom        = "Bfrtip",
    buttons    = c("csv", "excel"),
    pageLength = 15,
    scrollX    = TRUE,
    autoWidth  = FALSE,
    columnDefs = list(
      list(width = "80px",  targets = 0),  # Category
      list(width = "220px", targets = 1),  # NMC Condition
      list(width = "80px",  targets = 2),  # ICD-10
      list(width = "80px",  targets = 3),  # ICD-11
      list(width = "260px", targets = 4),  # Pathogen notes
      list(width = "260px", targets = 5),  # Likely miscodes
      list(width = "320px", targets = 6),  # Surveillance artefacts
      list(width = "320px", targets = 7),  # MACOD note
      list(width = "120px", targets = 8)   # Reporting timeframe
    )
  )
)
```

## Download the raw map

The underlying two-column verified map (NMC condition → primary UCoD ICD-10 code) used for joining to MACOD/VR data is available here:

- [`nmc_icd_full_reference.csv`](../../outputs/nmc_icd_full_reference.csv) — full reference with categories, UCoD ICD-10/ICD-11 codes, and coding notes
- Use the **CSV** / **Excel** export buttons above to download a filtered subset of the table

## Coverage note

```{r}
#| label: coverage-summary

n_conditions <- nrow(ref)
n_cat <- ref |> distinct(nmc_category) |> nrow()
```

This reference covers **`r n_conditions` NMC/IDSR conditions** across **`r n_cat` reporting categories**. One NMC condition — *Waterborne illness outbreak – undefined* — has no assigned UCoD ICD-10 code in the source map and is therefore excluded from ICD-based matching (e.g. in the MACOD comparison).
