---
title: "District QA: Death Counts, Crude Rates, and Map Checks"
description: "Simple district-level quality checks: do all districts report deaths, and which districts look high/low after matching deaths to MYPES population?"
date: "2026-06-17"
categories: [district, data-quality, crude-rates, deaths, mapping]
---

```{r}
#| label: setup
#| include: false

suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(flextable)
  library(scales)
  library(sf)
})

rda_candidates <- c(
  "district_veracity_results.rda",
  "projects/district_variable_veracity/district_veracity_results.rda"
)
rda_path <- rda_candidates[file.exists(rda_candidates)][1]

if (is.na(rda_path) || !file.exists(rda_path)) {
  stop("Run projects/district_variable_veracity/wrangling.R first.")
}
load(rda_path)
```

## Objective

This page keeps the checks deliberately simple:

1. plot district death counts,
2. confirm district coverage (which districts have zero deaths),
3. match deaths to district population (MYPES),
4. map crude death rates to flag potential over- or under-reporting areas.

No modelling or causal claims are made.

## Coverage summary

```{r}
#| label: coverage-summary

cat("Analysis year:", qa_summary$analysis_year, "\n")
cat("Districts in map layer:", qa_summary$n_districts_map, "\n")
cat("Districts with recorded deaths:", qa_summary$n_districts_with_deaths, "\n")
cat("Districts with zero recorded deaths:", qa_summary$n_districts_no_deaths, "\n")
cat("Unmatched death-district names (not in map):", qa_summary$n_death_district_names_not_in_map, "\n")
```

## Districts with zero deaths

```{r}
#| label: tbl-zero-deaths
#| tbl-cap: "Table 1. Districts with zero deaths in the analysis year."

zero_dt <- district_coverage[has_deaths == FALSE, .(province, district_name, deaths)]
setorder(zero_dt, province, district_name)

flextable(as.data.frame(zero_dt)) |>
  colformat_double(j = "deaths", digits = 0, big.mark = ",") |>
  theme_booktabs() |>
  bold(part = "header") |>
  autofit()
```

## Death counts by district

```{r}
#| label: fig-counts
#| fig-cap: "Fig. 1. Death counts by district."
#| fig-width: 9
#| fig-height: 8

fig_counts_bar
```

## Crude rate by district

```{r}
#| label: fig-rates
#| fig-cap: "Fig. 2. Crude death rate per 100,000 by district (deaths / MYPES population)."
#| fig-width: 9
#| fig-height: 8

fig_rates_bar
```

## District maps

```{r}
#| label: fig-map-counts
#| fig-cap: "Fig. 3. Map of district death counts."
#| fig-width: 8
#| fig-height: 6

fig_map_counts
```

```{r}
#| label: fig-map-rates
#| fig-cap: "Fig. 4. Map of district crude death rate per 100,000."
#| fig-width: 8
#| fig-height: 6

fig_map_rates
```

## Highest and lowest crude-rate districts

```{r}
#| label: tbl-high-low-rates
#| tbl-cap: "Table 2. Top/bottom 10 districts by crude death rate per 100,000."

high <- district_rate_table[!is.na(crude_rate_per_100k)][1:10,
  .(group = "Highest", province, district_name, deaths, population, crude_rate_per_100k)]
low <- district_rate_table[!is.na(crude_rate_per_100k)][.N-9:.N,
  .(group = "Lowest", province, district_name, deaths, population, crude_rate_per_100k)]
tbl_hl <- rbind(high, low, fill = TRUE)
setorder(tbl_hl, group, -crude_rate_per_100k)

flextable(as.data.frame(tbl_hl)) |>
  colformat_double(j = c("deaths", "population"), digits = 0, big.mark = ",") |>
  colformat_double(j = "crude_rate_per_100k", digits = 1, big.mark = ",") |>
  theme_booktabs() |>
  bold(part = "header") |>
  autofit()
```

## District name mismatches

```{r}
#| label: tbl-unmatched
#| tbl-cap: "Table 3. Death-district names that did not match the district map after standardization."

unmatched_tbl <- deaths_not_in_map
if (nrow(unmatched_tbl) == 0) {
  unmatched_tbl <- data.table(note = "No unmatched district names")
}

flextable(as.data.frame(unmatched_tbl)) |>
  theme_booktabs() |>
  bold(part = "header") |>
  autofit()
```

## Interpretation notes

- These are descriptive QA checks only.
- Districts with zero deaths or extreme crude rates should be reviewed for completeness/coding issues before interpreting as true epidemiological differences.
- Extreme high/low districts can reflect real variation, denominator mismatch, or under/over-reporting; this page is intended to flag where to investigate further.
