---
title: "Years of Life Lost: NMC Conditions in Comparative Perspective"
description: "YLL analysis comparing South Africa's Notifiable Medical Conditions (NMC) — including poisoning — against other injuries, violence, drowning, vaccine-preventable conditions, and cancers. National vital registration data 1997–2022."
date: "2026-03-10"
categories: [YLL, burden-of-disease, NMC, poisoning, injury-mortality, vital-registration, icd10]
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(data.table)
library(ggplot2)
library(flextable)
library(scales)

# ggrepel required for scatter plot — install if absent
if (!requireNamespace("ggrepel", quietly = TRUE)) install.packages("ggrepel")
library(ggrepel)

if (file.exists("yll_results.rda")) {
  load("yll_results.rda")
} else {
  stop("Run yll_wrangling.R first:\n  Rscript projects/yll_comparison/yll_wrangling.R")
}

fmt_n <- function(x) format(round(x), big.mark = ",")
```

## Background

**Years of Life Lost (YLL)** is a summary measure of premature mortality that weights each death by the remaining life expectancy the deceased would have had if they had not died. Unlike simple death counts, YLL captures *how early* deaths occur: a death at age 5 contributes far more YLL than a death at age 75, even though both count as one death.

This analysis applies the **South Africa 2022 national life table** (Statistics South Africa) as the reference standard. For each death in the vital registration dataset, YLL is calculated as the remaining life expectancy at the age of death. Results are summarised across cause groups spanning:

- *External causes* — poisoning, road traffic injuries, drowning, interpersonal violence, self-harm, falls, burns
- *NMC communicable conditions* — tuberculosis, HIV/AIDS, vaccine-preventable diseases, other notifiable infections
- *Chronic non-communicable conditions* — cancers, cardiovascular disease, diabetes

::: {.callout-note}
## Life table note

YLL values use an **approximated** South Africa 2022 combined-sex abridged life table derived from Statistics South Africa mid-year population estimates. Exact single-year values should be substituted from the official StatsSA P0302.3 (2022 edition) when published. All figures may differ slightly from GBD or WHO estimates, which use global frontier reference tables.

The dataset contains deaths from `r analysis_metadata$year_range`. The overall summary table covers all years; the "most recent year" table uses `r analysis_metadata$recent_year` only.
:::

---

## Notifiable Medical Conditions (NMC) — ICD-10 Mapping

The table below summarises South Africa's Notifiable Medical Conditions (NMCs) under the National Health Act (Government Gazette No. 38764, 2015) and their corresponding ICD-10 underlying-cause-of-death codes as expected in Stats SA vital registration data.

```{r}
#| label: tbl-nmc-map
#| tbl-cap: "Table 1. South Africa Notifiable Medical Conditions and corresponding ICD-10 underlying-cause-of-death codes."
#| column: page

if (exists("nmc_map")) {
  ft <- flextable(as.data.frame(nmc_map)) |>
    set_header_labels(
      condition      = "NMC condition",
      icd10_codes    = "ICD-10 UCD codes",
      nmc_category   = "Category",
      notes          = "Vaccine / intervention context"
    ) |>
    bold(part = "header") |>
    theme_booktabs() |>
    fontsize(size = 8.5, part = "all") |>
    color(~ nmc_category == "Environmental / toxic / external",
          color = "#8B0000", part = "body") |>
    bg(~ nmc_category == "Environmental / toxic / external",
       bg = "#FFF5F5", part = "body") |>
    bg(~ nmc_category == "Vaccine-preventable",
       bg = "#F0FFF4", part = "body") |>
    bg(~ nmc_category == "High-burden communicable",
       bg = "#EFF7FF", part = "body") |>
    width(j = 1, width = 2.6) |>
    width(j = 2, width = 1.4) |>
    width(j = 3, width = 1.8) |>
    width(j = 4, width = 3.0) |>
    valign(valign = "top", part = "body") |>
    set_table_properties(layout = "fixed")
  ft
} else {
  message("Run yll_wrangling.R")
}
```

::: {.callout-important}
## Coding caveat for external-cause NMCs

Poisoning deaths in Stats SA vital registration data are subject to systematic **intent misclassification**: because South Africa's death notification form lacks a certified manner-of-death field, ICD-10 Vol. 2 rules assign the *accidental* code by default when intent is unstated. As a result, the X40–X49 (accidental) category substantially over-counts genuinely accidental poisonings, absorbing cases of self-harm and assault. See the [Poisoning Deaths analysis](../poisoning_deaths/index.qmd) for a detailed discussion.
:::

---

## YLL Summary: All Years (`r analysis_metadata$year_range`)

```{r}
#| label: tbl-yll-overall
#| tbl-cap: !expr paste0("Table 2. Total YLL by cause group — all deaths, ", analysis_metadata$year_range, ". SA 2022 national life table (combined sex). 'Other cause' excluded from % calculation.")

if (exists("tbl_yll_overall_display")) {
  flextable(as.data.frame(tbl_yll_overall_display)) |>
    bold(part = "header") |>
    theme_booktabs() |>
    fontsize(size = 9, part = "all") |>
    bg(j = 1, i = ~ `Cause group` == "Poisoning (all intent)",
       bg = "#FFF5F5", part = "body") |>
    bg(j = 1, i = ~ `Cause group` %in% c("Tuberculosis","HIV/AIDS",
       "Vaccine-preventable (excl TB/HIV)","Other NMC communicable"),
       bg = "#EFF7FF", part = "body") |>
    bg(j = 1, i = ~ `Cause group` == "Malignant neoplasms (cancer)",
       bg = "#F5F5F5", part = "body") |>
    width(j = 1, width = 2.8) |>
    width(j = 2:5, width = 1.3) |>
    width(j = 6, width = 1.1) |>
    autofit()
} else {
  message("Run yll_wrangling.R")
}
```

---

## YLL Summary: Most Recent Year (`r if(exists("RECENT_YEAR")) RECENT_YEAR`)

```{r}
#| label: tbl-yll-recent
#| tbl-cap: !expr paste0("Table 3. Total YLL by cause group — ", if(exists("RECENT_YEAR")) RECENT_YEAR, " only. SA 2022 national life table.")

if (exists("tbl_yll_recent_display")) {
  flextable(as.data.frame(tbl_yll_recent_display)) |>
    bold(part = "header") |>
    theme_booktabs() |>
    fontsize(size = 9, part = "all") |>
    bg(j = 1, i = ~ `Cause group` == "Poisoning (all intent)",
       bg = "#FFF5F5", part = "body") |>
    bg(j = 1, i = ~ `Cause group` %in% c("Tuberculosis","HIV/AIDS",
       "Vaccine-preventable (excl TB/HIV)","Other NMC communicable"),
       bg = "#EFF7FF", part = "body") |>
    autofit()
} else {
  message("Run yll_wrangling.R")
}
```

---

## Figures

### Total YLL by cause group

Total YLL summed across all years. The length of each bar reflects both the number of deaths and how prematurely they occurred.

```{r}
#| label: fig-yll-total
#| fig-cap: !expr paste0("Fig. 1. Total YLL by cause group (", analysis_metadata$year_range, "). SA 2022 national life table. 'Other cause' excluded.")
#| fig-width: 9
#| fig-height: 7.5

if (exists("fig_yll_total")) fig_yll_total else message("Run yll_wrangling.R")
```

### Mean YLL per death (prematurity)

Mean YLL per death indicates how premature deaths are within each cause group. Groups at the top die youngest relative to the SA reference life table; groups at the bottom have deaths concentrated at older ages.

```{r}
#| label: fig-mean-yll
#| fig-cap: "Fig. 2. Mean YLL per death by cause group. Higher value = deaths occur at younger ages relative to the SA national life table. SA 2022 national life table (combined sex)."
#| fig-width: 9
#| fig-height: 7.5

if (exists("fig_mean_yll")) fig_mean_yll else message("Run yll_wrangling.R")
```

### Share of total YLL

```{r}
#| label: fig-yll-share
#| fig-cap: "Fig. 3. Percentage share of total YLL by cause group (all years, 'Other cause' excluded). Red shading = NMC poisoning/toxic; blue shading = communicable/NMC; grey = non-NMC chronic."
#| fig-width: 9
#| fig-height: 7.5

if (exists("fig_yll_share")) fig_yll_share else message("Run yll_wrangling.R")
```

### Deaths vs mean YLL: positioning causes across burden dimensions

The scatter plot places each cause group in a two-dimensional space. The x-axis measures total *volume* (deaths), the y-axis measures *prematurity* (mean YLL per death), and bubble area is proportional to *total YLL* — the product of both dimensions.

```{r}
#| label: fig-scatter
#| fig-cap: "Fig. 4. Positioning of cause groups by death volume (x), mean premature years (y), and total YLL (bubble size). Groups in the upper-left cause few but very premature deaths; groups in the lower-right cause many deaths that are less premature. SA 2022 national life table."
#| fig-width: 9
#| fig-height: 7

if (exists("fig_scatter")) fig_scatter else message("Run yll_wrangling.R")
```

### YLL trends over time

Annual YLL for selected cause groups. Years with incomplete registration or changes in coding practice will produce artefactual dips or spikes.

```{r}
#| label: fig-yll-trend
#| fig-cap: !expr paste0("Fig. 5. Annual YLL trends for selected cause groups (", analysis_metadata$year_range, "). SA 2022 national life table.")
#| fig-width: 11
#| fig-height: 5.5

if (exists("fig_yll_trend")) fig_yll_trend else message("Run yll_wrangling.R")
```

### YLL rank over time (bump chart)

Each line tracks the **YLL rank** of a cause group from year to year (rank 1 = the group with the most YLL that year). A rising line means a group is losing relative burden; a falling line means it is gaining ground compared with other causes. This view is particularly useful for policymakers asking *"is this problem getting worse relative to other causes?"*

```{r}
#| label: fig-rank-bump
#| fig-cap: !expr paste0("Fig. 5b. YLL rank by cause group, ", analysis_metadata$year_range, ". Rank 1 = highest YLL that year. Lines connect consecutive annual ranks; right-hand labels show the final-year rank and cause name. 'Other cause' excluded. SA 2022 national life table.")
#| fig-width: 12
#| fig-height: 7.5
#| out-width: "100%"

if (exists("fig_rank_bump")) fig_rank_bump else message("Run yll_wrangling.R")
```

### YLL by age group

The panel charts show the age distribution of YLL for each selected cause group. Because YLL weights young deaths more heavily, the peak of each panel will be shifted younger than the peak of the corresponding death count.

```{r}
#| label: fig-yll-age
#| fig-cap: "Fig. 6. YLL by age group for selected cause groups (all years). SA 2022 national life table. Y-axis scales differ between panels."
#| fig-width: 11
#| fig-height: 11

if (exists("fig_yll_age")) fig_yll_age else message("Run yll_wrangling.R")
```

### YLL by sex (most recent year)

```{r}
#| label: fig-yll-sex
#| fig-cap: !expr paste0("Fig. 7. YLL by cause group and sex, ", if(exists("RECENT_YEAR")) RECENT_YEAR, ". SA 2022 sex-specific life table. Blue = male; red = female.")
#| fig-width: 9
#| fig-height: 6

if (exists("fig_yll_sex")) fig_yll_sex else message("Run yll_wrangling.R")
```

---

## Poisoning YLL breakdown by intent

```{r}
#| label: tbl-poisoning-sub
#| tbl-cap: "Table 4. YLL for poisoning deaths by intent category (all years). Caution: 'Accidental/unspec.' substantially inflated by ICD-10 coding default — see poisoning analysis."

if (exists("yll_poisoning_sub")) {
  tbl_pois <- as.data.frame(yll_poisoning_sub)
  tbl_pois$Deaths  <- format(tbl_pois$Deaths, big.mark = ",")
  tbl_pois$YLL     <- format(tbl_pois$YLL,    big.mark = ",")
  names(tbl_pois) <- c("Intent", "Deaths", "YLL (total)", "Mean YLL/death")
  flextable(tbl_pois) |>
    bold(part = "header") |>
    theme_booktabs() |>
    fontsize(size = 9.5, part = "all") |>
    autofit()
} else {
  message("Run yll_wrangling.R")
}
```

---

## Methods note

### YLL formula

For each death $i$ at age $a_i$:

$$\text{YLL}_i = e(a_i)$$

where $e(a_i)$ is the **remaining life expectancy** from the SA 2022 national life table at age $a_i$. Total YLL for a cause group is $\sum_i \text{YLL}_i$; mean YLL per death is $\bar{\text{YLL}} = \text{Total YLL} / N$.

This follows the standard WHO/GBD approach (Murray & López 1996; GBD 2019), adapted to use a **country-specific** rather than frontier reference life table.

### Cause group assignment

Cause groups are assigned based on the 3-character ICD-10 underlying cause of death code using sequential prefix matching. Groups are **mutually exclusive** (first-match priority). Key boundary decisions:

| Boundary point | ICD-10 codes | Assignment |
|:---|:---|:---|
| Poisoning assault (X85) | Counted with *poisoning*, not with interpersonal violence | `Poisoning (all intent)` |
| Drowning — unintentional only | W65-W74 | `Drowning (unintentional)` |
| Self-harm by drowning / hanging | X71–X84 | `Self-harm (non-poisoning)` |
| Violence by firearm / sharp objects | X86–Y09 | `Interpersonal violence` |
| COVID-19 | U07–U08 | `Other NMC communicable` |
| TB meningitis (A17), TB of other organs | A15–A19 | `Tuberculosis` |

### Data limitations

1. **Intent misclassification** (external causes): see poisoning analysis.
2. **Age missingness**: YLL cannot be calculated where age is unknown; the `% missing age` column in Table 2 shows the fraction excluded per group.
3. **VR coverage**: Stats SA vital registration is known to have incomplete coverage, particularly for younger age groups and provinces with weaker civil registration systems.
4. **Life table provenance**: values used here are provisional approximations. Replace with official Stats SA P0302.3 (2022 edition) for publication.

---

## Next steps

- Add population denominators (StatsSA mid-year estimates) to compute **YLL rates per 100,000**.
- Apply **age-standardisation** to allow comparisons adjusted for South Africa's age structure.
- Disaggregate by **province** and **sex** for the full time series.
- **Sensitivity analysis** using the GBD 2019 global frontier life table (e₀ ≈ 88.9 years) for international comparability.
- Link to programmatic data: TB notifications, ART coverage, vaccination coverage, NCoDV forensic data for external causes.
