---
title: "Birth Registration — Coverage Results"
description: "Population coverage of DHA offices PLUS hospital-based DHA satellites, applying Method 3 (ward-level centroid distance)"
date: "2026-01-20"
categories: [access, DHA, birth registration, results]
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(ggplot2)
library(plotly)
library(flextable)
library(scales)
load("../../outputs/ward_level_dha_access.RData")
```

## Scope

Most births in South Africa occur in health facilities, and a subset of public hospitals host **DHA satellite desks** where birth registration is performed on-site. Birth registration coverage should therefore be computed against the **combined set** of registration nodes:

- `r n_standalone_dha` standalone DHA offices
- `r n_hospital_dha` hospital-based DHA satellites (curated list linked to MFL coordinates via [`DHA_access/link_DHA_to_MFL.r`](https://github.com/bridaybrummer/MRC_VR/blob/main/DHA_access/link_DHA_to_MFL.r))

All numbers use **[Method 3](methodology.qmd)** — ward centroid to nearest registration node — applied to this augmented node set.

## Headline coverage

::: callout-important
- **`r birth_pct_5km`%** of the population lives within **5 km** of a registration node
- **`r birth_pct_10km`%** of the population lives within **10 km** of a registration node
- **`r birth_pct_20km`%** of the population lives within **20 km** of a registration node
:::

## Coverage by distance band

```{r}
#| label: tbl-birth-coverage
birth_access_ft
```

## Birth vs death registration: side by side

```{r}
#| label: tbl-birth-vs-death
birth_vs_death_ft
```

Adding the `r n_hospital_dha` hospital satellites raises 5 km coverage by **`r round(birth_pct_5km - m3_pct_5km, 1)` percentage points** — a meaningful gain in areas where mothers would otherwise have to travel beyond their birthing facility to register the birth.

## Interactive map: registration nodes

Standalone DHA offices in **red**; hospital-based DHA satellites in **blue**. Hover for facility detail.

```{r}
#| label: fig-birth-nodes-plotly
#| fig-width: 10
#| fig-height: 7
birth_nodes_plotly
```

## Map: distance to nearest registration node

```{r}
#| label: fig-birth-map
#| fig-width: 10
#| fig-height: 8
birth_distance_map
```

## Cumulative coverage: births vs deaths

```{r}
#| label: fig-birth-vs-death-curve
#| fig-width: 10
#| fig-height: 6
birth_vs_death_curve_plot
```

## Notes on the hospital DHA linkage

The hospital-list-to-MFL match (see `DHA_access/dha_mfl_matched.csv`) is fuzzy and includes weak matches flagged for manual review. Coverage figures should be interpreted as the **upper bound under the current linkage** — manual curation of the *Weak — verify* matches and follow-up on unmatched private facilities will refine the estimate.

## Limitations

::: callout-warning
- **Service hours**: hospital DHA desks may have limited operating hours; we assume full availability.
- **Linkage quality**: ~9% of matches are flagged "Weak — verify". See `dha_mfl_unmatched.csv` and `dha_mfl_matched.csv` for the audit trail.
- All Method 3 limitations apply — centroid proxy, straight-line distance, static population. See [Methodology](methodology.qmd).
:::

For methodology, see [Methodology](methodology.qmd). For a check that DHA offices are in plausibly dense locations, see the [Sensitivity check](sensitivity.qmd). For alternative methods, see [Methods Comparison](methods_comparison.qmd).
