---
title: "Death Registration — Coverage Results"
description: "Population coverage of standalone DHA offices, applying Method 3 (ward-level centroid distance)"
date: "2026-01-20"
categories: [access, DHA, death registration, results]
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(ggplot2)
library(flextable)
library(scales)
load("../../outputs/ward_level_dha_access.RData")
```

## Scope

Death registration in South Africa takes place at **standalone DHA offices** (`r n_standalone_dha` offices with geocoded locations in the 2024 dataset). Unlike births, deaths are not commonly registered at health facilities, so the hospital-based DHA satellites are **not** included here — they are added in the [birth registration analysis](birth_registration.qmd).

All numbers below use **[Method 3](methodology.qmd)** — ward centroid to nearest DHA office.

## Headline coverage

::: callout-important
- **`r m3_pct_5km`%** of the population lives within **5 km** of a DHA office
- **`r m3_pct_10km`%** of the population lives within **10 km** of a DHA office
- **`r m3_pct_20km`%** of the population lives within **20 km** of a DHA office
- `r missing_pop` wards had missing population data and were excluded
:::

## Coverage by distance band

```{r}
#| label: tbl-death-coverage
ward_access_ft
```

## Map: distance to nearest DHA office

```{r}
#| label: fig-death-map
#| fig-width: 10
#| fig-height: 8
ward_distance_map
```

## Cumulative coverage curve

```{r}
#| label: fig-death-curve
#| fig-width: 10
#| fig-height: 6
coverage_curve_plot
```

## What this means for death registration

The headline figure — roughly **`r m3_pct_10km`%** of the population within 10 km of a DHA office — is the working population covered for death registration under realistic same-day travel. Coverage drops markedly in rural districts, which is where under-registration of deaths is most plausible.

The [sensitivity check](sensitivity.qmd) confirms that DHA offices sit, on average, in denser-than-average areas — consistent with population-weighted placement.

## Limitations

::: callout-warning
- **Centroid proxy**: large rural wards may misrepresent where people actually live.
- **Straight-line distance**: real travel cost is higher in mountainous / poorly connected districts.
- **Office capacity** is not modelled — overcrowding at busy urban offices can be a real access barrier even with low geographic distance.
- **Static population**: 2025 mid-year estimates may understate informal settlement growth.
:::

For a side-by-side comparison with alternative methods, see [Methods Comparison](methods_comparison.qmd).
