---
title: "DHA Service Access and Coverage"
description: "Population-weighted access to Department of Home Affairs offices for birth and death registration in South Africa"
date: "2026-01-20"
categories: [access, spatial, DHA, registration]
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(flextable)
ward_output <- "../../outputs/ward_level_dha_access.RData"
if (!file.exists(ward_output)) {
  message("Running ward-level wrangling script...")
  source("../../ward_level_dha_access_script.R", local = new.env())
}
load(ward_output)
```

## Overview

Birth and death registration in South Africa happens at Department of Home Affairs (DHA) offices. **Access to these offices is a precondition for complete vital registration**, and poor geographic access is a plausible driver of under-registration in remote areas.

This project answers: **what proportion of the South African population lives within reach of a DHA registration node?** Coverage estimates are reported separately for **death registration** (DHA offices only) and **birth registration** (DHA offices + hospital-based DHA satellites, since most births occur in facilities).

## How the analysis is organised

::: {.callout-tip}
## Preferred method
All headline results use **Method 3 — ward-level centroid distance**. Population is taken from observed Stats SA 2025 ward-level estimates; distance is the straight-line distance from each ward centroid to the nearest registration node.
:::

| Page | Purpose |
|---|---|
| [Methodology — Method 3](methodology.qmd) | How ward-level access is computed and why Method 3 is preferred |
| [Methods Comparison](methods_comparison.qmd) | Side-by-side comparison of three approaches (district inter-office, modelled grid, ward centroid) |
| [Sensitivity: 500 m density check](sensitivity.qmd) | Are DHA offices in plausibly dense areas? Modelled population density within 500 m of each node, benchmarked against the ward-density distribution |
| [**Death registration coverage**](death_registration.qmd) | Final results — DHA offices only |
| [**Birth registration coverage**](birth_registration.qmd) | Final results — DHA offices + hospital-based DHA satellites (from the DHA↔MFL linkage) |

## Headline numbers (Method 3)

```{r}
#| label: tbl-headline
#| echo: false
tibble::tribble(
  ~Indicator,                                ~`Death reg. (DHA only)`,         ~`Birth reg. (DHA + hospital)`,
  "% population within 5 km of a node",      paste0(m3_pct_5km,  "%"),         paste0(birth_pct_5km,  "%"),
  "% population within 10 km of a node",     paste0(m3_pct_10km, "%"),         paste0(birth_pct_10km, "%"),
  "% population within 20 km of a node",     paste0(m3_pct_20km, "%"),         paste0(birth_pct_20km, "%"),
  "Number of registration nodes",            as.character(n_standalone_dha),   as.character(n_standalone_dha + n_hospital_dha)
) |>
  flextable() |>
  bold(part = "header") |>
  bg(j = "Birth reg. (DHA + hospital)", bg = "#E8F5E9") |>
  theme_zebra() |>
  autofit() |>
  set_caption("National coverage — Method 3 (ward centroid to nearest registration node)")
```

## Data sources

- **DHA office locations** — 2024 DHA offices dataset (`r n_standalone_dha` standalone offices with coordinates)
- **Hospital-based DHA satellites** — `r n_hospital_dha` public hospitals matched to MFL coordinates via the linkage in `DHA_access/link_DHA_to_MFL.r`
- **Population** — Stats SA mid-year ward-level estimates (2025)
- **Boundaries** — MDB 2020 ward shapefile (4,468 wards)

## Limitations

- Ward centroids are proxies for population locations; large rural wards may have centroids far from where people actually live.
- Distances are straight-line, not road-network or travel time.
- Hospital satellites are assumed to offer the relevant registration service whenever they appear on the curated list — actual service hours and uptake are not modelled.
- Office capacity and service quality are not accounted for.

See each child page for method-specific limitations.
