---
title: "Methods Comparison"
description: "Four approaches to estimating DHA office population coverage, side by side"
date: "2026-01-20"
categories: [access, spatial, DHA, methods]
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(ggplot2)
library(plotly)
library(flextable)
library(scales)
load("../../outputs/ward_level_dha_access.RData")
```

# Background

We evaluated **four progressively refined approaches** for estimating the share of the South African population within reach of a DHA office. The headline reporting on [death registration](death_registration.qmd) and [birth registration](birth_registration.qmd) uses **Method 3**. This page documents the trade-offs.

# Shared data

- DHA office locations — 2024 DHA offices dataset
- Population — Stats SA mid-year estimates (2025)
- Boundaries — Districts (Stats SA) and Wards (MDB 2020)

# National overview

::: panel-tabset

## Population by district

```{r}
#| label: SA-pop-map
SA_pop
```

## Number of offices by district

```{r}
#| label: SA-offices-map
SA_N_off
```

## Office locations (interactive)

```{r}
#| label: office-locations-interactive
plotly::ggplotly(p_with_buffers, tooltip = "text")
```

:::

------------------------------------------------------------------------

# Method 1 — District-level office-to-office distance

For each district, take the median distance between each DHA office and its nearest neighbour. Halve it as a proxy for the typical one-way travel a resident must make. This single distance is applied to the entire district population.

## Results

```{r}
#| label: method1-results
m1_table
```

::: callout-important
Coverage: **`r m1_national$pct_5km`%** within 5 km, **`r m1_national$pct_10km`%** within 10 km, **`r m1_national$pct_20km`%** within 20 km.
:::

::: callout-warning
**Weakness** — one estimate per district ignores within-district variation. Underestimates coverage where most people live near one office.
:::

------------------------------------------------------------------------

# Method 2 — District-level modelled population grid

Build a 1 km grid across each district. Distribute the district population to grid points using an inverse-power decay around DHA offices ($w_i = 1/(d_i + 1)^{\alpha}$ with $\alpha = 1.5$). Distance is computed from each grid point to the nearest office.

## Modelled population

```{r}
#| label: modelled-population-map
if (exists("modelled_population_map")) modelled_population_map
```

## Coverage table

```{r}
#| label: method2-table
dha_access_summary_table
```

::: callout-important
Coverage: **`r m2_pct_5km`%** within 5 km, **`r m2_pct_10km`%** within 10 km, **`r m2_pct_20km`%** within 20 km.
:::

::: callout-warning
**Weakness** — population is modelled, not observed. The decay assumption (people cluster near offices) is exactly what we're trying to evaluate, so the method is partly circular for the policy question.
:::

------------------------------------------------------------------------

# Method 3 — Ward-level centroid distance (preferred) {#preferred}

Use **observed ward-level population data** (Stats SA 2025) for 4,468 wards. Distance is computed from each ward centroid to the nearest DHA office.

See [Methodology](methodology.qmd) for the full specification.

## Coverage by distance band

```{r}
#| label: method3-coverage-table
ward_access_ft
```

## Distance map

```{r}
#| label: method3-distance-map
#| fig-width: 10
#| fig-height: 8
ward_distance_map
```

## Cumulative coverage

```{r}
#| label: method3-cumulative-plot
#| fig-width: 10
#| fig-height: 6
coverage_curve_plot
```

::: callout-important
Coverage: **`r m3_pct_5km`%** within 5 km, **`r m3_pct_10km`%** within 10 km, **`r m3_pct_20km`%** within 20 km.
:::

::: callout-warning
**Weakness** — ward centroids are still a proxy. The [sensitivity check](sensitivity.qmd) tests how plausible the implied population distribution near DHA nodes is.
:::

------------------------------------------------------------------------

# Method 4 — WorldPop Grid + OSM Road Network (Western Cape) {#method4}

100m-resolution **observed** (satellite-derived) population from [WorldPop](https://www.worldpop.org/) is routed to the nearest DHA office over the actual **OpenStreetMap road network**, giving road-distance and driving-time coverage rather than straight-line distance. Implemented for the Western Cape only, in the companion [redistribute_facility_deaths](https://github.com/bridaybrummer/redistribute_facility_deaths) pipeline.

## Results (Western Cape, death registration only)

```{r}
#| label: method4-results

# Pre-computed from redistribute_facility_deaths pipeline (2026-06-09)
m4_results <- tibble::tribble(
  ~Measure, ~`< 5 km`, ~`< 10 km`, ~`< 20 km`,
  "Straight-line distance", "48.6%", "74.5%", "86.5%",
  "Road-network distance", "36.0%", "71.1%", "83.3%"
)

m4_time <- tibble::tribble(
  ~Measure, ~`< 10 min`, ~`< 30 min`, ~`< 60 min`,
  "Driving time", "71.9%", "91.1%", "97.6%"
)

flextable(m4_results) |>
  bold(part = "header") |>
  bg(i = 2, bg = "#FFF3E0") |>
  theme_zebra() |>
  autofit() |>
  set_caption("Method 4 — Population within distance of nearest DHA office (Western Cape)")

```

```{r}
#| label: method4-time-results

flextable(m4_time) |>
  bold(part = "header") |>
  bg(i = 1, bg = "#FFF3E0") |>
  theme_zebra() |>
  autofit() |>
  set_caption("Method 4 — Population within driving time of nearest DHA office (Western Cape)")
```

::: callout-important
Road distance: **36.0%** within 5 km, **71.1%** within 10 km, **83.3%** within 20 km. Driving time: **91.1%** within 30 minutes, **97.6%** within 60 minutes (death registration, DHA offices only, Western Cape).
:::

::: callout-warning
**Weakness** — Western Cape only; not yet run nationally. Assumes free-flow road speeds; excludes residential/unclassified streets; no birth-registration (hospital satellite) analysis has been run against this method.
:::

------------------------------------------------------------------------

# Side-by-side comparison

## Qualitative comparison

```{r}
#| label: comparison-table-qualitative

comparison_qual <- tibble::tribble(
  ~Feature, ~`Method 1`, ~`Method 2`, ~`Method 3`, ~`Method 4`,
  "Geographic unit", "District", "1 km grid within district", "Ward (~4,400 units)", "100 m population cell",
  "Population data", "District totals", "District totals (modelled to grid)", "Ward-level counts (observed)", "WorldPop satellite grid (observed)",
  "Population distribution", "Assumed uniform across district", "Modelled (inverse power decay near offices)", "Assigned to ward centroid", "Observed satellite-derived raster",
  "Distance measure", "Half-median inter-office distance", "Grid point to nearest office", "Ward centroid to nearest office", "Road-network distance & driving time",
  "Birth vs death split", "No", "No", "Yes (hospital satellites added)", "No",
  "Geographic coverage", "National", "National", "National", "Western Cape only",
  "Key strength", "Simple; no spatial computation", "Models within-district variation", "Uses actual sub-district population data", "Real road distances & travel times",
  "Key weakness", "Single estimate per district; ignores clustering", "Population model may not reflect reality", "Centroid may not represent large rural wards", "Not yet run nationally",
  "Computational cost", "Low", "High (1 km grid across SA)", "Moderate", "Very high (OSM routing)"
)

flextable(comparison_qual) |>
  bold(part = "header") |>
  bg(j = "Method 3", bg = "#E8F5E9") |>
  bg(j = "Method 4", bg = "#FFF3E0") |>
  theme_zebra() |>
  autofit() |>
  set_caption("Qualitative comparison of the four DHA access methodologies")
```

## Coverage results comparison

```{r}
#| label: comparison-results-table-all

comparison_res <- tibble::tribble(
  ~Metric, ~`Method 1`, ~`Method 2`, ~`Method 3`, ~`Method 4 (road)`,
  "% within 5 km",  paste0(m1_national$pct_5km, "%"),  paste0(m2_pct_5km, "%"),  paste0(m3_pct_5km, "%"),  "36.0%",
  "% within 10 km", paste0(m1_national$pct_10km, "%"), paste0(m2_pct_10km, "%"), paste0(m3_pct_10km, "%"), "71.1%",
  "% within 20 km", paste0(m1_national$pct_20km, "%"), paste0(m2_pct_20km, "%"), paste0(m3_pct_20km, "%"), "83.3%"
)

flextable(comparison_res) |>
  bold(part = "header") |>
  bg(j = "Method 3", bg = "#E8F5E9") |>
  bg(j = "Method 4 (road)", bg = "#FFF3E0") |>
  theme_zebra() |>
  autofit() |>
  set_caption("Population completeness (% within distance threshold) — Methods 1–3 are national, Method 4 is Western Cape only")
```

::: callout-note
Methods 1–3 use **straight-line** distance and cover **all of South Africa**. Method 4 uses **road-network** distance and covers **Western Cape only** — the figures are not directly comparable. Road distances are consistently higher than straight-line at the same threshold.
:::

::: callout-tip
## Which method should I use?

- **District-level dashboards**: Method 1 — simple and conservative.
- **Sub-district modelled estimates**: Method 2 — but be aware the population distribution is itself modelled.
- **National policy reporting**: **Method 3** — uses observed ward populations and provides the finest available spatial resolution.
- **Maximum physical accuracy (Western Cape)**: **Method 4** — road-network routing with observed satellite-derived population; currently limited to one province.
:::
