---
title: "Methods Comparison"
description: "Three approaches to estimating DHA office population coverage, side by side"
date: "2026-01-20"
categories: [access, spatial, DHA, methods]
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(ggplot2)
library(plotly)
library(flextable)
library(scales)
load("../../outputs/ward_level_dha_access.RData")
```

# Background

We evaluated **three progressively refined approaches** for estimating the share of the South African population within reach of a DHA office. The headline reporting on [death registration](death_registration.qmd) and [birth registration](birth_registration.qmd) uses **Method 3**. This page documents the trade-offs.

# Shared data

- DHA office locations — 2024 DHA offices dataset
- Population — Stats SA mid-year estimates (2025)
- Boundaries — Districts (Stats SA) and Wards (MDB 2020)

# National overview

::: panel-tabset

## Population by district

```{r}
#| label: SA-pop-map
SA_pop
```

## Number of offices by district

```{r}
#| label: SA-offices-map
SA_N_off
```

## Office locations (interactive)

```{r}
#| label: office-locations-interactive
plotly::ggplotly(p_with_buffers, tooltip = "text")
```

:::

------------------------------------------------------------------------

# Method 1 — District-level office-to-office distance

For each district, take the median distance between each DHA office and its nearest neighbour. Halve it as a proxy for the typical one-way travel a resident must make. This single distance is applied to the entire district population.

## Results

```{r}
#| label: method1-results
m1_table
```

::: callout-important
Coverage: **`r m1_national$pct_5km`%** within 5 km, **`r m1_national$pct_10km`%** within 10 km, **`r m1_national$pct_20km`%** within 20 km.
:::

::: callout-warning
**Weakness** — one estimate per district ignores within-district variation. Underestimates coverage where most people live near one office.
:::

------------------------------------------------------------------------

# Method 2 — District-level modelled population grid

Build a 1 km grid across each district. Distribute the district population to grid points using an inverse-power decay around DHA offices ($w_i = 1/(d_i + 1)^{\alpha}$ with $\alpha = 1.5$). Distance is computed from each grid point to the nearest office.

## Modelled population

```{r}
#| label: modelled-population-map
if (exists("modelled_population_map")) modelled_population_map
```

## Coverage table

```{r}
#| label: method2-table
dha_access_summary_table
```

::: callout-important
Coverage: **`r m2_pct_5km`%** within 5 km, **`r m2_pct_10km`%** within 10 km, **`r m2_pct_20km`%** within 20 km.
:::

::: callout-warning
**Weakness** — population is modelled, not observed. The decay assumption (people cluster near offices) is exactly what we're trying to evaluate, so the method is partly circular for the policy question.
:::

------------------------------------------------------------------------

# Method 3 — Ward-level centroid distance (preferred) {#preferred}

Use **observed ward-level population data** (Stats SA 2025) for 4,468 wards. Distance is computed from each ward centroid to the nearest DHA office.

See [Methodology](methodology.qmd) for the full specification.

## Coverage by distance band

```{r}
#| label: method3-coverage-table
ward_access_ft
```

## Distance map

```{r}
#| label: method3-distance-map
#| fig-width: 10
#| fig-height: 8
ward_distance_map
```

## Cumulative coverage

```{r}
#| label: method3-cumulative-plot
#| fig-width: 10
#| fig-height: 6
coverage_curve_plot
```

::: callout-important
Coverage: **`r m3_pct_5km`%** within 5 km, **`r m3_pct_10km`%** within 10 km, **`r m3_pct_20km`%** within 20 km.
:::

::: callout-warning
**Weakness** — ward centroids are still a proxy. The [sensitivity check](sensitivity.qmd) tests how plausible the implied population distribution near DHA nodes is.
:::

------------------------------------------------------------------------

# Side-by-side comparison

```{r}
#| label: comparison-table
comparison_ft
```

```{r}
#| label: comparison-results-table
comparison_results_ft
```

::: callout-tip
## Which method should I use?

- **District-level dashboards**: Method 1 — simple and conservative.
- **Sub-district modelled estimates**: Method 2 — but be aware the population distribution is itself modelled.
- **National policy reporting**: **Method 3** — uses observed ward populations and provides the finest available spatial resolution.
- **Maximum accuracy**: enumeration-area census population + road-network distances. Not currently available.
:::
