---
title: "Monitoring and Evaluation Dashboard"
description: "Comparing vital registration data with Tembisa model and NCODV-corrected estimates for HIV/AIDS, TB, and poisoning mortality, plus data quality indicators"
date: "2026-01-20"
categories: [mortality, HIV, TB, poisoning, data-quality, timeliness, non-hospital]
bibliography: references.bib
csl: ../../vancouver.csl
---

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(plotly)
library(htmlwidgets)
library(arrow)
library(data.table)
library(flextable)
library(conflicted)
library(DT)
library(readxl)
library(haven)
library(scales)

# Resolve common conflicts between lubridate and data.table
conflicts_prefer(
  data.table::hour,
  data.table::minute,
  data.table::second,
  data.table::year,
  data.table::month,
  data.table::mday,
  data.table::wday
)

# Resolve conflicts between plotly and graphics
conflicts_prefer(
  plotly::layout,
  plotly::style
)

# -------------------------------------------------------------------------
# Standardized data-quality TARGET (single source of truth for all figures)
# Literature basis: WHO/GBD cause-of-death usability convention.
#   - >20% of deaths assigned to ill-defined / garbage codes indicates
#     LOW-quality, low-usability cause-of-death data.
#   - <10% is the "high quality" benchmark.
# Primary sources: Mathers et al. (2005), Bull WHO 83(3):171-7 (ill-defined
#   % classification); Mahapatra et al. (2007), Lancet 370:1653-63 (CRVS
#   series); Naghavi et al. (2010), Popul Health Metr 8:9 (garbage codes).
# Target is rendered consistently as a RED DOTTED line + RED STAR (shape 8).
# -------------------------------------------------------------------------
dq_target          <- 20          # percent (usability upper bound)
dq_target_high     <- 10          # percent (high-quality benchmark, in text)
dq_target_colour   <- "#D32F2F"   # standardized red
dq_target_lty      <- "dotted"

# Auditability: capture render timestamp once
render_stamp <- format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")

# -------------------------------------------------------------------------
# Standardized 10-YEAR reporting window (M+E review, Aug 2026):
# "Dashboard to 10 years across the board" - restrict all trend figures to
# the most recent 10 death-year cohorts unless a figure needs longer history
# for methodological reasons (explicitly noted where that applies).
# -------------------------------------------------------------------------
window_last_year  <- 2022L
window_first_year <- window_last_year - 9L   # 2013-2022 inclusive = 10 years
window_years       <- window_first_year:window_last_year

# Source the wrangling script if outputs are missing or stale
output_file <- "../../outputs/me_plots.RData"
wrangling_script <- "../../ME_wrangling.r"

# Check if we need to re-run wrangling
if (!file.exists(output_file)) {
  message("Running wrangling script...")
  source(wrangling_script, local = new.env())
}

# Load the pre-built plots
load(output_file)

# Source dashboard helpers for excess deaths functions
helpersPath <- "../../dashboard/dashboard_helpers.r"
if (file.exists(helpersPath)) {
  source(helpersPath)
}

# Load analysis data for excess deaths calculations
analysisPath <- "../../LGH_AnalysisData_2022.feather"
if (file.exists(analysisPath)) {
  dt_analysis <- read_feather(analysisPath) %>% as.data.table()
}

# Load external causes data (unnatural deaths)
externalPath <- "../../Deaths2022_MRCversionFINAL.feather"  
if (file.exists(externalPath)) {
  dt_external <- read_feather(externalPath) %>% as.data.table()
  # Filter to external causes only (V, W, X, Y codes)
  dt_external <- dt_external[grepl("^[VWXY]", UnderlyingCause)]
} else {
  dt_external <- NULL
}
```

## Overview

::: {.callout-note appearance="minimal"}
**Report generated:** `r render_stamp` &nbsp;·&nbsp; *Timestamp shown for auditability — re-rendering this dashboard refreshes it against the latest source data.*
:::

This dashboard compares vital registration (VR) data with alternative mortality data sources for monitoring and evaluation purposes:

- **Tembisa Model**: Demographic projections for HIV/AIDS and TB mortality
- **NCODV Study**: Verbal autopsy-corrected estimates (2017-2018)
- **NMC Database**: Notifiable Medical Conditions surveillance data for TB and poisoning

## South Africa CRVS Monitoring Indicators (Cause-of-Death Focus) {#crvs-compendium}

The table below is sourced from the M+E team verified **V4.0 indicator compendium workbook** (September 2025) and is the authoritative SA CRVS cause-of-death indicator list used for this dashboard. It includes the standard indicator definition, the verified numerator, the verified denominator, and the applicable disaggregation level for each indicator. Use the column filters and search box to explore indicators by CRVS milestone, category, or indicator definition. The sections that follow track dashboard performance against this framework, starting with registration timeliness.

```{r}
#| label: crvs-compendium-table
#| echo: false

# Source: official M+E verified V4 workbook. We import the "Indicator List"
# sheet directly and preserve the workbook's CRVS-pathway ordering. The row
# number is generated to keep the table sequential while the V4 workbook's
# natural ordering is retained.
compendium_raw <- read_excel(
  "20250930_CRVSPerfMonIndCompendium_V4.0.xlsx",
  sheet = "Indicator List",
  skip = 1
)
compendium_raw <- compendium_raw[!is.na(compendium_raw$Indicator), ]
compendium_raw$No <- seq_len(nrow(compendium_raw))

compendium_display <- compendium_raw[, c(
  "No",
  "ID",
  "Unique ID",
  "Indicator",
  "Category",
  "CRVS milestone",
  "Recommended prioritization",
  "Disaggregation level",
  "Numerator",
  "Denominator",
  "Definition"
)]

colnames(compendium_display) <- c(
  "No",
  "Indicator ID",
  "Unique ID",
  "Indicator",
  "Category",
  "CRVS milestone",
  "Priority",
  "Disaggregation level",
  "Numerator",
  "Denominator",
  "Definition"
)

datatable(
  compendium_display,
  filter     = "top",
  extensions = "Buttons",
  rownames   = FALSE,
  options    = list(
    dom        = "Bfrtip",
    buttons    = c("csv", "excel"),
    pageLength = 15,
    scrollX    = TRUE,
    autoWidth  = FALSE,
    order      = list(list(0, "asc")),
    columnDefs = list(
      list(width = "40px",  targets = 0),   # No
      list(width = "120px", targets = 1),   # Indicator ID
      list(width = "120px", targets = 2),   # Unique ID
      list(width = "280px", targets = 3),   # Indicator
      list(width = "160px", targets = 5),   # CRVS milestone
      list(width = "180px", targets = 7),   # Disaggregation level
      list(width = "220px", targets = 8),   # Numerator
      list(width = "220px", targets = 9),   # Denominator
      list(width = "360px", targets = 10)    # Definition
    )
  )
)
```

---

## Report Timeliness {#timeliness}

Stats SA releases mortality reports with a delay of several years. This section tracks the timeliness of releases.

**Click on any point to open the corresponding Stats SA PDF report.**

```{r}
#| label: timeliness-plot

p_timeliness_interactive
```

### Timeliness vs. Completeness {#timeliness-completeness}

::: {.callout-warning icon=false}
## ⚠ Placeholder — content pending
This subsection is reserved for a **timeliness vs. completeness** figure/summary (distinct from the late-transfer cohort analysis below). **Not yet populated** — requires source data/methodology confirmation from the M+E team before this placeholder is replaced with a real figure.
:::

### Timeliness vs. Late-Transfer Completeness {#reg-tradeoff}

Each death record carries a **date of death** and a **date of registration**. The gap between them — the *registration lag* — drives a fundamental tradeoff: publishing statistics soon after the death year is **timely** but misses deaths transferred/registered late, while waiting longer captures those late transfers and is more **complete**. Using registration timing as a proxy, the curve below shows, for each death-year cohort, the cumulative share of its registered deaths captured by *N* years after death.

::: {.callout-important appearance="simple"}
The **2022 death cohort is not yet complete**: late transfers/registrations for 2022 deaths are still arriving, so its curve (like 2021's) is right-truncated and will keep rising slightly with future extracts.
:::

```{r}
#| label: reg-tradeoff-plot
#| echo: false
#| warning: false
#| message: false

reg_cols <- c("DeathYear", "DeathMonth", "DeathDay",
              "RegistrationYear", "RegistrationMonth", "RegistrationDay")

reg_raw <- read_feather("../../LGH_MasterFile_preCollapsed2022.feather",
                        col_select = reg_cols) |> as.data.table()
reg_raw[, (reg_cols) := lapply(.SD, function(x) as.integer(zap_labels(x))),
        .SDcols = reg_cols]

reg_raw[, raw_lag := RegistrationYear - DeathYear]
reg_raw[, lag_yr := pmax(raw_lag, 0L)]

# --- Data-quality flags to surface as limitations / checks ---
flag_partial2013 <- reg_raw[DeathYear == 2013, .N]      # partial cohort in extract
flag_neg_lag     <- reg_raw[raw_lag < 0L, .N]           # registered before death (impossible)
flag_floored     <- reg_raw[raw_lag < 0L, .N]           # records floored to lag 0
flag_late5       <- reg_raw[raw_lag > 5L, .N]           # very late (>5 yr) registrations

# Last-10-years window (2013-2022); 2013 remains excluded as a partial cohort,
# leaving 9 full death-year cohorts (2014-2022) - see limitations callout below.
reg_dt   <- reg_raw[DeathYear %in% window_years[window_years != 2013]]
max_reg  <- 2024L   # latest registration year present in the extract
max_lag  <- 5L

cohort_n <- reg_dt[, .(N = .N), by = DeathYear]
cum      <- reg_dt[, .(n = .N), by = .(DeathYear, lag = pmin(lag_yr, max_lag))]
grid     <- CJ(DeathYear = 2014:2022, lag = 0:max_lag)
cum      <- cum[grid, on = .(DeathYear, lag)][is.na(n), n := 0L]
setorder(cum, DeathYear, lag)
cum[cohort_n, N := i.N, on = "DeathYear"]
cum[, cum_pct := cumsum(n) / N * 100, by = DeathYear]

# Draw each cohort only up to its observed maximum lag (right-truncation honesty)
cum[, obs_max_lag := max_reg - DeathYear]
cum_plot <- cum[lag <= obs_max_lag]
cum_plot[, tt := paste0("Death cohort: ", DeathYear,
                        "<br>Within ", lag, " yr of death: ",
                        sprintf("%.2f%%", cum_pct))]

p_tradeoff <- ggplot(cum_plot,
                     aes(lag, cum_pct, colour = DeathYear,
                         group = DeathYear, text = tt)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.6) +
  scale_colour_viridis_c(name = "Death\nyear", breaks = seq(2014, 2022, 2)) +
  scale_x_continuous(breaks = 0:max_lag) +
  coord_cartesian(ylim = c(floor(min(cum_plot$cum_pct)), 100)) +
  labs(
    title    = "Timeliness vs. late-transfer completeness, by death-year cohort",
    subtitle = "Cumulative % of each cohort registered by N years after death\nDashed line = 1-year publication cut-off (stricter than the WHO 2-year CRVS completeness benchmark); rise to its right = completeness gained from late transfers",
    x        = "Years after death (registration lag)",
    y        = "Cumulative % of cohort registered"
  ) +
  theme_minimal(base_size = 12)

ggplotly(p_tradeoff, tooltip = "text") |>
  style(connectgaps = TRUE) |>
  layout(hovermode = "closest")
```

::: {.callout-note}
## Interpreting the Tradeoff
- The **height at lag 0** is the *timely* share — deaths registered in the same calendar year (≈ 98.5–99.2% across cohorts), so South African VR registration is highly timely.
- The **rise between lag 0 and lag 1** is the *completeness gain* from late transfers captured by waiting one extra year; it adds only ≈ 0.6–1.3 percentage points, after which the curves plateau.
- **Tradeoff implication:** publishing one year after the death year (dashed line) captures ≈ 99.8% of all eventually-registered deaths. This 1-year cut-off is **stricter** than the WHO-recommended 2-year interval for CRVS completeness assessment, so the completeness penalty for timely release is small.
- Recent cohorts (2021–2022) are **right-truncated** — they can only be observed to registration year `r max_reg` — so their curves stop at a shorter lag, and **2022 is not yet a complete cohort**.
:::

::: {.callout-warning icon=false}
## ⚠ Data limitations & checks to investigate
These handling decisions are applied above so the plot is honest, but each one flags a record-level issue that should be **checked at source** rather than silently corrected:

- **2013 is a partial cohort** (`r format(flag_partial2013, big.mark=",")` records) and is **excluded** from cohort comparisons — its same-year registration share (≈ 54%) reflects a truncated extract window, not poor timeliness. *Check:* confirm the extract's start date and whether earlier 2013 deaths live in a separate file.
- **Negative registration lags** (`r format(flag_neg_lag, big.mark=",")` record(s)) imply a death registered *before* it occurred — logically impossible. These are **floored to lag 0** for plotting. *Check:* audit these records for transposed or mis-keyed death/registration dates.
- **Very late registrations** (>5 years: `r format(flag_late5, big.mark=",")` records) are **capped at lag 5** in the cumulative curve. *Check:* verify whether these are genuine late registrations (e.g. delayed/contested deaths) or data-entry errors in the year fields.

None of these materially change the headline (≈ 99.8% within one year), but they should be triaged as part of routine VR data-quality control.
:::

## Data Quality: Garbage Codes {#garbage-codes}

Garbage codes are ICD-10 codes that provide insufficient information about the underlying cause of death. Monitoring the proportion of non-garbage codes is essential for assessing data quality.

```{r}
#| label: dq-data-load
#| include: false

dt_raw <- read_feather("../../LGH_MasterFile_preCollapsed2022.feather") |>
  as.data.table()

dt_raw[, `:=`(
  sex           = as_factor(sex),
  DeathInst     = as_factor(DeathInst),
  DeathProvince = as_factor(DeathProvince),
  Ascertainment = as_factor(Ascertainment)
)]

# Facility-type classification of DeathInst (8 raw levels), used throughout
# the "Non-Hospital Deaths" section. This mapping is a judgment call, not an
# official CRVS standard — see the definitions callout under "Non-Hospital
# Deaths" for the full rationale on each category, in particular:
#   - Nursing home -> Out-of-hospital: matches this dashboard's framing of
#     "non-hospital" deaths (at home, nursing homes, etc.) and reflects that
#     these are not acute in-hospital deaths.
#   - Dead on arrival -> Out-of-hospital: the death itself occurred before/
#     outside hospital care, even though it was certified at a facility.
#   - Unknown/Unspecified -> its own explicit "Unspecified/Unknown" level
#     (never silently dropped as NA) so it is always visible in tables and
#     is excluded, with that exclusion stated, from the non-hospital % trend.
dt_raw[, facility_type := fcase(
  DeathInst %in% c("Hospital", "Emergency room/Outpatient"),      "In-hospital",
  DeathInst %in% c("Home", "Nursing home", "Dead on arrival",
                    "Other"),                                     "Out-of-hospital",
  DeathInst %in% c("Unknown", "Unspecified"),                     "Unspecified/Unknown",
  default = "Unspecified/Unknown"
)]

# Data-quality flags per the confirmed 5-category GBD/Naghavi unusable-code
# crosswalk [@nannan2023]; garbage_flag_label's 5 categories match Table 6
# of Nannan et al. (2023) exactly. X59 sits within Category 5 (insufficiently
# specified, alongside Y10-Y34) rather than as its own garbage_flag category.
dt_raw[, `:=`(
  is_garbage      = garbage_flag > 0L,
  is_illdefined   = garbage_flag_label == "Symptoms, signs and ill-defined conditions",
  is_x59          = grepl("^X59", CauseA),
  is_unspecified_facility = as.character(DeathInst) == "Unspecified"
)]

dt_dq <- dt_raw[DeathYear %in% window_years]
```

### Overview of Death Distribution

::: {.panel-tabset}

#### By Province
```{r}
#| label: overview-province-plot

ggplotly(p_overview_province, tooltip = "text") %>%
  layout(hovermode = "closest")
```

#### By Institution
```{r}
#| label: overview-instance-plot

ggplotly(p_overview_instance, tooltip = "text") %>%
  layout(hovermode = "closest")
```

:::

### Garbage Code Trends {#garbage-trends}

Proportion of **non-garbage** codes over time (last `r length(window_years)` years). The **national trend** is shown first; each province is then shown on its **own panel** so that provinces are not compared directly against each other.

```{r}
#| label: garbage-trend-fn
#| include: false

make_nongarbage_plot <- function(province = NULL) {
  d <- if (is.null(province)) dt_dq else dt_dq[DeathProvince == province]
  agg <- d[, .(n = .N, pct_nongarbage = mean(!is_garbage)), by = DeathYear]
  ttl <- if (is.null(province)) "National" else as.character(province)

  p <- ggplot(agg, aes(x = DeathYear, y = pct_nongarbage,
             text = paste0(ttl, "<br>Year: ", DeathYear,
                           "<br>Non-garbage: ", scales::percent(pct_nongarbage, accuracy = 0.1),
                           "<br>N = ", scales::comma(n)))) +
    geom_hline(yintercept = 0.8, linetype = dq_target_lty,
               colour = dq_target_colour, linewidth = 0.7) +
    annotate("point", x = min(agg$DeathYear), y = 0.8, shape = 8,
             colour = dq_target_colour, size = 3) +
    geom_line(colour = "#1976D2", linewidth = 1) +
    geom_point(colour = "#1976D2", size = 2) +
    scale_x_continuous(breaks = window_years) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
    labs(x = NULL, y = "Proportion non-garbage", title = ttl) +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          plot.title  = element_text(face = "bold", size = 11))

  # connectgaps: ggplotly's per-point tooltip conversion inserts NA gaps between
  # points, which otherwise leaves the connecting line invisible.
  ggplotly(p, tooltip = "text") |> style(connectgaps = TRUE) |> layout(hovermode = "closest")
}

provinces_garbage <- sort(unique(as.character(dt_dq$DeathProvince)))
```

```{r}
#| label: garbage-trend-tabs
#| results: asis
#| echo: false

cat("::: {.panel-tabset}\n\n")
cat("#### National\n\n")
print(htmltools::tagList(make_nongarbage_plot(NULL)))
cat("\n\n")
for (prov in provinces_garbage) {
  cat("#### ", prov, "\n\n", sep = "")
  print(htmltools::tagList(make_nongarbage_plot(prov)))
  cat("\n\n")
}
cat(":::\n\n")
```

::: {.callout-tip}
## Interpreting the Plots
The **red dotted line / red star (★) at 80%** is the standardized data-quality target: ≥ 80% non-garbage (equivalently **≤ 20% garbage/ill-defined**) is the WHO/GBD usability cut-point above which cause-of-death data are considered acceptable; **< 10% garbage** marks high quality [@mathers2005; @mahapatra2007; @naghavi2010].
:::

## Data Quality: Ill-Defined vs. Garbage/Unusable Codes {#illdefined-garbage}

::: {.callout-note appearance="simple"}
**Confirmed definitions** — garbage/ill-defined/unusable categories below match the 5-category Global Burden of Disease (Naghavi et al. 2010) unusable-code crosswalk as applied to South African CRVS data by the SAMRC's rapid assessment [@nannan2023]. Full definitions, ICD-10 code lists, and international-standard caveats are in [Methods: Ill-Defined and Garbage/Unusable Cause-of-Death Definitions](#methods-illdefined).

- **Garbage = Unusable** (used interchangeably in this section).
- **Ill-defined (natural)** = ICD-10 **R00–R99** symptoms/signs, confirmed here to match `garbage_flag_label == "Symptoms, signs and ill-defined conditions"` exactly (`r comma(sum(dt_dq$is_illdefined))` deaths, last `r length(window_years)` years). Nannan et al. (2023) report this category nationally at ~13% of deaths (2000-2017), above their cited ideal level of **below 5%**.
- Garbage/unusable is understood as **ill-defined natural (R00–R99) + impossible + intermediate + immediate + insufficiently specified causes** (the last of which includes **X59**, exposure to unspecified factor). Nannan et al. (2023) report ~32% of SA deaths nationally coded to unusable causes overall (2000-2017).
:::

```{r}
#| label: illdefined-garbage-national
#| echo: false

nat_ig <- dt_dq[, .(
  n           = .N,
  pct_garbage = mean(is_garbage),
  pct_illdef  = mean(is_illdefined)
), by = DeathYear]

nat_ig_long <- melt(nat_ig, id.vars = c("DeathYear", "n"),
                     measure.vars = c("pct_garbage", "pct_illdef"),
                     variable.name = "indicator", value.name = "pct")
nat_ig_long[, indicator_label := fifelse(indicator == "pct_garbage",
                                          "Garbage / unusable (%)",
                                          "Ill-defined natural, R00-R99 (%)")]
nat_ig_long[, tt := paste0(indicator_label, "<br>Year: ", DeathYear,
                           "<br>", scales::percent(pct, accuracy = 0.1),
                           "<br>N = ", scales::comma(n))]

p_ig <- ggplot(nat_ig_long, aes(x = DeathYear, y = pct, colour = indicator_label,
                                group = indicator_label, text = tt)) +
  geom_hline(yintercept = dq_target / 100, linetype = dq_target_lty,
             colour = dq_target_colour, linewidth = 0.7) +
  annotate("point", x = min(nat_ig_long$DeathYear), y = dq_target / 100, shape = 8,
           colour = dq_target_colour, size = 3) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = window_years) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_colour_manual(values = c("Garbage / unusable (%)" = "#D32F2F",
                                  "Ill-defined natural, R00-R99 (%)" = "#F77F00")) +
  labs(x = NULL, y = "Percentage of deaths", colour = NULL,
       title = "National: Garbage/Unusable vs. Ill-Defined (R00-R99)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1))

ggplotly(p_ig, tooltip = "text") |>
  style(connectgaps = TRUE) |>
  layout(hovermode = "closest", legend = list(orientation = "h", y = -0.25))
```

::: {.callout-note}
**X59** (exposure to unspecified factor) appears `r comma(sum(dt_dq$is_x59))` times as Cause A over the last `r length(window_years)` years — too small nationally to plot as its own line, but it falls within Category 5 ("Insufficiently specified causes within ICD chapters", alongside Y10-Y34) of the confirmed unusable-code crosswalk [@nannan2023] and is captured within the garbage/unusable total above.
:::

### Unspecified Facility Type {#unspecified-facility}

Shown **on its own** (not combined with other institution types), since the M+E review flagged this category as disproportionately large.

```{r}
#| label: unspecified-facility-trend
#| echo: false

unspec_trend <- dt_dq[, .(n = .N, pct_unspec = mean(is_unspecified_facility)), by = DeathYear]

p_unspec <- ggplot(unspec_trend, aes(x = DeathYear, y = pct_unspec,
           text = paste0("Year: ", DeathYear,
                        "<br>Unspecified facility: ", scales::percent(pct_unspec, accuracy = 0.1),
                        "<br>N = ", scales::comma(n)))) +
  geom_line(colour = "#6A1B9A", linewidth = 1, aes(group = 1)) +
  geom_point(colour = "#6A1B9A", size = 2) +
  scale_x_continuous(breaks = window_years) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = NULL, y = "Proportion of deaths",
       title = "National: Death Institution Recorded as 'Unspecified'") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplotly(p_unspec, tooltip = "text") |> style(connectgaps = TRUE) |> layout(hovermode = "closest")
```

### Data Completeness: Sex and Facility Type {#missingness-table}

Combined completeness table — sex and facility-type proportions with a single **missingness** column, last `r length(window_years)` years.

```{r}
#| label: missingness-table
#| echo: false

sex_tab <- dt_dq[, .(N = .N), by = .(Category = as.character(sex))]
sex_tab[, `:=`(Variable = "Sex", Pct = round(N / sum(N) * 100, 1),
               Missing = Category %in% c("Undeterminable", "Unspecified") | is.na(Category))]

fac_tab <- dt_dq[, .(N = .N), by = .(Category = facility_type)]
fac_tab[, `:=`(Variable = "Facility type", Pct = round(N / sum(N) * 100, 1),
               Missing = Category == "Unspecified/Unknown")]

missing_combined <- rbind(sex_tab, fac_tab)[, .(Variable, Category, `N` = comma(N),
                                                 `% of variable` = Pct,
                                                 `Missing?` = fifelse(Missing, "Yes", ""))]

flextable(as.data.frame(missing_combined)) |>
  set_caption("Sex and facility-type completeness (proportions with missingness flag)") |>
  bold(part = "header") |>
  merge_v(j = "Variable") |>
  autofit()
```

## Non-Hospital Deaths {#non-hospital}

Deaths occurring outside hospitals (at home, nursing homes, etc.) represent an important indicator for health system access and end-of-life care patterns.

::: {.callout-note appearance="simple"}
## How facility type is classified
Death institution (`DeathInst`, 8 raw levels) is grouped into 3 explicit categories for this section — none are silently dropped:

| `DeathInst` value | Grouped as | Rationale |
|---|---|---|
| Hospital, Emergency room/Outpatient | **In-hospital** | Acute facility-based care |
| Home, Nursing home, Dead on arrival, Other | **Out-of-hospital** | Death occurred outside acute hospital care (see note on Nursing home / Dead on arrival below) |
| Unknown, Unspecified | **Unspecified/Unknown** | Reported separately, **excluded** from the non-hospital % trend denominator below |

- **Nursing home** is grouped as **Out-of-hospital**, consistent with this section's framing ("at home, nursing homes, etc.").
- **Dead on arrival** is grouped as **Out-of-hospital** on the basis that the death itself occurred before/outside hospital care, even though it was certified at a facility — this is a judgment call, not an official CRVS convention, and should be confirmed with the M+E team.
- **Unspecified/Unknown** deaths are tracked in their own right in [Unspecified Facility Type](#unspecified-facility) and the [completeness table](#missingness-table) above, and are excluded (not imputed) from the non-hospital proportion charts so the trend reflects only records with a known facility type.
:::

```{r}
#| label: nonhospital-province-plot

ggplotly(p_nonhospital_province, tooltip = "text") %>%
  layout(hovermode = "closest")
```

As with the garbage-code trends above, the **national trend** lands first, followed by a **tab per province** so provinces are not compared directly against each other.

```{r}
#| label: nonhospital-trend-fn
#| include: false

make_nonhospital_plot <- function(province = NULL) {
  # Excludes "Unspecified/Unknown" facility type from the denominator (see the
  # definitions callout above); reported separately, not silently dropped.
  d <- if (is.null(province)) dt_dq[facility_type != "Unspecified/Unknown"] else dt_dq[DeathProvince == province & facility_type != "Unspecified/Unknown"]
  agg <- d[, .(n = .N, pct_nonhosp = mean(facility_type == "Out-of-hospital")), by = DeathYear]
  ttl <- if (is.null(province)) "National" else as.character(province)

  p <- ggplot(agg, aes(x = DeathYear, y = pct_nonhosp,
             text = paste0(ttl, "<br>Year: ", DeathYear,
                           "<br>Non-hospital: ", scales::percent(pct_nonhosp, accuracy = 0.1),
                           "<br>N = ", scales::comma(n)))) +
    geom_line(colour = "#E63946", linewidth = 1) +
    geom_point(colour = "#E63946", size = 2) +
    scale_x_continuous(breaks = window_years) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
    labs(x = NULL, y = "Proportion non-hospital", title = ttl) +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          plot.title  = element_text(face = "bold", size = 11))

  ggplotly(p, tooltip = "text") |> style(connectgaps = TRUE) |> layout(hovermode = "closest")
}

provinces_nonhosp <- sort(unique(as.character(dt_dq$DeathProvince)))
```

```{r}
#| label: nonhospital-trend-tabs
#| results: asis
#| echo: false

cat("::: {.panel-tabset}\n\n")
cat("#### National\n\n")
print(htmltools::tagList(make_nonhospital_plot(NULL)))
cat("\n\n")
for (prov in provinces_nonhosp) {
  cat("#### ", prov, "\n\n", sep = "")
  print(htmltools::tagList(make_nonhospital_plot(prov)))
  cat("\n\n")
}
cat(":::\n\n")
```

::: {.callout-tip}
## Interpreting Non-Hospital Deaths
Hover over any point to see:
- Non-hospital deaths (N) and percentage
- Total deaths for comparison
- Percentage contribution to national non-hospital deaths

The **grey dashed line at 50%** is a **descriptive reference only** — it is *not* a literature-based quality target. It is styled differently (grey, no star) from the standardized red cause-of-death target to avoid implying a benchmark that the published CRVS literature does not define.
:::

---

## HIV/AIDS Deaths {#hiv}

### HIV/AIDS Mortality: VR vs Models

```{r}
#| label: hiv-plot

ggplotly(p_hiv + coord_cartesian(xlim = range(window_years)), tooltip = "text") %>%
  style(connectgaps = TRUE) %>%
  layout(hovermode = "closest")
```

::: {.callout-note}
## Key Findings
- VR data substantially underestimates HIV/AIDS mortality compared to the Tembisa model
- The NCODV study suggests ~23% of deaths may be HIV-related (2017-2018)
- The gap has narrowed over time but remains significant
:::

::: {.callout-important icon=false}
## So what? — Decision this data informs

**Monitored action: correct certification and coding of HIV/AIDS as the underlying cause of death.**

The gap between VR and the Tembisa model is driven mainly by HIV deaths being mis-certified to indicator or "garbage" causes (e.g. TB, pneumonia, anaemia, wasting) rather than to HIV/AIDS — a recognised driver of cause-of-death undercount in high-prevalence settings [@mathers2005; @naghavi2010].

- **Trigger:** a persistent VR-vs-model gap, and an NCODV reattribution share (~23% of deaths HIV-related, 2017–2018) far exceeding the HIV fraction recorded in VR.
- **Decision:** prioritise MCCD certifier training on naming HIV/AIDS as the underlying cause, and apply NCODV-based reattribution before reporting HIV mortality from VR.
- **Expected effect:** the VR series moves **toward** the Tembisa model line as certification improves.
- **Monitor:** re-render this dashboard each cycle (see the *Report generated* timestamp) and track narrowing of the VR-vs-model gap.
:::

## TB Deaths {#tb}

### TB Mortality: VR vs Models

```{r}
#| label: tb-plot

ggplotly(p_tb + coord_cartesian(xlim = range(window_years)), tooltip = "text") %>%
  style(connectgaps = TRUE) %>%
  layout(hovermode = "closest")
```

::: {.callout-important icon=false}
## So what? — Decision this data informs

**Monitored action: correct certification and coding of TB as the underlying cause of death.**

TB mortality in VR diverges from the Tembisa model and NMC notifications largely because TB deaths are mis-certified to terminal or ill-defined causes, or lost where HIV/TB co-morbidity is not correctly sequenced on the MCCD [@mathers2005; @naghavi2010].

- **Trigger:** a VR count that sits below the Tembisa model and/or NMC notification trend for the same period.
- **Decision:** prioritise MCCD certifier training on correct underlying-cause selection for TB (including HIV/TB sequencing), and reconcile VR against NMC notifications.
- **Expected effect:** the VR series converges **toward** the model and notification trends as certification and coding improve.
- **Monitor:** re-render this dashboard each cycle (see the *Report generated* timestamp) and track convergence of the VR, model, and notification series.
:::

## Poisoning Deaths {#poisoning}

### Poisoning Mortality: VR vs NMC

```{r}
#| label: poisoning-plot

ggplotly(p_poisoning + coord_cartesian(xlim = range(window_years)), tooltip = "text") %>%
  style(connectgaps = TRUE) %>%
  layout(hovermode = "closest")
```

::: {.callout-note}
## Key Findings
- VR data is limited to pesticide-poisoning ICD-10 codes: **X48** (accidental), **X68** (intentional self-poisoning), **X87** (assault), and **Y18** (undetermined intent) — matching the scope of the NMC agricultural/pesticide comparison
- NMC tracks notifications of agricultural and stock remedy poisoning cases
- The comparison shows notification patterns vs recorded deaths from poisoning
:::

### Total Unnatural Deaths

```{r}
#| label: unnatural-plot
#| eval: false

ggplotly(p_unnatural, tooltip = "text") %>%
  style(connectgaps = TRUE) %>%
  layout(hovermode = "closest")
```

::: {.callout-tip}
## Understanding Unnatural Deaths
This includes all external causes of death (ICD-10 chapters V, W, X, Y):
- Transport accidents (V codes)
- Other external causes like falls, drowning (W codes)  
- Poisoning and exposure (X codes)
- Assault and intentional harm (Y codes)

Monitoring unnatural deaths helps assess injury prevention efforts and overall safety trends.
:::

## Excess Deaths Summary {#excess-deaths}

This section provides a comprehensive overview of excess mortality by cause of death from 2019-2022, showing which conditions contributed most to excess deaths during the COVID-19 pandemic period.

::: {.panel-tabset}

#### Combined Natural & External Causes

```{r}
#| label: excess-deaths-combined-table
#| fig-cap: "Combined excess deaths by Leading Group of Causes (LGH) and external causes, 2019-2022"

if (exists("dt_analysis") && exists("tabulate_excess_grand_total_with_external")) {
  # Generate the enhanced table with both natural and external causes
  if (exists("dt_external")) {
    excess_table <- tabulate_excess_grand_total_with_external(dt_analysis, dt_external, years = 2019:2022, include_external = TRUE)
  } else {
    # Fallback to natural causes only
    excess_table <- tabulate_excess_grand_total(dt_analysis, years = 2019:2022)
  }
  excess_table
} else {
  cat("Data or functions not available for excess deaths analysis")
}
```

::: {.callout-note}
## Understanding the Combined Table
- **Natural causes** (LGH codes): Positive values indicate excess deaths, negative values (green) indicate deficit deaths
- **External causes** (Unnatural): Shows actual death counts (no baseline model available)
- **Total column** shows cumulative excess/deficit across all years
- **Color coding**: Green = deficit, Red = high excess (≥10,000/year or ≥50,000 total)
- Data includes major external cause categories: road traffic accidents, falls, assault (gunshots), suicide, etc.
:::

#### Natural Causes Only

```{r}
#| label: excess-deaths-natural-only
#| fig-cap: "Excess deaths by Leading Group of Causes (LGH) - natural causes only, 2019-2022"

if (exists("dt_analysis") && exists("tabulate_excess_grand_total_with_external")) {
  # Generate table with natural causes only
  excess_table_natural <- tabulate_excess_grand_total_with_external(dt_analysis, dt_external, years = 2019:2022, include_external = FALSE)
  excess_table_natural
} else {
  cat("Data or functions not available for natural causes analysis")
}
```

::: {.callout-note}
## Understanding Natural Causes Table
- Shows **excess/deficit deaths** compared to baseline models for disease-related mortality
- **Positive values** indicate more deaths than expected (excess mortality)
- **Negative values** (green) indicate fewer deaths than expected (deficit mortality)
- **LGH classification** groups causes into epidemiologically meaningful categories
:::

#### External Causes Only  

```{r}
#| label: excess-deaths-external-only
#| fig-cap: "Unnatural deaths by external cause category, 2019-2022"
#| eval: false

if (exists("dt_external") && exists("tabulate_external_causes_only")) {
  # Generate table with external causes only
  external_table <- tabulate_external_causes_only(dt_external, years = 2019:2022)
  external_table
} else {
  cat("External causes data or functions not available")
}
```

::: {.callout-note}
## Understanding External Causes Table  
- Shows **actual death counts** for injury-related mortality (no baseline models available)
- **Major categories**: Road traffic accidents (V01-V99), falls (W00-W19), assault/gunshots (X85-Y09), suicide (X60-X84), accidental poisoning (X40-X49), etc.
- **High counts** (≥1,000/year or ≥5,000 total) are highlighted in red
- Useful for monitoring injury prevention and safety trends
:::

:::

## Data Sources

| Source | Description | Coverage |
|--------|-------------|----------|
| Stats SA P0309.3 | Mortality and causes of death | 2006-2022 |
| Thembisa Model v4.8 | HIV/AIDS demographic projections | 1997-2025 |
| NCODV Study | Verbal autopsy validation | 2017-2018 |
| NMC Database | TB and poisoning surveillance notifications | 2017-2024 |

## Methods

### Garbage Code Classification

Deaths are classified using the WHO garbage code framework. A death is considered "not garbage" if the underlying cause code provides meaningful epidemiological information.

### Ill-Defined and Garbage/Unusable Cause-of-Death Definitions {#methods-illdefined}

::: {.callout-note appearance="simple"}
**Status: confirmed.** The definitions below match the 5-category unusable-code crosswalk (based on Naghavi et al. 2010's GBD classification) as applied to South African CRVS data by the SAMRC's rapid assessment of cause-of-death data [@nannan2023]. The `garbage_flag_label` categories already present in this dataset's source pipeline correspond exactly to Nannan et al. (2023)'s five categories.
:::

- **Ill-defined (natural) causes**: ICD-10 Chapter XVIII, **R00–R99** ("Symptoms, signs and abnormal clinical/laboratory findings, not elsewhere classified"). In this dataset this corresponds exactly to `garbage_flag_label == "Symptoms, signs and ill-defined conditions"` (verified 1:1 against a direct `R00-R99` regex on `UnderlyingCause`). Nannan et al. (2023) report this nationally at ~13% of deaths (2000-2017), above their cited ideal level of **below 5%**.
- **Garbage / unusable codes** (used interchangeably here): the broader GBD "unusable code" set — causes too vague, intermediate, or implausible to be the true underlying cause of death. Operationally this dashboard treats **any `garbage_flag > 0`** category as garbage/unusable, composed of the five categories confirmed in Nannan et al. (2023), Table 6:
  1. Ill-defined natural causes (R00–R99, above);
  2. Impossible causes of death (e.g. essential primary hypertension, atherosclerosis, long-term sequelae — implausible as an underlying cause);
  3. Intermediate causes of death (e.g. heart failure, septicaemia, peritonitis, osteomyelitis, pulmonary embolism — a consequence of another condition, not the root cause);
  4. Immediate causes of death (e.g. cardiac arrest, respiratory failure — a terminal mode of dying, not a specific disease);
  5. Insufficiently specified causes within ICD chapters (not-otherwise-specified codes, e.g. unspecified cancer site).
  
  Nannan et al. (2023) report ~32% of SA deaths nationally coded to unusable causes overall (2000-2017), with ill-defined natural causes the largest single contributor (12-14%).
- **X59** ("Exposure to unspecified factor causing other and unspecified injury") falls within **Category 5 (Insufficiently specified)** per the GBD Appendix 5 code list (alongside Y10-Y34), not as its own standalone `garbage_flag` category in the source pipeline — it is monitored separately here (see [Data Quality: Ill-Defined vs. Garbage/Unusable Codes](#illdefined-garbage)) because of its specific relevance to injury/undetermined-intent deaths.

**Caveats vs. international (WHO/GBD) standards:**

- The WHO/GBD garbage-code framework additionally reclassifies some causes *by age and sex* (e.g. certain cardiovascular or cancer codes are only "garbage" in specific age groups) [@naghavi2010]. The current `garbage_flag` in this dataset is **not age/sex-stratified**, so the % garbage reported here is not guaranteed to be directly comparable to published GBD "garbage-code" percentages for other countries — treat cross-study comparisons as indicative, not exact.
- The 80%/20% usability thresholds displayed on these charts are the **Mahapatra (2007) CRVS-series convention** [@mahapatra2007], consistent with the wider WHO/GBD literature [@mathers2005; @naghavi2010], and are a **general usability benchmark**, not a South Africa-specific target — Nannan et al. (2023) instead cite a stricter **below 5%** ideal specifically for ill-defined natural causes (R00-R99) in the SA context, separate from the overall garbage/unusable total.
- "Not-specified" facility/institution (`DeathInst == "Unspecified"`) is a **data-collection quality issue**, distinct from ill-defined *cause* coding — the two are reported separately in this dashboard and should not be conflated when interpreting "garbage" percentages.

### Data Sources

- **VR Data**: Individual death records from `LGH_MasterFile_preCollapsedAll.feather`
- **Tembisa**: Age-specific mortality outputs from version 4.8
- **NMC**: Notifiable conditions from the surveillance database

---

## References {.unnumbered}

::: {#refs}
:::
