---
title: "Monitoring and Evaluation Dashboard"
description: "Comparing vital registration data with Tembisa model and NCODV-corrected estimates for HIV/AIDS, TB, and poisoning mortality, plus data quality indicators"
date: "2026-01-20"
categories: [mortality, HIV, TB, poisoning, data-quality, timeliness, non-hospital]
bibliography: references.bib
csl: ../../vancouver.csl
---

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(plotly)
library(htmlwidgets)
library(arrow)
library(data.table)
library(flextable)
library(conflicted)
library(DT)
library(readxl)
library(haven)
library(scales)

# Resolve common conflicts between lubridate and data.table
conflicts_prefer(
  data.table::hour,
  data.table::minute,
  data.table::second,
  data.table::year,
  data.table::month,
  data.table::mday,
  data.table::wday
)

# Resolve conflicts between plotly and graphics
conflicts_prefer(
  plotly::layout
)

# -------------------------------------------------------------------------
# Standardized data-quality TARGET (single source of truth for all figures)
# Literature basis: WHO/GBD cause-of-death usability convention.
#   - >20% of deaths assigned to ill-defined / garbage codes indicates
#     LOW-quality, low-usability cause-of-death data.
#   - <10% is the "high quality" benchmark.
# Primary sources: Mathers et al. (2005), Bull WHO 83(3):171-7 (ill-defined
#   % classification); Mahapatra et al. (2007), Lancet 370:1653-63 (CRVS
#   series); Naghavi et al. (2010), Popul Health Metr 8:9 (garbage codes).
# Target is rendered consistently as a RED DOTTED line + RED STAR (shape 8).
# -------------------------------------------------------------------------
dq_target          <- 20          # percent (usability upper bound)
dq_target_high     <- 10          # percent (high-quality benchmark, in text)
dq_target_colour   <- "#D32F2F"   # standardized red
dq_target_lty      <- "dotted"

# Auditability: capture render timestamp once
render_stamp <- format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")

# Source the wrangling script if outputs are missing or stale
output_file <- "../../outputs/me_plots.RData"
wrangling_script <- "../../ME_wrangling.r"

# Check if we need to re-run wrangling
if (!file.exists(output_file)) {
  message("Running wrangling script...")
  source(wrangling_script, local = new.env())
}

# Load the pre-built plots
load(output_file)

# Source dashboard helpers for excess deaths functions
helpersPath <- "../../dashboard/dashboard_helpers.r"
if (file.exists(helpersPath)) {
  source(helpersPath)
}

# Load analysis data for excess deaths calculations
analysisPath <- "../../LGH_AnalysisData_2022.feather"
if (file.exists(analysisPath)) {
  dt_analysis <- read_feather(analysisPath) %>% as.data.table()
}

# Load external causes data (unnatural deaths)
externalPath <- "../../Deaths2022_MRCversionFINAL.feather"  
if (file.exists(externalPath)) {
  dt_external <- read_feather(externalPath) %>% as.data.table()
  # Filter to external causes only (V, W, X, Y codes)
  dt_external <- dt_external[grepl("^[VWXY]", UnderlyingCause)]
} else {
  dt_external <- NULL
}
```

## Overview

::: {.callout-note appearance="minimal"}
**Report generated:** `r render_stamp` &nbsp;·&nbsp; *Timestamp shown for auditability — re-rendering this dashboard refreshes it against the latest source data.*
:::

This dashboard compares vital registration (VR) data with alternative mortality data sources for monitoring and evaluation purposes:

- **Tembisa Model**: Demographic projections for HIV/AIDS and TB mortality
- **NCODV Study**: Verbal autopsy-corrected estimates (2017-2018)
- **NMC Database**: Notifiable Medical Conditions surveillance data for TB and poisoning

## HIV/AIDS Deaths {#hiv}

### HIV/AIDS Mortality: VR vs Models

```{r}
#| label: hiv-plot

ggplotly(p_hiv, tooltip = "text") %>%
  layout(hovermode = "closest")
```

::: {.callout-note}
## Key Findings
- VR data substantially underestimates HIV/AIDS mortality compared to the Tembisa model
- The NCODV study suggests ~23% of deaths may be HIV-related (2017-2018)
- The gap has narrowed over time but remains significant
:::

::: {.callout-important icon=false}
## So what? — Decision this data informs

**Monitored action: correct certification and coding of HIV/AIDS as the underlying cause of death.**

The gap between VR and the Tembisa model is driven mainly by HIV deaths being mis-certified to indicator or "garbage" causes (e.g. TB, pneumonia, anaemia, wasting) rather than to HIV/AIDS — a recognised driver of cause-of-death undercount in high-prevalence settings [@mathers2005; @naghavi2010].

- **Trigger:** a persistent VR-vs-model gap, and an NCODV reattribution share (~23% of deaths HIV-related, 2017–2018) far exceeding the HIV fraction recorded in VR.
- **Decision:** prioritise MCCD certifier training on naming HIV/AIDS as the underlying cause, and apply NCODV-based reattribution before reporting HIV mortality from VR.
- **Expected effect:** the VR series moves **toward** the Tembisa model line as certification improves.
- **Monitor:** re-render this dashboard each cycle (see the *Report generated* timestamp) and track narrowing of the VR-vs-model gap.
:::

## TB Deaths {#tb}

### TB Mortality: VR vs Models

```{r}
#| label: tb-plot

ggplotly(p_tb, tooltip = "text") %>%
  layout(hovermode = "closest")
```

::: {.callout-important icon=false}
## So what? — Decision this data informs

**Monitored action: correct certification and coding of TB as the underlying cause of death.**

TB mortality in VR diverges from the Tembisa model and NMC notifications largely because TB deaths are mis-certified to terminal or ill-defined causes, or lost where HIV/TB co-morbidity is not correctly sequenced on the MCCD [@mathers2005; @naghavi2010].

- **Trigger:** a VR count that sits below the Tembisa model and/or NMC notification trend for the same period.
- **Decision:** prioritise MCCD certifier training on correct underlying-cause selection for TB (including HIV/TB sequencing), and reconcile VR against NMC notifications.
- **Expected effect:** the VR series converges **toward** the model and notification trends as certification and coding improve.
- **Monitor:** re-render this dashboard each cycle (see the *Report generated* timestamp) and track convergence of the VR, model, and notification series.
:::

## Poisoning Deaths {#poisoning}

### Poisoning Mortality: VR vs NMC

```{r}
#| label: poisoning-plot

ggplotly(p_poisoning, tooltip = "text") %>%
  layout(hovermode = "closest")
```

::: {.callout-note}
## Key Findings
- VR data captures poisoning deaths using ICD-10 codes X40-X49 (accidental poisoning) and X60-X69 (intentional self-poisoning)
- NMC tracks notifications of agricultural and stock remedy poisoning cases
- The comparison shows notification patterns vs recorded deaths from poisoning
:::

### Total Unnatural Deaths

```{r}
#| label: unnatural-plot
#| eval: false

ggplotly(p_unnatural, tooltip = "text") %>%
  layout(hovermode = "closest")
```

::: {.callout-tip}
## Understanding Unnatural Deaths
This includes all external causes of death (ICD-10 chapters V, W, X, Y):
- Transport accidents (V codes)
- Other external causes like falls, drowning (W codes)  
- Poisoning and exposure (X codes)
- Assault and intentional harm (Y codes)

Monitoring unnatural deaths helps assess injury prevention efforts and overall safety trends.
:::

## Excess Deaths Summary {#excess-deaths}

This section provides a comprehensive overview of excess mortality by cause of death from 2019-2022, showing which conditions contributed most to excess deaths during the COVID-19 pandemic period.

::: {.panel-tabset}

#### Combined Natural & External Causes

```{r}
#| label: excess-deaths-combined-table
#| fig-cap: "Combined excess deaths by Leading Group of Causes (LGH) and external causes, 2019-2022"

if (exists("dt_analysis") && exists("tabulate_excess_grand_total_with_external")) {
  # Generate the enhanced table with both natural and external causes
  if (exists("dt_external")) {
    excess_table <- tabulate_excess_grand_total_with_external(dt_analysis, dt_external, years = 2019:2022, include_external = TRUE)
  } else {
    # Fallback to natural causes only
    excess_table <- tabulate_excess_grand_total(dt_analysis, years = 2019:2022)
  }
  excess_table
} else {
  cat("Data or functions not available for excess deaths analysis")
}
```

::: {.callout-note}
## Understanding the Combined Table
- **Natural causes** (LGH codes): Positive values indicate excess deaths, negative values (green) indicate deficit deaths
- **External causes** (Unnatural): Shows actual death counts (no baseline model available)
- **Total column** shows cumulative excess/deficit across all years
- **Color coding**: Green = deficit, Red = high excess (≥10,000/year or ≥50,000 total)
- Data includes major external cause categories: road traffic accidents, falls, assault (gunshots), suicide, etc.
:::

#### Natural Causes Only

```{r}
#| label: excess-deaths-natural-only
#| fig-cap: "Excess deaths by Leading Group of Causes (LGH) - natural causes only, 2019-2022"

if (exists("dt_analysis") && exists("tabulate_excess_grand_total_with_external")) {
  # Generate table with natural causes only
  excess_table_natural <- tabulate_excess_grand_total_with_external(dt_analysis, dt_external, years = 2019:2022, include_external = FALSE)
  excess_table_natural
} else {
  cat("Data or functions not available for natural causes analysis")
}
```

::: {.callout-note}
## Understanding Natural Causes Table
- Shows **excess/deficit deaths** compared to baseline models for disease-related mortality
- **Positive values** indicate more deaths than expected (excess mortality)
- **Negative values** (green) indicate fewer deaths than expected (deficit mortality)
- **LGH classification** groups causes into epidemiologically meaningful categories
:::

#### External Causes Only  

```{r}
#| label: excess-deaths-external-only
#| fig-cap: "Unnatural deaths by external cause category, 2019-2022"
#| eval: false

if (exists("dt_external") && exists("tabulate_external_causes_only")) {
  # Generate table with external causes only
  external_table <- tabulate_external_causes_only(dt_external, years = 2019:2022)
  external_table
} else {
  cat("External causes data or functions not available")
}
```

::: {.callout-note}
## Understanding External Causes Table  
- Shows **actual death counts** for injury-related mortality (no baseline models available)
- **Major categories**: Road traffic accidents (V01-V99), falls (W00-W19), assault/gunshots (X85-Y09), suicide (X60-X84), accidental poisoning (X40-X49), etc.
- **High counts** (≥1,000/year or ≥5,000 total) are highlighted in red
- Useful for monitoring injury prevention and safety trends
:::

:::

## Data Quality: Garbage Codes {#garbage-codes}

Garbage codes are ICD-10 codes that provide insufficient information about the underlying cause of death. Monitoring the proportion of non-garbage codes is essential for assessing data quality.

### Overview of Death Distribution

::: {.panel-tabset}

#### By Province
```{r}
#| label: overview-province-plot

ggplotly(p_overview_province, tooltip = "text") %>%
  layout(hovermode = "closest")
```

#### By Institution
```{r}
#| label: overview-instance-plot

ggplotly(p_overview_instance, tooltip = "text") %>%
  layout(hovermode = "closest")
```

:::

### Garbage Code Trends

::: {.panel-tabset}

#### By Province
```{r}
#| label: garbage-province-plot

ggplotly(p_garbage_province, tooltip = "text") %>%
  layout(hovermode = "closest")
```

#### By Institution Type
```{r}
#| label: garbage-instance-plot

ggplotly(p_garbage_instance, tooltip = "text") %>%
  layout(hovermode = "closest")

  # Look at the proportion of locations by province s
```

:::

::: {.callout-tip}
## Interpreting the Plots
Hover over any point to see:
- The proportion of non-garbage codes
- Total deaths (N) in that province/institution
- Percentage contribution to national deaths

The **red dotted line / red star (★) at 80%** is the standardized data-quality target: ≥ 80% non-garbage (equivalently **≤ 20% garbage/ill-defined**) is the WHO/GBD usability cut-point above which cause-of-death data are considered acceptable; **< 10% garbage** marks high quality [@mathers2005; @mahapatra2007; @naghavi2010].
:::

## Non-Hospital Deaths {#non-hospital}

Deaths occurring outside hospitals (at home, nursing homes, etc.) represent an important indicator for health system access and end-of-life care patterns.

::: {.panel-tabset}

#### By Province (Total)
```{r}
#| label: nonhospital-province-plot

ggplotly(p_nonhospital_province, tooltip = "text") %>%
  layout(hovermode = "closest")
```

#### Trend Over Time
```{r}
#| label: nonhospital-trend-plot

ggplotly(p_nonhospital_trend, tooltip = "text") %>%
  layout(hovermode = "closest")
```

:::

::: {.callout-tip}
## Interpreting Non-Hospital Deaths
Hover over any point to see:
- Non-hospital deaths (N) and percentage
- Total deaths for comparison
- Percentage contribution to national non-hospital deaths

The **grey dashed line at 50%** is a **descriptive reference only** — it is *not* a literature-based quality target. It is styled differently (grey, no star) from the standardized red cause-of-death target to avoid implying a benchmark that the published CRVS literature does not define.
:::

## Report Timeliness {#timeliness}

Stats SA releases mortality reports with a delay of several years. This section tracks the timeliness of releases.

**Click on any point to open the corresponding Stats SA PDF report.**

```{r}
#| label: timeliness-plot

p_timeliness_interactive
```

### Registration Delay: Timeliness vs. Completeness Tradeoff {#reg-tradeoff}

Each death record carries a **date of death** and a **date of registration**. The gap between them — the *registration lag* — drives a fundamental tradeoff: publishing statistics soon after the death year is **timely** but misses deaths that are registered late, while waiting longer captures those late registrations and is more **complete**. Using registration timing as a proxy, the curve below shows, for each death-year cohort, the cumulative share of its registered deaths captured by *N* years after death.

```{r}
#| label: reg-tradeoff-plot
#| echo: false
#| warning: false
#| message: false

reg_cols <- c("DeathYear", "DeathMonth", "DeathDay",
              "RegistrationYear", "RegistrationMonth", "RegistrationDay")

reg_raw <- read_feather("../../LGH_MasterFile_preCollapsed2022.feather",
                        col_select = reg_cols) |> as.data.table()
reg_raw[, (reg_cols) := lapply(.SD, function(x) as.integer(zap_labels(x))),
        .SDcols = reg_cols]

reg_raw[, raw_lag := RegistrationYear - DeathYear]
reg_raw[, lag_yr := pmax(raw_lag, 0L)]

# --- Data-quality flags to surface as limitations / checks ---
flag_partial2013 <- reg_raw[DeathYear == 2013, .N]      # partial cohort in extract
flag_neg_lag     <- reg_raw[raw_lag < 0L, .N]           # registered before death (impossible)
flag_floored     <- reg_raw[raw_lag < 0L, .N]           # records floored to lag 0
flag_late5       <- reg_raw[raw_lag > 5L, .N]           # very late (>5 yr) registrations

# Full-year cohorts only (2013 is a partial cohort in this extract)
reg_dt   <- reg_raw[DeathYear %in% 2014:2022]
max_reg  <- 2024L   # latest registration year present in the extract
max_lag  <- 5L

cohort_n <- reg_dt[, .(N = .N), by = DeathYear]
cum      <- reg_dt[, .(n = .N), by = .(DeathYear, lag = pmin(lag_yr, max_lag))]
grid     <- CJ(DeathYear = 2014:2022, lag = 0:max_lag)
cum      <- cum[grid, on = .(DeathYear, lag)][is.na(n), n := 0L]
setorder(cum, DeathYear, lag)
cum[cohort_n, N := i.N, on = "DeathYear"]
cum[, cum_pct := cumsum(n) / N * 100, by = DeathYear]

# Draw each cohort only up to its observed maximum lag (right-truncation honesty)
cum[, obs_max_lag := max_reg - DeathYear]
cum_plot <- cum[lag <= obs_max_lag]
cum_plot[, tt := paste0("Death cohort: ", DeathYear,
                        "<br>Within ", lag, " yr of death: ",
                        sprintf("%.2f%%", cum_pct))]

p_tradeoff <- ggplot(cum_plot,
                     aes(lag, cum_pct, colour = DeathYear,
                         group = DeathYear, text = tt)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.6) +
  scale_colour_viridis_c(name = "Death\nyear", breaks = seq(2014, 2022, 2)) +
  scale_x_continuous(breaks = 0:max_lag) +
  coord_cartesian(ylim = c(floor(min(cum_plot$cum_pct)), 100)) +
  labs(
    title    = "Registration timeliness vs. completeness, by death-year cohort",
    subtitle = "Cumulative % of each cohort registered by N years after death\nDashed line = 1-year publication cut-off; rise to its right = completeness gained by waiting",
    x        = "Years after death (registration lag)",
    y        = "Cumulative % of cohort registered"
  ) +
  theme_minimal(base_size = 12)

ggplotly(p_tradeoff, tooltip = "text") |>
  layout(hovermode = "closest")
```

::: {.callout-note}
## Interpreting the Tradeoff
- The **height at lag 0** is the *timely* share — deaths registered in the same calendar year (≈ 98.5–99.2% across cohorts), so South African VR registration is highly timely.
- The **rise between lag 0 and lag 1** is the *completeness gain* from waiting one extra year; it adds only ≈ 0.6–1.3 percentage points, after which the curves plateau.
- **Tradeoff implication:** publishing one year after the death year (dashed line) captures ≈ 99.8% of all eventually-registered deaths, so the completeness penalty for timely release is small.
- Recent cohorts (2021–2022) are **right-truncated** — they can only be observed to registration year `r max_reg` — so their curves stop at a shorter lag.
:::

::: {.callout-warning icon=false}
## ⚠ Data limitations & checks to investigate
These handling decisions are applied above so the plot is honest, but each one flags a record-level issue that should be **checked at source** rather than silently corrected:

- **2013 is a partial cohort** (`r format(flag_partial2013, big.mark=",")` records) and is **excluded** from cohort comparisons — its same-year registration share (≈ 54%) reflects a truncated extract window, not poor timeliness. *Check:* confirm the extract's start date and whether earlier 2013 deaths live in a separate file.
- **Negative registration lags** (`r format(flag_neg_lag, big.mark=",")` record(s)) imply a death registered *before* it occurred — logically impossible. These are **floored to lag 0** for plotting. *Check:* audit these records for transposed or mis-keyed death/registration dates.
- **Very late registrations** (>5 years: `r format(flag_late5, big.mark=",")` records) are **capped at lag 5** in the cumulative curve. *Check:* verify whether these are genuine late registrations (e.g. delayed/contested deaths) or data-entry errors in the year fields.

None of these materially change the headline (≈ 99.8% within one year), but they should be triaged as part of routine VR data-quality control.
:::

## Data Sources

| Source | Description | Coverage |
|--------|-------------|----------|
| Stats SA P0309.3 | Mortality and causes of death | 2006-2022 |
| Thembisa Model v4.8 | HIV/AIDS demographic projections | 1997-2025 |
| NCODV Study | Verbal autopsy validation | 2017-2018 |
| NMC Database | TB and poisoning surveillance notifications | 2017-2024 |

## Methods

### Garbage Code Classification

Deaths are classified using the WHO garbage code framework. A death is considered "not garbage" if the underlying cause code provides meaningful epidemiological information.

### Data Sources

- **VR Data**: Individual death records from `LGH_MasterFile_preCollapsedAll.feather`
- **Tembisa**: Age-specific mortality outputs from version 4.8
- **NMC**: Notifiable conditions from the surveillance database

---

## South Africa CRVS Monitoring Indicators (Cause-of-Death Focus) {#crvs-compendium}

The table below lists the **South Africa CRVS monitoring indicators with a cause-of-death focus** (v1, June 2025). This is the SA-contextualised indicator set spanning the CRVS pathway — from notification and registration through medical certification, cause-of-death coding, quality control, and vital-statistics generation. Use the column filters and search box to explore indicators by CRVS milestone or performance domain.

```{r}
#| label: crvs-compendium-table
#| echo: false

compendium_raw <- read_excel(
  "20260506_SA_CRVS_monitoring_indicators_COD_focus_V1.xlsx",
  sheet = "Indicators"
)
compendium_raw <- compendium_raw[!is.na(compendium_raw$Indicator), ]

datatable(
  compendium_raw[, c("Indicator", "CRVS milestone", "Performance domain",
                     "Desagregation level", "Numerator", "Denominator",
                     "Compendium APP")],
  colnames = c("Disaggregation level" = "Desagregation level",
               "In compendium APP"    = "Compendium APP"),
  filter     = "top",
  extensions = "Buttons",
  rownames   = FALSE,
  options    = list(
    dom        = "Bfrtip",
    buttons    = c("csv", "excel"),
    pageLength = 15,
    scrollX    = TRUE,
    autoWidth  = FALSE,
    columnDefs = list(
      list(width = "260px", targets = 0),   # Indicator
      list(width = "180px", targets = 4),   # Numerator
      list(width = "180px", targets = 5)    # Denominator
    )
  )
)
```

---

## Data Quality: Facility-Type Decision Framework {#facility-dq}

Using a **decision-first** approach, this section identifies which `facility_type` presents the greatest data quality burden. Three indicators aligned to the SA CRVS monitoring set are computed from individual death records (2015–2022):

| Indicator | SA monitoring indicator | Definition |
|---|---|---|
| Any garbage CoD | *Percent of medically certified deaths coded with an unusable cause* | Underlying cause coded as any garbage category (flag > 0) |
| Ill-defined CoD | *Percent of medically certified deaths with ill-defined cause of death* | Underlying cause is a symptom / ill-defined (R-codes; flag = 1) |
| Non-medical ascertainment | *Percent of deaths occurring outside health facilities with no medically certified cause of death* | Death certified by family interview, "Other", or unspecified (not by a medical practitioner or autopsy) |

**Facility types** are grouped as: *In-hospital* (Hospital, Emergency room/Outpatient, Dead on arrival, Nursing home) and *Out-of-hospital* (Home, Other). Records with Unspecified/Unknown institution are excluded from facility comparisons.

```{r}
#| label: dq-data-load
#| include: false

dt_raw <- read_feather("../../LGH_MasterFile_preCollapsed2022.feather") |>
  as.data.table()

dt_raw[, `:=`(
  DeathInst     = as_factor(DeathInst),
  DeathProvince = as_factor(DeathProvince),
  Ascertainment = as_factor(Ascertainment)
)]

dt_raw[, facility_type := fcase(
  DeathInst %in% c("Hospital", "Emergency room/Outpatient",
                   "Dead on arrival", "Nursing home"),  "In-hospital",
  DeathInst %in% c("Home", "Other"),                   "Out-of-hospital",
  default = NA_character_
)]

medical_vals <- c(
  "Autopsy",
  "Post mortem examination",
  "Opinion of attending medical practitioner",
  "Opinion of attending medical practitioner on duty",
  "Opinion of registered professional nurse",
  "Autopsy results may be available later"
)

dt_raw[, `:=`(
  is_garbage    = garbage_flag > 0L,
  is_illdefined = garbage_flag == 1L,
  is_nonmedical = !(as.character(Ascertainment) %in% medical_vals)
)]

dt_dq <- dt_raw[DeathYear %in% 2015:2022 & !is.na(facility_type)]
```

### Summary: Which Facility Type Has the Worst Data Quality? {#facility-dq-summary}

```{r}
#| label: dq-summary-table
#| echo: false

summary_dq <- dt_dq[, .(
  `Deaths (N)`            = comma(.N),
  `Any garbage CoD (%)`   = round(mean(is_garbage)    * 100, 1),
  `Ill-defined CoD (%)`   = round(mean(is_illdefined) * 100, 1),
  `Non-medical cert. (%)` = round(mean(is_nonmedical) * 100, 1)
), by = .(facility_type)][order(-`Any garbage CoD (%)`)]

setnames(summary_dq, "facility_type", "Facility type")

flextable(as.data.frame(summary_dq)) |>
  set_caption("Data quality indicators by facility type, South Africa 2015–2022 (pooled)") |>
  bold(part = "header") |>
  bg(i = 1, bg = "#FFF3E0", part = "body") |>
  autofit()
```

::: {.callout-warning}
The facility type with the **highest** any-garbage percentage (top row) is the primary focus for data quality improvement efforts.
:::

---

### National Trends: In-Hospital vs. Out-of-Hospital {#national-dq-trends}

Three quality metrics tracked annually, 2015–2022.

```{r}
#| label: national-dq-panel
#| echo: false
#| fig-height: 5

nat_dq <- dt_dq[, .(
  n              = .N,
  pct_garbage    = mean(is_garbage)    * 100,
  pct_illdefined = mean(is_illdefined) * 100,
  pct_nonmedical = mean(is_nonmedical) * 100
), by = .(DeathYear, facility_type)]

nat_long <- melt(
  nat_dq,
  id.vars      = c("DeathYear", "facility_type", "n"),
  measure.vars = c("pct_garbage", "pct_illdefined", "pct_nonmedical"),
  variable.name = "indicator", value.name = "pct"
)

nat_long[, indicator_label := fcase(
  indicator == "pct_garbage",    "Any garbage CoD (%) [CoD073]",
  indicator == "pct_illdefined", "Ill-defined CoD (%) [CoD045]",
  indicator == "pct_nonmedical", "Non-medical ascertainment (%)"
)]

nat_long[, tooltip_text := paste0(
  facility_type, "\n",
  "Year: ", DeathYear, "\n",
  round(pct, 1), "%\n",
  "N = ", comma(n)
)]

# Mahapatra (2007) 20% target applies only to the garbage / ill-defined facets
target_nat <- data.table(
  indicator_label = c("Any garbage CoD (%) [CoD073]",
                      "Ill-defined CoD (%) [CoD045]"),
  yint = dq_target,
  xstar = 2015
)

p_nat <- ggplot(nat_long,
  aes(x = DeathYear, y = pct, colour = facility_type,
      group = facility_type, text = tooltip_text)) +
  geom_hline(data = target_nat, aes(yintercept = yint),
             linetype = dq_target_lty, colour = dq_target_colour,
             linewidth = 0.7, inherit.aes = FALSE) +
  geom_point(data = target_nat, aes(x = xstar, y = yint),
             shape = 8, colour = dq_target_colour, size = 3,
             inherit.aes = FALSE) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.2) +
  facet_wrap(~indicator_label, ncol = 3, scales = "free_y") +
  scale_colour_manual(
    values = c("In-hospital" = "#1976D2", "Out-of-hospital" = "#E64A19")
  ) +
  scale_x_continuous(breaks = 2015:2022) +
  labs(x = NULL, y = "Percentage (%)", colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position  = "bottom",
    axis.text.x      = element_text(angle = 45, hjust = 1),
    strip.text       = element_text(size = 10, face = "bold"),
    panel.spacing    = unit(1.2, "lines")
  )

ggplotly(p_nat, tooltip = "text") |>
  layout(hovermode = "closest", legend = list(orientation = "h", y = -0.2))
```

::: {.callout-tip appearance="simple"}
The **red dotted line / red star (★)** marks the standardized **20% target**: the WHO/GBD usability cut-point above which cause-of-death data are considered **low quality** because too many deaths are assigned to ill-defined / garbage codes [@mathers2005; @mahapatra2007; @naghavi2010]. Values **below** the line are usable; **< 10%** is the high-quality benchmark [@mathers2005]. The target applies to the garbage and ill-defined panels only.
:::

---

### Provincial Breakdown: In-Hospital vs. Out-of-Hospital {#provincial-dq}

Each panel below shows all nine provinces. Lines track each facility type across 2015–2022.

::: {.panel-tabset}

#### Any Garbage CoD (%)

```{r}
#| label: prov-garbage-panel
#| echo: false
#| fig-height: 9

prov_dq <- dt_dq[, .(
  n              = .N,
  pct_garbage    = mean(is_garbage)    * 100,
  pct_illdefined = mean(is_illdefined) * 100,
  pct_nonmedical = mean(is_nonmedical) * 100
), by = .(DeathYear, DeathProvince, facility_type)]

p_prov_gg <- ggplot(prov_dq,
  aes(x = DeathYear, y = pct_garbage, colour = facility_type,
      group = facility_type,
      text = paste0(facility_type, "\n", DeathProvince,
                    "\nYear: ", DeathYear,
                    "\n", round(pct_garbage, 1), "% garbage\n",
                    "N = ", comma(n)))) +
  geom_hline(yintercept = dq_target, linetype = dq_target_lty,
             colour = dq_target_colour, linewidth = 0.6) +
  annotate("point", x = 2015, y = dq_target, shape = 8,
           colour = dq_target_colour, size = 2.5) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.8) +
  facet_wrap(~DeathProvince, ncol = 3) +
  scale_colour_manual(
    values = c("In-hospital" = "#1976D2", "Out-of-hospital" = "#E64A19")
  ) +
  scale_x_continuous(breaks = c(2015, 2017, 2019, 2021),
                     labels = c("'15","'17","'19","'21")) +
  labs(x = NULL, y = "Any garbage CoD (%)", colour = NULL) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    axis.text.x     = element_text(angle = 45, hjust = 1),
    strip.text      = element_text(size = 9, face = "bold"),
    panel.spacing   = unit(0.8, "lines")
  )

ggplotly(p_prov_gg, tooltip = "text") |>
  layout(hovermode = "closest", legend = list(orientation = "h", y = -0.1))
```

#### Ill-Defined CoD (%)

```{r}
#| label: prov-illdefined-panel
#| echo: false
#| fig-height: 9

p_prov_id <- ggplot(prov_dq,
  aes(x = DeathYear, y = pct_illdefined, colour = facility_type,
      group = facility_type,
      text = paste0(facility_type, "\n", DeathProvince,
                    "\nYear: ", DeathYear,
                    "\n", round(pct_illdefined, 1), "% ill-defined\n",
                    "N = ", comma(n)))) +
  geom_hline(yintercept = dq_target, linetype = dq_target_lty,
             colour = dq_target_colour, linewidth = 0.6) +
  annotate("point", x = 2015, y = dq_target, shape = 8,
           colour = dq_target_colour, size = 2.5) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.8) +
  facet_wrap(~DeathProvince, ncol = 3) +
  scale_colour_manual(
    values = c("In-hospital" = "#1976D2", "Out-of-hospital" = "#E64A19")
  ) +
  scale_x_continuous(breaks = c(2015, 2017, 2019, 2021),
                     labels = c("'15","'17","'19","'21")) +
  labs(x = NULL, y = "Ill-defined CoD (%)", colour = NULL) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    axis.text.x     = element_text(angle = 45, hjust = 1),
    strip.text      = element_text(size = 9, face = "bold"),
    panel.spacing   = unit(0.8, "lines")
  )

ggplotly(p_prov_id, tooltip = "text") |>
  layout(hovermode = "closest", legend = list(orientation = "h", y = -0.1))
```

#### Non-Medical Ascertainment (%)

```{r}
#| label: prov-nonmedical-panel
#| echo: false
#| fig-height: 9

p_prov_nm <- ggplot(prov_dq,
  aes(x = DeathYear, y = pct_nonmedical, colour = facility_type,
      group = facility_type,
      text = paste0(facility_type, "\n", DeathProvince,
                    "\nYear: ", DeathYear,
                    "\n", round(pct_nonmedical, 1), "% non-medical\n",
                    "N = ", comma(n)))) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.8) +
  facet_wrap(~DeathProvince, ncol = 3) +
  scale_colour_manual(
    values = c("In-hospital" = "#1976D2", "Out-of-hospital" = "#E64A19")
  ) +
  scale_x_continuous(breaks = c(2015, 2017, 2019, 2021),
                     labels = c("'15","'17","'19","'21")) +
  labs(x = NULL, y = "Non-medical ascertainment (%)", colour = NULL) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    axis.text.x     = element_text(angle = 45, hjust = 1),
    strip.text      = element_text(size = 9, face = "bold"),
    panel.spacing   = unit(0.8, "lines")
  )

ggplotly(p_prov_nm, tooltip = "text") |>
  layout(hovermode = "closest", legend = list(orientation = "h", y = -0.1))
```

:::

::: {.callout-important icon=false}
## So what? — Decision this data informs

**Monitored action: completion of the *Medical Certification of Cause of Death (MCCD)* course on the Knowledge Hub.**

Poor cause-of-death quality is driven mainly by how certifiers complete the MCCD. These panels turn that into a concrete decision:

- **Trigger:** any facility type / province sitting **above the 20% target** (red dotted line / ★) on the *garbage* or *ill-defined* panels — the WHO/GBD threshold above which cause-of-death data are deemed low quality / unusable [@mathers2005; @mahapatra2007; @naghavi2010].
- **Decision:** prioritise that facility type / province for **MCCD course enrolment and completion** on the Knowledge Hub.
- **Expected effect:** garbage and ill-defined percentages move **toward and below** the red target line in subsequent years.
- **Monitor:** re-render this dashboard each cycle (see the *Report generated* timestamp at the top) and track movement of the In-hospital (blue) vs Out-of-hospital (orange) lines relative to the target.

A persistent gap between In-hospital and Out-of-hospital lines signals a structural certification deficit — typically community/home deaths lacking a medically certified cause — and indicates where MCCD training and support will yield the largest quality gains.

*Indicator definitions:* **Any garbage CoD** = underlying cause too vague or implausible to be informative (WHO garbage code framework); **Ill-defined CoD** = ICD-10 Chapter XVIII (R00–R99) symptoms/signs (CoD045 proxy); **Non-medical ascertainment** = death attested by a family member, "Other", or left unspecified (no Mahapatra target applies).
:::

---

## References {.unnumbered}

::: {#refs}
:::

