---
title: "Rob's Request: Adjusted CYP and Recorded Births"
description: "Redistributing long-acting contraceptive protection across calendar years and comparing it with recorded births."
date: "2026-09-21"
categories: [DHIS2, births, contraception, CYP, sensitivity]
---

```{r}
#| label: setup
#| include: false

suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(plotly)
  library(scales)
  library(DT)
})

rda_path <- here::here("projects/dhis_hospital_mortality/dhis_results.rda")
if (!file.exists(rda_path)) {
  stop("Run projects/dhis_hospital_mortality/wrangling.R first.")
}
load(rda_path)

theme_mrc <- theme_bw(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "#e8edf2"),
    legend.position = "bottom"
  )

ply_config <- function(plot) {
  plot |> config(
    displaylogo = FALSE,
    modeBarButtonsToRemove = c("lasso2d", "select2d")
  )
}
```

## At a glance: objectives, adjustments, outputs

This page tests Rob's proposed change to the CYP calculation against the
original method used in the [main indicators
report](index.qmd#couple-years-of-protection-cyp), where each service is
credited its full CYP factor in the year it was delivered.

```{r}
#| label: tbl-objectives-adjustments-outputs
#| tbl-cap: "What each part of this analysis tests, what it changes relative to the original CYP method, and where to find the result."

objectives_table <- data.table(
  objective = c(
    "Isolate the effect of changing WHEN protection is credited",
    "Test Rob's full proposal (timing change + revised factors)",
    "Check sensitivity to the 2015 DHIS start (left-truncation)",
    "Test whether adjusted CYP tracks recorded births better than the original"
  ),
  adjustment = c(
    "Timing only: spread implant/IUCD protection across protection years, original pill/implant factors kept unchanged",
    "Full proposal: spread implant/IUCD protection AND revise factors (pill 12 cycles = 1 CYP, implant 2.5 total CYP)",
    "Compare observed-cohorts-only vs a steady-state warm-up that back-fills pre-2015 insertions",
    "Correlate each CYP series (original, timing-only, full proposal) against births in year N-1 and N+1"
  ),
  output = c(
    "Trend chart below; 'Timing' columns in the annual difference table",
    "Trend chart below; 'Full proposal' columns in the annual difference table, method-level difference chart",
    "Trend chart below (both new series plotted together)",
    "Correlation table and CYP-vs-births scatter plot, later in this page"
  )
)

datatable(
  objectives_table,
  colnames = c("Objective", "Adjustment applied", "Output"),
  rownames = FALSE,
  options = list(dom = "t", pageLength = 10, scrollX = TRUE)
)
```

## Measures compared

The **old measure** is the standard CYP presented in the main DHIS report. It
credits all expected protection from a long-acting method in its insertion
year. The **new measure** spreads implant and IUCD protection across the
calendar years in which it is expected to remain active and applies the
proposed pill and implant factors.

A bridge measure, **timing only**, redistributes implant and IUCD protection
while preserving every old standard factor, isolating the timing change from
the additional factor changes in the full proposal.

::: {.callout-important}
This is an exploratory ecological comparison of two national annual series. A
correlation does not estimate contraceptive effectiveness or establish that
contraception caused the observed birth trend. Both series may be affected by
reporting completeness, method mix, population size and unmeasured changes over
time.
:::

Recorded births are defined as **live births in facilities plus births born
alive before arrival**. Stillbirths are excluded. The standard CYP series is
unchanged from the main report.

```{r}
#| label: tbl-old-new-summary
#| tbl-cap: "The three CYP definitions compared in this analysis."

definition_comparison <- data.table(
  measure = c("Old: standard CYP", "Bridge: timing only", "New: full proposal"),
  calendar_allocation = c(
    "All CYP credited in service year",
    "Implant and IUCD spread across protection years",
    "Implant and IUCD spread across protection years"
  ),
  pill_factor = c("15 cycles = 1 CYP", "15 cycles = 1 CYP", "12 cycles = 1 CYP"),
  implant_total = c("3.5 CYP", "3.5 CYP", "2.5 CYP"),
  purpose = c(
    "Previous analysis and baseline",
    "Isolates the effect of changing timing",
    "Tests all proposed changes together"
  )
)

datatable(
  definition_comparison,
  colnames = c(
    "Measure", "Calendar allocation", "Pill factor",
    "Implant total", "Role in comparison"
  ),
  rownames = FALSE,
  options = list(dom = "t", scrollX = TRUE)
)
```

## Allocation assumptions

The adjusted measure applies each year's method-specific weight to services
recorded in year $N-k$, where $k$ is the protection-year offset. Thus, adjusted
protection in calendar year $N$ is

$$
\operatorname{Adjusted\ CYP}_N =
\sum_m \sum_k \operatorname{Services}_{m,N-k} w_{m,k}.
$$

The full proposal uses an implant profile of $1.00, 0.80, 0.70$, totalling 2.5
CYP. For the timing-only bridge, the same relative profile is scaled to total
3.5 CYP so that only timing changes. The working IUCD profile starts at 0.85
and 0.75 and declines over ten years, totalling 4.6 CYP in both new measures.
Pills use 12 cycles per CYP in the full proposal and the old 15-cycle factor in
the timing-only bridge. These are scenario assumptions, not locally estimated
continuation rates.

::: {.callout-warning}
The full old-versus-new difference combines a timing change with two factor
changes: pill credit rises by 25% per cycle and total implant credit falls by
28.6% per insertion. It must not be interpreted as the effect of continuation
allocation alone. The timing-only bridge provides that separate comparison.
:::

```{r}
#| label: tbl-adjusted-cyp-matrix
#| tbl-cap: "Calendar-year allocation weights used for adjusted CYP. Offset 0 is the distribution or insertion year."

matrix_display <- copy(ADJUSTED_CYP_MATRIX)
matrix_display[, protection_year := paste0("N+", protection_year)]

datatable(
  matrix_display,
  colnames = c("Method", "Protection year", "CYP per service"),
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(pageLength = 15, dom = "Bfrtip", buttons = list("csv", "excel"))
) |>
  formatRound("adjusted_cyp", digits = 3)
```

```{r}
#| label: tbl-adjusted-cyp-matrix-audit
#| tbl-cap: "Check that calendar-year weights sum to the intended total CYP per service."

datatable(
  adjusted_cyp_matrix_audit[, .(
    short_label,
    protection_years,
    adjusted_total_cyp = round(adjusted_total_cyp, 3),
    standard_cyp_factor = round(standard_cyp_factor, 3),
    difference_from_standard = round(difference_from_standard, 3)
  )],
  colnames = c(
    "Method", "Allocated years", "Adjusted total CYP",
    "Standard CYP factor", "Adjusted minus standard"
  ),
  rownames = FALSE,
  options = list(dom = "t", pageLength = 10)
)
```

## Pre-2015 cohorts

The DHIS series begins in 2015. Consequently, active implant and IUCD cohorts
inserted before 2015 cannot be observed. Two versions of the full proposal show
the effect of this boundary:

- **Observed cohorts only:** uses only insertions observed from 2015 onward and
  is therefore left-truncated in early years.
- **Steady-state initialization:** assumes the 2015 method-specific count
  also occurred in each required pre-2015 year. This is a transparent warm-up
  assumption, not an estimate of historical insertions.

The steady-state version is used for the main old-versus-new comparison. The
observed-cohort version is retained only as a boundary sensitivity check.

```{r}
#| label: fig-adjusted-cyp-trend
#| fig-cap: "Old standard CYP, timing-only redistribution, and the full proposed new CYP, 2015–2025. The observed-cohort series shows sensitivity to missing pre-2015 cohorts."

cyp_trend <- copy(cyp_births_aligned_annual)
cyp_trend[, series := fifelse(cyp_definition == "Standard", "Old: standard CYP", scenario)]

trend_plot <- ggplot(
  cyp_trend,
  aes(
    x = year, y = cyp_value, colour = series, group = series,
    text = paste0(series, "<br>Year: ", year,
                  "<br>CYP: ", format(round(cyp_value), big.mark = ","))
  )
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  scale_x_continuous(breaks = 2015:2025) +
  scale_y_continuous(labels = comma) +
  scale_colour_manual(
    values = c(
      "Old: standard CYP" = "#6a3d9a",
      "Timing only: standard factors redistributed" = "#1f78b4",
      "Proposed new: steady-state initialization" = "#1b9e77",
      "Proposed new: observed cohorts only" = "#d95f02"
    ),
    name = NULL
  ) +
  labs(x = NULL, y = "Couple-years of protection") +
  theme_mrc

ggplotly(trend_plot, tooltip = "text") |> ply_config()
```

## Differences from old standard CYP

The timing-only columns isolate the change from moving protection across
calendar years. The full-proposal columns add the changed pill and implant
factors.

```{r}
#| label: tbl-annual-old-new-difference
#| tbl-cap: "Annual absolute and percentage differences from old standard CYP."

annual_comparison <- merge(
  agg_cyp_national_annual_total[, .(year, old_standard_cyp = cyp)],
  dcast(
    agg_adjusted_cyp_annual[
      scenario %in% c(
        "Timing only: standard factors redistributed",
        "Proposed new: steady-state initialization"
      )
    ],
    year ~ scenario,
    value.var = "adjusted_cyp"
  ),
  by = "year"
)
setnames(
  annual_comparison,
  c(
    "Timing only: standard factors redistributed",
    "Proposed new: steady-state initialization"
  ),
  c("timing_only_cyp", "new_proposed_cyp")
)
annual_comparison[, `:=`(
  timing_difference = timing_only_cyp - old_standard_cyp,
  new_difference = new_proposed_cyp - old_standard_cyp
)]
annual_comparison[, `:=`(
  timing_pct_difference = 100 * timing_difference / old_standard_cyp,
  new_pct_difference = 100 * new_difference / old_standard_cyp
)]

datatable(
  annual_comparison[, .(
    year,
    old_standard_cyp = round(old_standard_cyp),
    timing_only_cyp = round(timing_only_cyp),
    timing_difference = round(timing_difference),
    timing_pct_difference = round(timing_pct_difference, 1),
    new_proposed_cyp = round(new_proposed_cyp),
    new_difference = round(new_difference),
    new_pct_difference = round(new_pct_difference, 1)
  )],
  colnames = c(
    "Year", "Old standard", "Timing only", "Timing difference",
    "Timing %", "Full proposal", "Full difference", "Full %"
  ),
  rownames = FALSE,
  extensions = "Buttons",
  options = list(
    pageLength = 12, dom = "Bfrtip", buttons = list("csv", "excel"),
    scrollX = TRUE
  )
) |>
  formatCurrency(
    c("old_standard_cyp", "timing_only_cyp", "timing_difference",
      "new_proposed_cyp", "new_difference"),
    currency = "", interval = 3, mark = ",", digits = 0
  )
```

```{r}
#| label: fig-method-old-new-difference
#| fig-cap: !expr paste0("Method-level contribution to the difference from old standard CYP in ", max(agg_cyp_national_annual$year), ". Positive values increase total CYP; negative values reduce it.")

latest_comparison_year <- max(agg_cyp_national_annual$year)
standard_method_latest <- agg_cyp_national_annual[
  year == latest_comparison_year,
  .(short_label, old_cyp = cyp)
]
alternative_method_latest <- dcast(
  agg_adjusted_cyp_method_annual[
    year == latest_comparison_year &
      scenario %in% c(
        "Timing only: standard factors redistributed",
        "Proposed new: steady-state initialization"
      )
  ],
  short_label ~ scenario,
  value.var = "adjusted_cyp"
)
method_difference <- merge(
  standard_method_latest, alternative_method_latest,
  by = "short_label"
)
setnames(
  method_difference,
  c(
    "Timing only: standard factors redistributed",
    "Proposed new: steady-state initialization"
  ),
  c("timing_only_cyp", "new_proposed_cyp")
)
method_difference_long <- melt(
  method_difference,
  id.vars = c("short_label", "old_cyp"),
  measure.vars = c("timing_only_cyp", "new_proposed_cyp"),
  variable.name = "comparison",
  value.name = "alternative_cyp"
)
method_difference_long[, difference := alternative_cyp - old_cyp]
method_difference_long[, comparison := factor(
  comparison,
  levels = c("timing_only_cyp", "new_proposed_cyp"),
  labels = c("Timing only minus old", "Full proposal minus old")
)]

method_plot <- ggplot(
  method_difference_long,
  aes(
    x = reorder(short_label, difference), y = difference, fill = comparison,
    text = paste0(
      short_label, "<br>", comparison,
      "<br>Difference: ", format(round(difference), big.mark = ","), " CYP"
    )
  )
) +
  geom_hline(yintercept = 0, colour = "grey45") +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("#1f78b4", "#1b9e77"), name = NULL) +
  labs(x = NULL, y = paste0("Difference from old standard CYP in ", latest_comparison_year)) +
  theme_mrc

ggplotly(method_plot, tooltip = "text") |> ply_config()
```

## Alignment with recorded births

Two temporal alignments are evaluated with Pearson's correlation coefficient:

- **Previous-year births:** CYP in year $N$ versus births in year $N-1$, the
  relationship motivating this sensitivity analysis.
- **Following-year births:** CYP in year $N$ versus births in year $N+1$, the
  hypothesised forecasting relationship.

Each estimate uses ten annual pairs. No p-values are reported because the short,
serially ordered time series does not satisfy the independence assumption of a
simple correlation test.

```{r}
#| label: tbl-cyp-birth-correlation
#| tbl-cap: "Pearson correlations between national CYP and recorded births under the two year alignments."

correlation_display <- copy(cyp_birth_correlation)
correlation_display[, series := fifelse(
  cyp_definition == "Standard", "Old: standard CYP", scenario
)]
old_correlations <- correlation_display[series == "Old: standard CYP",
  .(alignment, old_correlation = correlation)]
correlation_display <- old_correlations[correlation_display, on = "alignment"]
correlation_display[, difference_from_old := correlation - old_correlation]

datatable(
  correlation_display[, .(
    series,
    alignment,
    cyp_years = paste0(first_cyp_year, "–", last_cyp_year),
    n_year_pairs,
    correlation = round(correlation, 3),
    difference_from_old = round(difference_from_old, 3)
  )],
  colnames = c(
    "CYP series", "Birth alignment", "CYP years", "Year pairs",
    "Pearson r", "Difference in r from old"
  ),
  rownames = FALSE,
  options = list(dom = "t", pageLength = 10)
)
```

```{r}
#| label: fig-cyp-following-births
#| fig-cap: "CYP in year N against recorded births in year N+1. Lines are descriptive linear fits, not forecasting models."

following_plot_data <- cyp_births_aligned_annual[!is.na(following_year_births)]
following_plot_data <- following_plot_data[
  scenario != "Proposed new: observed cohorts only"
]
following_plot_data[, series := fifelse(
  cyp_definition == "Standard", "Old: standard CYP", scenario
)]

following_plot <- ggplot(
  following_plot_data,
  aes(
    x = cyp_value, y = following_year_births, colour = series,
    text = paste0(
      series, "<br>CYP year: ", year,
      "<br>CYP: ", format(round(cyp_value), big.mark = ","),
      "<br>Birth year: ", year + 1L,
      "<br>Recorded births: ", format(round(following_year_births), big.mark = ",")
    )
  )
) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 0.7) +
  geom_point(size = 2) +
  facet_wrap(~series, scales = "free_x") +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  scale_colour_manual(
    values = c(
      "Old: standard CYP" = "#6a3d9a",
      "Timing only: standard factors redistributed" = "#1f78b4",
      "Proposed new: steady-state initialization" = "#1b9e77"
    ),
    guide = "none"
  ) +
  labs(x = "CYP in year N", y = "Recorded births in year N+1") +
  theme_mrc +
  theme(strip.text = element_text(size = 9))

ggplotly(following_plot, tooltip = "text") |> ply_config()
```

## Correlation with births: old versus new

Births in the following year: old standard CYP $r =
`r sprintf("%.3f", cyp_birth_correlation[cyp_definition == "Standard" & alignment == "Births in following year (N+1)", correlation])`$,
timing only $r =
`r sprintf("%.3f", cyp_birth_correlation[scenario == "Timing only: standard factors redistributed" & alignment == "Births in following year (N+1)", correlation])`$,
full proposal $r =
`r sprintf("%.3f", cyp_birth_correlation[scenario == "Proposed new: steady-state initialization" & alignment == "Births in following year (N+1)", correlation])`$.

Births in the previous year: old standard CYP $r =
`r sprintf("%.3f", cyp_birth_correlation[cyp_definition == "Standard" & alignment == "Births in previous year (N-1)", correlation])`$,
timing only $r =
`r sprintf("%.3f", cyp_birth_correlation[scenario == "Timing only: standard factors redistributed" & alignment == "Births in previous year (N-1)", correlation])`$,
full proposal $r =
`r sprintf("%.3f", cyp_birth_correlation[scenario == "Proposed new: steady-state initialization" & alignment == "Births in previous year (N-1)", correlation])`$.

## Underlying aligned data

```{r}
#| label: tbl-cyp-birth-aligned-data
#| tbl-cap: "Exportable annual data used in the CYP–birth alignment analysis."

aligned_export <- copy(cyp_births_aligned_annual)
aligned_export[, series := fifelse(
  cyp_definition == "Standard", "Old: standard CYP", scenario
)]

datatable(
  aligned_export[, .(
    series, cyp_year = year,
    cyp_value = round(cyp_value),
    previous_year_births = round(previous_year_births),
    following_year_births = round(following_year_births)
  )],
  colnames = c("CYP series", "CYP year", "CYP", "Births in N-1", "Births in N+1"),
  rownames = FALSE,
  filter = "top",
  extensions = "Buttons",
  options = list(pageLength = 15, dom = "Bfrtip", buttons = list("csv", "excel"))
)
```

## Original request

::: {.callout-note collapse="true"}
## Email from Rob (kept verbatim for reference)

While the couple year protection rate (CYPR) is highly (negatively)
correlated with the numbers of births in the previous calendar year from
2016 to 2023 (i.e. when births are high, women go on contraception (and
vice versa) - sort of justifying the trend in births over this period - but
this does not hold for the 2024 births (and 2025 CYPR).

Also, ideally, one would prefer to find an association of the CYPR with
births in the following year, but this is very weak (mainly because of the
way the CYPR is determined - in particular, for example implants are given a
weight of 2.5 in the year of implant and then not included in the following
years).

So I was wondering if you (because you would be much more efficient at doing
it and it would be useful to have the measure available on your GitHub
site) could consider computing an Adjusted CYPR where the numerator is
computed as described below. Then we could check how well it works, and
consider possible refinements.

**Structuring the adjusted forecasting table**

| Contraceptive method | Raw distribution (Year N) | Standard CYP reporting (credited entirely to Year N) | Adjusted allocation (protection timeline) |
|---|---|---|---|
| Oral contraceptive pills | 1200 cycles | 100 CYP (12 cycles = 1 CYP) | 100% credited to Year N (impacts Year N+1 births) |
| Sub-dermal implant | 100 units | 250 CYP (credited completely to Year N) | Year N: 100 CYP; Year N+1: 80 CYP; Year N+2: 70 CYP |
| Copper IUD | 100 units | 460 CYP (credited completely to Year N) | Year N: 85 CYP; Year N+1: 75 CYP; ...and so on for 10 years |

**Apply a continuation decay rate (the matrix).** Do not just divide a
3-year implant evenly by 3 (0.33 per year). Real-world protection decreases
over time because women have implants removed early due to side effects,
planning a pregnancy, or reaching menopause. Instead, construct a
continuation matrix based on local survey data (e.g. Demographic and Health
Surveys) or standard global drop-out assumptions (see the [IDS Heart CYP
guide](https://archive.ids.ac.uk/heart/doc_lib/family-planning-couple-years-of-protection-cyp/index.html)
and the [Data4Impact CYP
page](https://www.data4impactproject.org/prh/family-planning/fp/couple-years-of-protection-cyp)).

- Sub-dermal implants (3-year total CYP factor = 2.5)
- Copper T IUD (10-year total CYP factor = 4.6)
:::
