---
title: "Sensitivity: Condom Utilization in Couple-Years of Protection"
description: "How much does total national CYP change if we assume distributed condoms are not all actually used?"
date: "2026-08-20"
categories: [DHIS2, contraception, CYP, sensitivity]
---

```{r}
#| label: setup
#| include: false

suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(plotly)
  library(scales)
  library(DT)
})

rda_path <- here::here("projects/dhis_hospital_mortality/dhis_results.rda")
if (!file.exists(rda_path)) {
  stop("Run projects/dhis_hospital_mortality/wrangling.R first.")
}
load(rda_path)

theme_mrc <- theme_bw(base_size = 12) +
  theme(panel.grid.minor = element_blank(), strip.background = element_rect(fill = "grey90"))

ply_config <- function(p) {
  p |> config(displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d"))
}
```

## Why this check?

Couple-years of protection (CYP) treats every distributed condom as if it were used correctly for its full protective purpose. Two things make that assumption shakier for condoms than for clinic-administered methods (injectables, implants, IUCD, sterilisation): the DHIS2 figure is **units distributed to facilities**, not confirmed use, and there is no data here on whether distributed condoms were used correctly (or at all).

This page holds the standard USAID/FP2030 factor (120 units = 1 CYP) fixed and instead varies **assumed real-world utilization** — the share of distributed condoms assumed actually used — to show how sensitive total national CYP is to that assumption.

::: {.callout-important}
The primary report ([DHIS2 Hospital Indicators](index.qmd)) always uses the **"Standard (100% used, current default)"** scenario below. This page is a supplementary robustness check — it does not change any number reported elsewhere.
:::

## Scenarios

```{r}
#| label: tbl-scenarios
datatable(
  CONDOM_CYP_SCENARIOS[, .(scenario, utilization = percent(utilization, accuracy = 1))],
  colnames = c("Scenario", "Assumed condom utilization"),
  rownames = FALSE,
  options  = list(dom = "t")
)
```

## Total national CYP under each scenario

```{r}
#| label: fig-condom-sensitivity-total
#| fig-cap: "Total national CYP by year, under each condom utilization scenario. Non-condom methods are unaffected in every scenario."

sens_plot <- ggplot(agg_condom_sensitivity_annual,
    aes(x = year, y = total_cyp, colour = scenario,
        text = paste0(scenario, "<br>Year: ", year,
                      "<br>Total CYP: ", format(round(total_cyp), big.mark = ","),
                      "<br>Condom CYP: ", format(round(condom_cyp), big.mark = ",")))) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(breaks = 2015:2025) +
  scale_colour_manual(values = c(
    "Excluded" = "#d95f02",
    "Conservative (50% used)" = "#7570b3",
    "Standard (100% used, current default)" = "#1b9e77"
  ), name = NULL) +
  labs(x = NULL, y = "Total couple-years of protection") +
  theme_mrc

ggplotly(sens_plot, tooltip = "text") |>
  layout(legend = list(orientation = "h", y = -0.2)) |>
  ply_config()
```

## Latest-year comparison

```{r}
#| label: tbl-condom-sensitivity-latest
#| tbl-cap: "Total national CYP by scenario, most recent year, with % difference from the Standard (current default) scenario."

latest_sens_year <- max(agg_condom_sensitivity_annual$year)
sens_latest <- agg_condom_sensitivity_annual[year == latest_sens_year]
standard_total <- sens_latest[scenario == "Standard (100% used, current default)", total_cyp]
sens_latest[, pct_vs_standard := (total_cyp - standard_total) / standard_total]

datatable(
  sens_latest[order(scenario),
    .(scenario, condom_cyp = round(condom_cyp), total_cyp = round(total_cyp),
      pct_vs_standard = percent(pct_vs_standard, accuracy = 0.1))],
  colnames = c("Scenario", paste0("Condom CYP (", latest_sens_year, ")"),
               paste0("Total CYP (", latest_sens_year, ")"), "% vs Standard"),
  rownames = FALSE,
  options  = list(dom = "t")
)
```
