---
title: "DHIS2 Hospital Indicators — Exploratory Analysis"
description: "Trend exploration of NDoH DHIS2 hospital-level data (births, deaths, contraception) for South Africa, 2015–2025. Exploratory only — not for inference."
date: "2026-06-13"
categories: [DHIS2, hospitals, births, deaths, contraception, exploratory]
---

```{r}
#| label: setup
#| include: false

suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(plotly)
  library(scales)
  library(DT)
  library(htmltools)
  library(lubridate)
})

# ── Load wrangling output ─────────────────────────────────────────────────────
rda_path <- here::here("projects/dhis_hospital_mortality/dhis_results.rda")
if (!file.exists(rda_path)) {
  stop("Run projects/dhis_hospital_mortality/wrangling.R first.")
}
load(rda_path)

# ── Colour palette (consistent across all tabs) ───────────────────────────────
PROV_COLOURS <- c(
  "Eastern Cape"  = "#e41a1c",
  "Free State"    = "#ff7f00",
  "Gauteng"       = "#984ea3",
  "KwaZulu-Natal" = "#377eb8",
  "Limpopo"       = "#4daf4a",
  "Mpumalanga"    = "#a65628",
  "North West"    = "#f781bf",
  "Northern Cape" = "#999999",
  "Western Cape"  = "#1f78b4"
)

# Theme for static ggplots (also used inside ggplotly)
theme_mrc <- theme_bw(base_size = 12) +
  theme(
    panel.grid.minor  = element_blank(),
    strip.background  = element_rect(fill = "#e8edf2"),
    plot.title        = element_text(face = "bold", size = 13),
    plot.subtitle     = element_text(size = 10, colour = "grey40"),
    legend.position   = "bottom"
  )

# ── Shared plotly config: tidy toolbar, hi-res PNG export ──────────────────────
ply_config <- function(p) {
  config(
    p,
    displaylogo = FALSE,
    modeBarButtonsToRemove = c("lasso2d", "select2d", "autoScale2d",
                               "hoverClosestCartesian", "hoverCompareCartesian"),
    toImageButtonOptions = list(format = "png", scale = 2)
  )
}

# ── Small-multiples builder ───────────────────────────────────────────────────
# Faceted, free-y trend panels so EVERY indicator is interpretable at its own
# scale (large- and small-magnitude indicators no longer compete on one axis).
# `time_col` is either "year" (numeric) or "period_date" (Date).
make_small_multiples <- function(dt, time_col,
                                 date_axis = FALSE,
                                 line_col  = "#1f6eb5",
                                 ncol      = 3) {
  dt <- copy(dt)
  dt[, x_plot := get(time_col)]
  dt[, tt := paste0(
    short_label, "<br>",
    if (date_axis) format(x_plot, "%b %Y") else paste("Year:", x_plot),
    "<br>Count: ", format(value, big.mark = ",")
  )]

  g <- ggplot(dt, aes(x = x_plot, y = value, group = short_label, text = tt)) +
    geom_line(colour = line_col, linewidth = 0.6) +
    geom_point(colour = line_col, size = 1.1) +
    facet_wrap(~ short_label, scales = "free_y", ncol = ncol) +
    scale_y_continuous(labels = comma) +
    labs(x = NULL, y = "Count") +
    theme_mrc +
    theme(legend.position = "none",
          strip.text  = element_text(size = 8),
          axis.text.x = element_text(angle = 45, hjust = 1, size = 7))

  g <- if (date_axis) {
    g + scale_x_date(date_breaks = "6 months", date_labels = "%b %y")
  } else {
    g + scale_x_continuous(breaks = seq(2015, 2025, 2))
  }

  ggplotly(g, tooltip = "text") |> ply_config()
}

# ── Province explorer builder ─────────────────────────────────────────────────
# One indicator at a time (dropdown) across all provinces, with a Linear/Log
# y-axis toggle so small provinces remain interpretable against large ones.
# Unified hover compares all provinces at the selected year.
make_province_explorer <- function(dt, domain_label) {
  dt        <- dt[!is.na(province)]
  indicators <- unique(dt$short_label)
  provs     <- sort(unique(dt$province))
  n_provs   <- length(provs)

  fig <- plot_ly()
  for (ind in indicators) {
    df_i <- dt[short_label == ind]
    for (pv in provs) {
      df_p <- df_i[province == pv]
      fig <- add_trace(
        fig,
        data        = df_p,
        x = ~year, y = ~value,
        name        = pv,
        type = "scatter", mode = "lines+markers",
        visible     = (ind == indicators[1]),
        legendgroup = pv,
        showlegend  = (ind == indicators[1]),
        line   = list(color = unname(PROV_COLOURS[pv]), width = 2),
        marker = list(color = unname(PROV_COLOURS[pv]), size = 5),
        hovertemplate = paste0("<b>", pv, "</b><br>", ind,
                               "<br>%{x}: %{y:,}<extra></extra>")
      )
    }
  }

  # Indicator dropdown — toggles which block of province traces is visible
  ind_buttons <- lapply(seq_along(indicators), function(i) {
    vis <- rep(FALSE, length(indicators) * n_provs)
    vis[((i - 1) * n_provs + 1):(i * n_provs)] <- TRUE
    list(method = "update", label = indicators[i],
         args = list(list(visible = vis),
                     list(title = paste0(domain_label, ": ", indicators[i],
                                         " — by province"))))
  })

  # Linear / Log y-axis toggle
  scale_buttons <- list(
    list(method = "relayout", label = "Linear",
         args = list(list(`yaxis.type` = "linear",
                          `yaxis.title.text` = "Count"))),
    list(method = "relayout", label = "Log",
         args = list(list(`yaxis.type` = "log",
                          `yaxis.title.text` = "Count (log scale)")))
  )

  fig |>
    layout(
      title     = paste0(domain_label, ": ", indicators[1], " — by province"),
      hovermode = "x unified",
      xaxis     = list(title = "", dtick = 1, showspikes = TRUE,
                       spikemode = "across", spikethickness = 1),
      yaxis     = list(title = "Count", tickformat = ","),
      legend    = list(orientation = "h", y = -0.22),
      margin    = list(t = 90),
      updatemenus = list(
        list(type = "dropdown", x = 0.0, xanchor = "left", y = 1.22,
             buttons = ind_buttons, showactive = TRUE,
             pad = list(t = 2, b = 2)),
        list(type = "buttons", direction = "right",
             x = 0.62, xanchor = "left", y = 1.22,
             buttons = scale_buttons, showactive = TRUE,
             pad = list(t = 2, b = 2))
      ),
      annotations = list(
        list(text = "Indicator:", x = 0.0, xref = "paper",
             y = 1.30, yref = "paper", xanchor = "left",
             showarrow = FALSE, font = list(size = 11, color = "grey40")),
        list(text = "y-axis:", x = 0.62, xref = "paper",
             y = 1.30, yref = "paper", xanchor = "left",
             showarrow = FALSE, font = list(size = 11, color = "grey40"))
      )
    ) |>
    ply_config()
}
```

::: {.callout-note}
**Exploratory only.** This page presents DHIS2 hospital data supplied by Pam Groenewald (NDoH). The data have not been quality-audited. Trends shown are descriptive; no causal or inferential claims should be drawn. For comparison with official MACOD data, see the dedicated comparison section (planned).
:::

## Data provenance & audit trail

```{r}
#| label: audit-table
#| echo: true

# Print the audit record so the page self-documents its inputs
cat("Source files:\n")
cat("  ", audit$source_annual,  "\n")
cat("  ", audit$source_monthly, "\n\n")
cat("Annual sheets read    :", paste(audit$annual_sheets, collapse = " | "), "\n")
cat("Monthly sheets read   :", paste(audit$monthly_sheets, collapse = " | "), "\n")
cat("Annual year range     :", audit$year_range_annual[1], "–", audit$year_range_annual[2], "\n")
cat("Monthly period range  :", format(audit$period_range_monthly[1]),
    "–", format(audit$period_range_monthly[2]), "\n")
cat("Wrangling run at      :", format(audit$run_timestamp), "\n\n")
cat("Dataset sizes (long rows):\n")
cat("  dhis_births :", nrow(dhis_births), "\n")
cat("  dhis_deaths :", nrow(dhis_deaths), "\n")
cat("  dhis_contra :", nrow(dhis_contra), "\n")
```

## Indicator reference

The 27 DHIS2 indicators are grouped into three domains. This table is the single source of truth used to split the data.

```{r}
#| label: indicator-table
datatable(
  indicator_meta[, .(domain, short_label, indicator)][order(domain, short_label)],
  colnames    = c("Domain", "Short label", "Full DHIS2 indicator name"),
  rownames    = FALSE,
  options     = list(pageLength = 30, dom = "ftp"),
  caption     = "Table 1. Indicator domain map"
)
```

---

## Reporting completeness {#sec-completeness}

Before interpreting any trend, inspect how many facilities were submitting data in each period. Spikes or drops in counts indicate data quality issues rather than true events.

```{r}
#| label: fig-completeness
#| fig-cap: "Annual reporting completeness — number of distinct facility–province–district combinations submitting data per year and domain. Expect stable or slowly growing counts; sudden drops signal missing data."
#| fig-height: 5

# Three-panel: one per domain
comp_plot <- ggplot(agg_completeness_annual,
                    aes(x = year, y = n_facilities, fill = domain,
                        text = paste0("Domain: ", domain,
                                      "<br>Year: ", year,
                                      "<br>Facilities: ", format(n_facilities, big.mark = ",")))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~domain, ncol = 3, scales = "free_y",
             labeller = labeller(domain = c(
               births = "Births / deliveries / BCG",
               deaths = "Deaths",
               contra = "Contraception / reproductive"
             ))) +
  scale_x_continuous(breaks = 2015:2025) +
  scale_fill_manual(values = c(births = "#2196F3", deaths = "#f44336", contra = "#4CAF50")) +
  labs(x = NULL, y = "Facilities reporting") +
  theme_mrc +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplotly(comp_plot, tooltip = "text") |>
  layout(margin = list(b = 60)) |>
  ply_config()
```

---

## Births, deliveries & BCG {#sec-births}

### National annual trends

```{r}
#| label: fig-births-national
#| fig-cap: "National annual counts for each births/delivery/BCG indicator, 2015–2025. Each panel has its OWN y-axis (free scale) so small-volume indicators (e.g. teen deliveries) are interpretable alongside high-volume ones (e.g. live births). Hover for exact counts. 2026 is excluded (incomplete year)."
#| fig-height: 7

make_small_multiples(
  agg_national_annual[domain == "births"],
  time_col = "year",
  line_col = "#1f6eb5"
)
```

### By province

```{r}
#| label: fig-births-province
#| fig-cap: "Provincial annual counts. Pick an indicator from the dropdown and toggle Linear / Log on the y-axis — the log view keeps low-volume provinces (e.g. Northern Cape) interpretable against high-volume ones (Gauteng, KZN). Hover compares all provinces at a year."

make_province_explorer(
  agg_prov_annual[domain == "births"],
  domain_label = "Births / deliveries"
)
```

---

## Deaths {#sec-deaths}

### National annual trends

```{r}
#| label: fig-deaths-national
#| fig-cap: "National annual counts for each death indicator, 2015–2025, on independent (free) y-axes so low-count indicators (maternal death, 7–28 day deaths) are interpretable alongside the much larger total. 2026 is excluded (incomplete year). Note: 'Inpatient deaths (total)' is a separately reported aggregate and is NOT the sum of the age-specific panels."
#| fig-height: 7

make_small_multiples(
  agg_national_annual[domain == "deaths"],
  time_col = "year",
  line_col = "#c0392b"
)
```

### Monthly trend (Apr 2024 – Mar 2026)

Monthly data from the second file allow finer temporal resolution over the most recent period. Each indicator is shown on its own scale so seasonality and short-term shifts are visible regardless of magnitude.

```{r}
#| label: fig-deaths-monthly
#| fig-cap: "Monthly national counts for each death indicator (Apr 2024 – Mar 2026), independent y-axes. Hover for month-level values."
#| fig-height: 7

make_small_multiples(
  agg_national_monthly[domain == "deaths"],
  time_col  = "period_date",
  date_axis = TRUE,
  line_col  = "#c0392b"
)
```

### By province

```{r}
#| label: fig-deaths-province
#| fig-cap: "Provincial annual death counts. Pick an indicator and toggle Linear / Log. Log view recommended when comparing provinces of very different size."

make_province_explorer(
  agg_prov_annual[domain == "deaths"],
  domain_label = "Deaths"
)
```

---

## Contraception & reproductive {#sec-contra}

### National annual trends

```{r}
#| label: fig-contra-national
#| fig-cap: "National annual counts for each contraception / reproductive indicator, 2015–2025, on independent (free) y-axes so rarer procedures (e.g. male sterilisation, TOP) are interpretable alongside high-volume methods (oral pill, injectables). 2026 is excluded (incomplete year)."
#| fig-height: 8

make_small_multiples(
  agg_national_annual[domain == "contra"],
  time_col = "year",
  line_col = "#1e8449"
)
```

### Monthly trend (Apr 2024 – Mar 2026)

```{r}
#| label: fig-contra-monthly
#| fig-cap: "Monthly national counts for each contraception indicator (Apr 2024 – Mar 2026), independent y-axes."
#| fig-height: 8

make_small_multiples(
  agg_national_monthly[domain == "contra"],
  time_col  = "period_date",
  date_axis = TRUE,
  line_col  = "#1e8449"
)
```

### By province

```{r}
#| label: fig-contra-province
#| fig-cap: "Provincial annual contraception counts. Pick an indicator and toggle Linear / Log on the y-axis."

make_province_explorer(
  agg_prov_annual[domain == "contra"],
  domain_label = "Contraception"
)
```

---

## Facility-level raw data browser {#sec-facility}

Search, sort and filter the full annual dataset at facility level.

```{r}
#| label: tbl-facility
#| tbl-cap: "Facility-level annual totals. Use the search box and column filters to explore specific provinces, districts or indicators."

datatable(
  agg_facility_totals[, .(domain, short_label, province, district, facility,
                           total_value, years_reported)],
  colnames  = c("Domain", "Indicator", "Province", "District", "Facility",
                "Total (all years)", "Years reporting"),
  rownames  = FALSE,
  filter    = "top",
  extensions = "Buttons",
  options   = list(
    pageLength = 20,
    dom        = "Bfrtip",
    buttons    = list("csv", "excel"),
    scrollX    = TRUE,
    columnDefs = list(list(className = "dt-right", targets = c(5, 6)))
  )
) |>
  formatCurrency("total_value", currency = "", interval = 3, mark = ",", digits = 0)
```

---

## Notes on methodology

### Data sources
- **Annual file**: `DHIS DATA 2015_2026 (2).xls` — six sheets covering 2015–2026 in wide format (one column per year). The 2022–2026 period is split across three sheets by indicator domain.
- **Monthly file**: `DHIS DATA APR 2024_MAR 2026_test.xls` — four sheets with monthly columns (e.g. "April 2024").

### Processing steps (see `wrangling.R`)
1. Each sheet read with `readxl::read_excel(..., col_types = "text")` to prevent type coercion errors.
2. Fixed columns 1–4 (Province, District, Sub-district, Facility) cleaned by stripping the two-letter DHIS2 province code prefix and trailing administrative suffixes.
3. Value columns melted to long format; values coerced to numeric after removing non-numeric characters.
4. Indicators classified into three domains (births, deaths, contra) via `indicator_meta` lookup table.
5. Monthly sheets de-duplicated on facility + indicator + period after binding.
6. **The 2026 calendar year is dropped from the annual series** (`year <= 2025`) because it is incomplete and would otherwise show a spurious partial-year drop. Monthly data, which is complete per-month through Mar 2026, is retained in the monthly figures.
7. No imputation of missing values; `NA` counts are excluded from aggregations.

### Limitations
- Annual counts in the main file are calendar-year aggregates; months within a year are not individually available before April 2024.
- The annual series ends in 2025; 2026 is excluded as incomplete. The monthly figures extend to Mar 2026.
- The `Inpatient deaths - total` indicator is a separately reported aggregate that may not equal the sum of age-specific death indicators — they are shown together but not summed.
- Facility coverage varies by year (see [Reporting completeness](#sec-completeness)). Declining counts in recent years are likely incomplete reporting, not true declines.
- These data have not been validated against the official MACOD or StatsSA vital registration series.
