---
title: "DHIS2 Hospital Indicators — Exploratory Analysis"
description: "Trend exploration of NDoH DHIS2 hospital-level data (births, deaths, contraception) for South Africa, 2015–2025. Exploratory only — not for inference."
date: "2026-06-13"
categories: [DHIS2, hospitals, births, deaths, contraception, exploratory]
---

```{r}
#| label: setup
#| include: false

suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(plotly)
  library(scales)
  library(DT)
  library(htmltools)
  library(lubridate)
})

# ── Load wrangling output ─────────────────────────────────────────────────────
rda_path <- here::here("projects/dhis_hospital_mortality/dhis_results.rda")
if (!file.exists(rda_path)) {
  stop("Run projects/dhis_hospital_mortality/wrangling.R first.")
}
load(rda_path)

# Monthly provincial aggregate used to prioritize month-level province views.
if (!exists("agg_prov_monthly")) {
  agg_prov_monthly <- dhis_monthly[!is.na(province) & province != "",
    .(value = sum(value, na.rm = TRUE)),
    by = .(domain, indicator, short_label, province, year, month, period_date)
  ][order(domain, indicator, province, period_date)]
}

# ── Colour palette (consistent across all tabs) ───────────────────────────────
PROV_COLOURS <- c(
  "Eastern Cape"  = "#e41a1c",
  "Free State"    = "#ff7f00",
  "Gauteng"       = "#984ea3",
  "KwaZulu-Natal" = "#377eb8",
  "Limpopo"       = "#4daf4a",
  "Mpumalanga"    = "#a65628",
  "North West"    = "#f781bf",
  "Northern Cape" = "#999999",
  "Western Cape"  = "#1f78b4"
)

# Theme for static ggplots (also used inside ggplotly)
theme_mrc <- theme_bw(base_size = 12) +
  theme(
    panel.grid.minor  = element_blank(),
    strip.background  = element_rect(fill = "#e8edf2"),
    plot.title        = element_text(face = "bold", size = 13),
    plot.subtitle     = element_text(size = 10, colour = "grey40"),
    legend.position   = "bottom"
  )

# ── Shared plotly config: tidy toolbar, hi-res PNG export ──────────────────────
ply_config <- function(p) {
  config(
    p,
    displaylogo = FALSE,
    modeBarButtonsToRemove = c("lasso2d", "select2d", "autoScale2d",
                               "hoverClosestCartesian", "hoverCompareCartesian"),
    toImageButtonOptions = list(format = "png", scale = 2)
  )
}

# ── Small-multiples builder ───────────────────────────────────────────────────
# Faceted, free-y trend panels so EVERY indicator is interpretable at its own
# scale (large- and small-magnitude indicators no longer compete on one axis).
# `time_col` is either "year" (numeric) or "period_date" (Date).
make_small_multiples <- function(dt, time_col,
                                 date_axis = FALSE,
                                 line_col  = "#1f6eb5",
                                 ncol      = 3) {
  dt <- copy(dt)
  dt[, x_plot := get(time_col)]
  dt[, tt := paste0(
    short_label, "<br>",
    if (date_axis) format(x_plot, "%b %Y") else paste("Year:", x_plot),
    "<br>Count: ", format(value, big.mark = ",")
  )]

  g <- ggplot(dt, aes(x = x_plot, y = value, group = short_label, text = tt)) +
    geom_line(colour = line_col, linewidth = 0.6) +
    geom_point(colour = line_col, size = 1.1) +
    facet_wrap(~ short_label, scales = "free_y", ncol = ncol) +
    scale_y_continuous(labels = comma) +
    labs(x = NULL, y = "Count") +
    theme_mrc +
    theme(legend.position = "none",
          strip.text  = element_text(size = 8),
          axis.text.x = element_text(angle = 45, hjust = 1, size = 7))

  g <- if (date_axis) {
    g + scale_x_date(date_breaks = "6 months", date_labels = "%b %y")
  } else {
    g + scale_x_continuous(breaks = seq(2015, 2025, 2))
  }

  ggplotly(g, tooltip = "text") |> ply_config()
}

# ── Province explorer builder ─────────────────────────────────────────────────
# One indicator at a time (dropdown) across all provinces, with a Linear/Log
# y-axis toggle so small provinces remain interpretable against large ones.
# Works for annual or monthly series.
make_province_explorer <- function(dt, domain_label,
                                   x_col = "year",
                                   date_axis = FALSE) {
  dt        <- dt[!is.na(province)]
  dt[, x_plot := get(x_col)]
  dt[, hover_text := if (date_axis) {
    paste0("<b>", province, "</b><br>", short_label,
           "<br>", format(x_plot, "%b %Y"), ": ", format(value, big.mark = ","))
  } else {
    paste0("<b>", province, "</b><br>", short_label,
           "<br>", x_plot, ": ", format(value, big.mark = ","))
  }]
  indicators <- unique(dt$short_label)
  provs     <- sort(unique(dt$province))
  n_provs   <- length(provs)

  fig <- plot_ly()
  for (ind in indicators) {
    df_i <- dt[short_label == ind]
    for (pv in provs) {
      df_p <- df_i[province == pv]
      fig <- add_trace(
        fig,
        data        = df_p,
        x = ~x_plot, y = ~value,
        text        = ~hover_text,
        name        = pv,
        type = "scatter", mode = "lines+markers",
        visible     = (ind == indicators[1]),
        legendgroup = pv,
        showlegend  = (ind == indicators[1]),
        line   = list(color = unname(PROV_COLOURS[pv]), width = 2),
        marker = list(color = unname(PROV_COLOURS[pv]), size = 5),
        hovertemplate = "%{text}<extra></extra>"
      )
    }
  }

  # Indicator dropdown — toggles which block of province traces is visible
  ind_buttons <- lapply(seq_along(indicators), function(i) {
    vis <- rep(FALSE, length(indicators) * n_provs)
    vis[((i - 1) * n_provs + 1):(i * n_provs)] <- TRUE
    list(method = "update", label = indicators[i],
         args = list(list(visible = vis),
                     list(title = paste0(domain_label, ": ", indicators[i],
                                         " — by province"))))
  })

  # Linear / Log y-axis toggle
  scale_buttons <- list(
    list(method = "relayout", label = "Linear",
         args = list(list(`yaxis.type` = "linear",
                          `yaxis.title.text` = "Count"))),
    list(method = "relayout", label = "Log",
         args = list(list(`yaxis.type` = "log",
                          `yaxis.title.text` = "Count (log scale)")))
  )

  fig |>
    layout(
      title     = paste0(domain_label, ": ", indicators[1], " — by province"),
      hovermode = "x unified",
      xaxis     = if (date_axis) {
        list(title = "", tickformat = "%b %y", showspikes = TRUE,
             spikemode = "across", spikethickness = 1)
      } else {
        list(title = "", dtick = 1, showspikes = TRUE,
             spikemode = "across", spikethickness = 1)
      },
      yaxis     = list(title = "Count", tickformat = ","),
      legend    = list(orientation = "h", y = -0.22),
      margin    = list(t = 90),
      updatemenus = list(
        list(type = "dropdown", x = 0.0, xanchor = "left", y = 1.22,
             buttons = ind_buttons, showactive = TRUE,
             pad = list(t = 2, b = 2)),
        list(type = "buttons", direction = "right",
             x = 0.62, xanchor = "left", y = 1.22,
             buttons = scale_buttons, showactive = TRUE,
             pad = list(t = 2, b = 2))
      ),
      annotations = list(
        list(text = "Indicator:", x = 0.0, xref = "paper",
             y = 1.30, yref = "paper", xanchor = "left",
             showarrow = FALSE, font = list(size = 11, color = "grey40")),
        list(text = "y-axis:", x = 0.62, xref = "paper",
             y = 1.30, yref = "paper", xanchor = "left",
             showarrow = FALSE, font = list(size = 11, color = "grey40"))
      )
    ) |>
    ply_config()
}
```

::: {.callout-note}
**Exploratory only.** This page presents DHIS2 hospital data supplied by Pam Groenewald (NDoH). The data have not been quality-audited. Trends shown are descriptive; no causal or inferential claims should be drawn. For comparison with official MACOD data, see the dedicated comparison section (planned).
:::

## Data provenance & audit trail

```{r}
#| label: audit-table
#| echo: true

# Print the audit record so the page self-documents its inputs
cat("Source files:\n")
cat("  ", audit$source_annual,  "\n")
cat("  ", audit$source_monthly, "\n\n")
cat("Annual sheets read    :", paste(audit$annual_sheets, collapse = " | "), "\n")
cat("Monthly sheets read   :", paste(audit$monthly_sheets, collapse = " | "), "\n")
cat("Annual year range     :", audit$year_range_annual[1], "–", audit$year_range_annual[2], "\n")
cat("Monthly period range  :", format(audit$period_range_monthly[1]),
    "–", format(audit$period_range_monthly[2]), "\n")
cat("Wrangling run at      :", format(audit$run_timestamp), "\n\n")
cat("Dataset sizes (long rows):\n")
cat("  dhis_births :", nrow(dhis_births), "\n")
cat("  dhis_deaths :", nrow(dhis_deaths), "\n")
cat("  dhis_contra :", nrow(dhis_contra), "\n")
```

## Indicator reference

The 27 DHIS2 indicators are grouped into three domains. This table is the single source of truth used to split the data.

```{r}
#| label: indicator-table
datatable(
  indicator_meta[, .(domain, short_label, indicator)][order(domain, short_label)],
  colnames    = c("Domain", "Short label", "Full DHIS2 indicator name"),
  rownames    = FALSE,
  options     = list(pageLength = 30, dom = "ftp"),
  caption     = "Table 1. Indicator domain map"
)
```

---

## Reporting completeness {#sec-completeness}

Before interpreting any trend, inspect how many facilities were submitting data in each month. Spikes or drops in counts indicate data quality issues rather than true events.

```{r}
#| label: fig-completeness
#| fig-cap: "Monthly reporting completeness — number of distinct facility–province–district combinations submitting data per month and domain (Apr 2024 – Mar 2026)."
#| fig-height: 5

# Three-panel: one per domain
comp_plot <- ggplot(agg_completeness_monthly,
          aes(x = period_date, y = n_facilities, colour = domain,
                        text = paste0("Domain: ", domain,
                                      "<br>Month: ", format(period_date, "%b %Y"),
                                      "<br>Facilities: ", format(n_facilities, big.mark = ",")))) +
  geom_line(show.legend = FALSE, linewidth = 0.8) +
  geom_point(show.legend = FALSE, size = 1.2) +
  facet_wrap(~domain, ncol = 3, scales = "free_y",
             labeller = labeller(domain = c(
               births = "Births / deliveries / BCG",
               deaths = "Deaths",
               contra = "Contraception / reproductive"
             ))) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %y") +
  scale_colour_manual(values = c(births = "#2196F3", deaths = "#f44336", contra = "#4CAF50")) +
  labs(x = NULL, y = "Facilities reporting") +
  theme_mrc +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplotly(comp_plot, tooltip = "text") |>
  layout(margin = list(b = 60)) |>
  ply_config()
```

---

## Births, deliveries & BCG {#sec-births}

### National monthly trends (primary)

```{r}
#| label: fig-births-monthly
#| fig-cap: "Monthly national counts for births/delivery/BCG indicators (Apr 2024 – Mar 2026), independent y-axes."
#| fig-height: 7

make_small_multiples(
  agg_national_monthly[domain == "births"],
  time_col  = "period_date",
  date_axis = TRUE,
  line_col  = "#1f6eb5"
)
```

### By province (monthly)

```{r}
#| label: fig-births-province-monthly
#| fig-cap: "Provincial monthly births/delivery counts. Pick an indicator from the dropdown and toggle Linear / Log on the y-axis."

make_province_explorer(
  agg_prov_monthly[domain == "births"],
  domain_label = "Births / deliveries",
  x_col = "period_date",
  date_axis = TRUE
)
```

### Annual context

```{r}
#| label: fig-births-national
#| fig-cap: "National annual counts for each births/delivery/BCG indicator, 2015–2025. Each panel has its OWN y-axis (free scale) so small-volume indicators (e.g. teen deliveries) are interpretable alongside high-volume ones (e.g. live births). Hover for exact counts. 2026 is excluded (incomplete year)."
#| fig-height: 7

make_small_multiples(
  agg_national_annual[domain == "births"],
  time_col = "year",
  line_col = "#1f6eb5"
)
```

---

## Deaths {#sec-deaths}

### Monthly trend (Apr 2024 – Mar 2026, primary)

Monthly data allow finer temporal resolution over the most recent period. Each indicator is shown on its own scale so seasonality and short-term shifts are visible regardless of magnitude.

```{r}
#| label: fig-deaths-monthly
#| fig-cap: "Monthly national counts for each death indicator (Apr 2024 – Mar 2026), independent y-axes. Hover for month-level values."
#| fig-height: 7

make_small_multiples(
  agg_national_monthly[domain == "deaths"],
  time_col  = "period_date",
  date_axis = TRUE,
  line_col  = "#c0392b"
)
```

### By province (monthly)

```{r}
#| label: fig-deaths-province-monthly
#| fig-cap: "Provincial monthly death counts. Pick an indicator and toggle Linear / Log to compare high- and low-volume provinces."

make_province_explorer(
  agg_prov_monthly[domain == "deaths"],
  domain_label = "Deaths",
  x_col = "period_date",
  date_axis = TRUE
)
```

### Annual context

```{r}
#| label: fig-deaths-national
#| fig-cap: "National annual counts for each death indicator, 2015–2025, on independent (free) y-axes so low-count indicators (maternal death, 7–28 day deaths) are interpretable alongside the much larger total. 2026 is excluded (incomplete year). Note: 'Inpatient deaths (total)' is a separately reported aggregate and is NOT the sum of the age-specific panels."
#| fig-height: 7

make_small_multiples(
  agg_national_annual[domain == "deaths"],
  time_col = "year",
  line_col = "#c0392b"
)
```

---

## Contraception & reproductive {#sec-contra}

::: {.callout-note collapse="false"}
## Methods: facility counts, couple-years of protection, and monthly QA

**Facility counts.** A facility is counted as "reporting" for the contraception
domain in a given period (year or month) if it has at least one non-missing
service record for **any** contraceptive/reproductive indicator in that
period. Facilities are identified by the combination of `province + district
+ facility name` — the source files carry no separate facility ID/code, so
name clashes across districts are not resolved beyond this key. The same
definition underlies the [reporting completeness](#sec-completeness) panel
above and the monthly reporting-rate analysis below.

**Couple-years of protection (CYP).** Raw commodity/procedure counts are not
directly comparable across methods (e.g. one implant insertion protects for
years; one oral-pill cycle protects for a month). CYP converts each service
into a common annual-protection unit using standard USAID/FP2030 conversion
factors:

```{r}
#| label: tbl-cyp-factors
#| tbl-cap: "CYP conversion factors applied to each contraceptive service/procedure."

datatable(
  cyp_factors[, .(short_label,
                  cyp_factor = round(cyp_factor, 3),
                  interpretation = fifelse(
                    short_label == "Medroxyprogesterone", "4 injections (3-monthly) = 1 CYP",
                    fifelse(short_label == "Norethisterone", "6 injections (2-monthly) = 1 CYP",
                    fifelse(short_label == "Oral pill", "15 cycles = 1 CYP",
                    fifelse(short_label == "IUCD", "1 insertion = 4.6 CYP (avg use-life)",
                    fifelse(short_label == "Sub-dermal implant", "1 insertion = 3.5 CYP (avg use-life)",
                    "1 procedure = 10 CYP"))))))],
  colnames = c("Method", "CYP factor", "Interpretation"),
  rownames = FALSE,
  options  = list(pageLength = 10, dom = "t")
)
```

Termination of pregnancy is **not** a contraceptive method and is excluded
from CYP. Sterilisation and TOP counts remain shown separately in raw-count
form throughout this section.

**Monthly data-quality checks.** Two checks are applied to the Apr 2024 – Mar
2026 monthly file: (1) a district-level **reporting rate**, comparing each
month's number of reporting facilities against that district's own
highest-observed facility count in the window (so district size doesn't bias
the comparison); and (2) a **within-facility outlier flag**, using a robust
(median/MAD-based) z-score computed separately for each facility × indicator
series, so a facility is only ever compared against its own history. Points
with $|z| > 3.5$ (the Iglewicz–Hoaglin threshold) are flagged as outliers.
:::

### Monthly trend (Apr 2024 – Mar 2026, primary)

```{r}
#| label: fig-contra-monthly
#| fig-cap: "Monthly national counts for each contraception indicator (Apr 2024 – Mar 2026), independent y-axes."
#| fig-height: 8

make_small_multiples(
  agg_national_monthly[domain == "contra"],
  time_col  = "period_date",
  date_axis = TRUE,
  line_col  = "#1e8449"
)
```

### By province (monthly)

```{r}
#| label: fig-contra-province-monthly
#| fig-cap: "Provincial monthly contraception counts. Pick an indicator and toggle Linear / Log on the y-axis."

make_province_explorer(
  agg_prov_monthly[domain == "contra"],
  domain_label = "Contraception",
  x_col = "period_date",
  date_axis = TRUE
)
```

### Annual context

```{r}
#| label: fig-contra-national
#| fig-cap: "National annual counts for each contraception / reproductive indicator, 2015–2025, on independent (free) y-axes so rarer procedures (e.g. male sterilisation, TOP) are interpretable alongside high-volume methods (oral pill, injectables). 2026 is excluded (incomplete year)."
#| fig-height: 8

make_small_multiples(
  agg_national_annual[domain == "contra"],
  time_col = "year",
  line_col = "#1e8449"
)
```

### Couple-years of protection (CYP)

National annual CYP, by method and combined. TOP and raw commodity counts are shown elsewhere in this section; this view expresses only methods with a defined CYP conversion factor (see Methods above).

```{r}
#| label: fig-contra-cyp-national
#| fig-cap: "National couple-years of protection (CYP) by method, 2015–2025. Each panel has its own y-axis."
#| fig-height: 6

cyp_plot_dt <- copy(agg_cyp_national_annual)
setnames(cyp_plot_dt, "cyp", "value")
make_small_multiples(cyp_plot_dt, time_col = "year", line_col = "#6a3d9a")
```

```{r}
#| label: fig-contra-cyp-total
#| fig-cap: "Total national CYP across all methods, 2015–2025."

cyp_total_plot <- ggplot(agg_cyp_national_annual_total, aes(x = year, y = cyp,
      text = paste0("Year: ", year, "<br>Total CYP: ", format(round(cyp), big.mark = ",")))) +
  geom_col(fill = "#6a3d9a") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(breaks = 2015:2025) +
  labs(x = NULL, y = "Couple-years of protection") +
  theme_mrc

ggplotly(cyp_total_plot, tooltip = "text") |> ply_config()
```

### Monthly reporting quality (Apr 2024 – Mar 2026)

Districts are ranked by their **mean monthly reporting rate** (facilities reporting that month ÷ that district's own peak facility count in the window). None fall below the conventional 70% "poor reporting" threshold in this dataset; the table and chart below show the 15 lowest-ranked districts for transparency.

```{r}
#| label: tbl-contra-poor-reporting
#| tbl-cap: "Districts with the lowest mean monthly reporting rate for contraception indicators, Apr 2024 – Mar 2026 (worst 15 shown)."

datatable(
  head(agg_contra_district_poor_reporting, 15)[
    , .(province, district,
        mean_reporting_rate = percent(mean_reporting_rate, accuracy = 1),
        min_reporting_rate  = percent(min_reporting_rate, accuracy = 1),
        n_months)],
  colnames = c("Province", "District", "Mean reporting rate", "Min reporting rate", "Months observed"),
  rownames = FALSE,
  options  = list(pageLength = 15, dom = "ftp")
)
```

```{r}
#| label: fig-contra-district-reporting
#| fig-cap: "Monthly reporting rate for the 8 lowest-ranked districts (facilities reporting ÷ district's peak facility count). The dashed line marks the conventional 70% threshold — no district drops below it in this dataset, but relative dips are still visible."
#| fig-height: 6

worst_districts <- head(agg_contra_district_poor_reporting, 8)$district

rep_plot_dt <- agg_contra_completeness_district_monthly[district %in% worst_districts]

rep_plot <- ggplot(rep_plot_dt, aes(x = period_date, y = reporting_rate,
      colour = district, group = interaction(province, district),
      text = paste0(district, " (", province, ")<br>",
                    format(period_date, "%b %Y"), "<br>",
                    "Reporting rate: ", percent(reporting_rate, accuracy = 1)))) +
  geom_hline(yintercept = 0.7, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 0.5, alpha = 0.8) +
  scale_y_continuous(labels = percent) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %y") +
  labs(x = NULL, y = "Reporting rate", colour = NULL) +
  theme_mrc +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1))

ggplotly(rep_plot, tooltip = "text") |> ply_config()
```

### Reporting heatmaps

Heatmaps make gaps and outlier months visible at a glance, without pre-selecting "worst" districts.

**All districts, national overview.** Rows are sorted by mean reporting rate (worst at the top). Green = fully reporting that month relative to the district's own peak; red = a large share of facilities went silent.

```{r}
#| label: fig-contra-heatmap-district
#| fig-cap: "District x month reporting-rate heatmap, Apr 2024 – Mar 2026. Reporting rate = facilities reporting that month ÷ district's own peak facility count in the window."
#| fig-height: 9

dist_mat_dt <- dcast(agg_contra_heatmap_district, district_label ~ period_date,
                     value.var = "reporting_rate")
dist_labels <- dist_mat_dt$district_label
dist_mat    <- as.matrix(dist_mat_dt[, -1, with = FALSE])
rownames(dist_mat) <- dist_labels
month_labs  <- format(as.Date(colnames(dist_mat)), "%b %y")

plot_ly(
  z = dist_mat, x = month_labs, y = dist_labels, type = "heatmap",
  colorscale = list(c(0, "#b2182b"), c(0.5, "#f7f7f7"), c(1, "#1a9850")),
  zmin = 0, zmax = 1,
  hovertemplate = paste0("%{y}<br>%{x}<br>Reporting rate: %{z:.0%}<extra></extra>"),
  colorbar = list(title = "Reporting<br>rate", tickformat = ".0%")
) |>
  layout(
    xaxis  = list(title = NULL, tickangle = -45),
    yaxis  = list(title = NULL, tickfont = list(size = 8)),
    margin = list(l = 160)
  ) |>
  ply_config()
```

**Facility level, one district at a time.** Colour = **percentage of that facility's own maximum monthly volume**, calculated as:

$$
\%\ of\ own\ max_{i,t} = \frac{\text{total contraceptive volume, facility } i,\ \text{month } t}{\max_{t}(\text{total contraceptive volume, facility } i)} \times 100
$$

100% therefore always marks a facility's own single best-reporting month in the window — it is **not** a fixed national or district target, so facilities are only ever judged against their own history, never against each other. A month sitting at, say, 20% means that facility reported only a fifth of the volume it is capable of reporting, which is an immediate, self-scaled signal of a completeness gap or genuine outlier — regardless of whether the facility is large or small. Blank cells are months with **no reported service at all** for that facility — the clearest signal of under-recording.

```{r}
#| label: fig-contra-heatmap-facility
#| fig-cap: "Facility x month heatmap for a selected district. Colour = monthly volume as a percentage of that facility's own maximum monthly volume in the window (100% = facility's own best month; 0% = reported nothing). Blank = facility reported nothing that month. Use the dropdown to switch district."
#| fig-height: 7

fac_grid <- agg_contra_facility_monthly_grid
fac_grid[, district_prov := paste0(district, " (", province, ")")]
fac_grid[, facility_label := facility]

districts_avail <- fac_grid[, .(n_fac = uniqueN(facility)), by = district_prov][
  order(district_prov)
]$district_prov
month_labs_fac <- format(all_months <- sort(unique(fac_grid$period_date)), "%b %y")

fig_fac <- plot_ly()
for (i in seq_along(districts_avail)) {
  d <- districts_avail[i]
  sub <- fac_grid[district_prov == d]
  wide <- dcast(sub, facility_label ~ period_date, value.var = "pct_of_max")
  fac_labels <- wide$facility_label
  mat <- as.matrix(wide[, -1, with = FALSE])
  # Align columns to the full month range (some districts may be missing an
  # early/late month if no facility in that district reported at all then)
  mat_full <- matrix(NA_real_, nrow = nrow(mat), ncol = length(all_months),
                     dimnames = list(fac_labels, as.character(all_months)))
  mat_full[, colnames(mat)] <- mat

  fig_fac <- add_trace(
    fig_fac,
    z = mat_full, x = month_labs_fac, y = fac_labels, type = "heatmap",
    visible = (i == 1),
    colorscale = list(c(0, "#b2182b"), c(0.5, "#f7f7f7"), c(1, "#1a9850")),
    zmin = 0, zmax = 100,
    hovertemplate = paste0("%{y}<br>%{x}<br>% of own max: %{z:.0f}%<extra></extra>"),
    colorbar = list(title = "% of own<br>max month", ticksuffix = "%")
  )
}

dist_buttons <- lapply(seq_along(districts_avail), function(i) {
  vis <- rep(FALSE, length(districts_avail))
  vis[i] <- TRUE
  list(method = "update", label = districts_avail[i],
       args = list(list(visible = vis),
                   list(title = paste0("Facility reporting: ", districts_avail[i]))))
})

fig_fac |>
  layout(
    title  = paste0("Facility reporting: ", districts_avail[1]),
    xaxis  = list(title = NULL, tickangle = -45),
    yaxis  = list(title = NULL, tickfont = list(size = 8)),
    margin = list(l = 140, t = 80),
    updatemenus = list(
      list(type = "dropdown", x = 0, xanchor = "left", y = 1.15,
           buttons = dist_buttons, showactive = TRUE)
    )
  ) |>
  ply_config()
```

### Monthly outlier flags

Facility-months where a contraceptive indicator deviates sharply (robust $|z| > 3.5$) from that same facility's own history — likely data-entry errors, stockouts, or reporting artefacts rather than true service-delivery shifts.

```{r}
#| label: tbl-contra-outliers
#| tbl-cap: "Flagged facility-month outliers for contraception indicators, Apr 2024 – Mar 2026 (largest deviations first)."

datatable(
  agg_contra_outliers_monthly[, .(province, district, facility, short_label,
                                   period = format(period_date, "%b %Y"),
                                   value, z_robust = round(z_robust, 1))],
  colnames  = c("Province", "District", "Facility", "Indicator", "Period", "Value", "Robust z"),
  rownames  = FALSE,
  filter    = "top",
  options   = list(pageLength = 15, dom = "ftp", scrollX = TRUE)
)
```

---

## Facility-level raw data browser {#sec-facility}

Search, sort and filter the full annual dataset at facility level.

```{r}
#| label: tbl-facility
#| tbl-cap: "Facility-level annual totals. Use the search box and column filters to explore specific provinces, districts or indicators."

datatable(
  agg_facility_totals[, .(domain, short_label, province, district, facility,
                           total_value, years_reported)],
  colnames  = c("Domain", "Indicator", "Province", "District", "Facility",
                "Total (all years)", "Years reporting"),
  rownames  = FALSE,
  filter    = "top",
  extensions = "Buttons",
  options   = list(
    pageLength = 20,
    dom        = "Bfrtip",
    buttons    = list("csv", "excel"),
    scrollX    = TRUE,
    columnDefs = list(list(className = "dt-right", targets = c(5, 6)))
  )
) |>
  formatCurrency("total_value", currency = "", interval = 3, mark = ",", digits = 0)
```

---

## Notes on methodology

### Data sources
- **Annual file**: `DHIS DATA 2015_2026 (2).xls` — six sheets covering 2015–2026 in wide format (one column per year). The 2022–2026 period is split across three sheets by indicator domain.
- **Monthly file**: `DHIS DATA APR 2024_MAR 2026_test.xls` — four sheets with monthly columns (e.g. "April 2024").

### Processing steps (see `wrangling.R`)
1. Each sheet read with `readxl::read_excel(..., col_types = "text")` to prevent type coercion errors.
2. Fixed columns 1–4 (Province, District, Sub-district, Facility) cleaned by stripping the two-letter DHIS2 province code prefix and trailing administrative suffixes.
3. Value columns melted to long format; values coerced to numeric after removing non-numeric characters.
4. Indicators classified into three domains (births, deaths, contra) via `indicator_meta` lookup table.
5. Monthly sheets de-duplicated on facility + indicator + period after binding.
6. **The 2026 calendar year is dropped from the annual series** (`year <= 2025`) because it is incomplete and would otherwise show a spurious partial-year drop. Monthly data, which is complete per-month through Mar 2026, is retained in the monthly figures.
7. No imputation of missing values; `NA` counts are excluded from aggregations.

### Limitations
- Annual counts in the main file are calendar-year aggregates; months within a year are not individually available before April 2024.
- The annual series ends in 2025; 2026 is excluded as incomplete. The monthly figures extend to Mar 2026.
- The `Inpatient deaths - total` indicator is a separately reported aggregate that may not equal the sum of age-specific death indicators — they are shown together but not summed.
- Facility coverage varies by year and month (see [Reporting completeness](#sec-completeness)). Declining counts are often incomplete reporting, not true declines.
- These data have not been validated against the official MACOD or StatsSA vital registration series.
