---
title: "DHIS2 Hospital Indicators — Exploratory Analysis"
description: "Trend exploration of NDoH DHIS2 hospital-level data (births, deaths, contraception) for South Africa, 2015–2025. Exploratory only — not for inference."
date: "2026-06-13"
categories: [DHIS2, hospitals, births, deaths, contraception, exploratory]
---

```{r}
#| label: setup
#| include: false

suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(plotly)
  library(scales)
  library(DT)
  library(htmltools)
  library(lubridate)
})

# ── Load wrangling output ─────────────────────────────────────────────────────
rda_path <- here::here("projects/dhis_hospital_mortality/dhis_results.rda")
if (!file.exists(rda_path)) {
  stop("Run projects/dhis_hospital_mortality/wrangling.R first.")
}
load(rda_path)

# Monthly provincial aggregate used to prioritize month-level province views.
if (!exists("agg_prov_monthly")) {
  agg_prov_monthly <- dhis_monthly[!is.na(province) & province != "",
    .(value = sum(value, na.rm = TRUE)),
    by = .(domain, indicator, short_label, province, year, month, period_date)
  ][order(domain, indicator, province, period_date)]
}

# ── Colour palette (consistent across all tabs) ───────────────────────────────
PROV_COLOURS <- c(
  "Eastern Cape"  = "#e41a1c",
  "Free State"    = "#ff7f00",
  "Gauteng"       = "#984ea3",
  "KwaZulu-Natal" = "#377eb8",
  "Limpopo"       = "#4daf4a",
  "Mpumalanga"    = "#a65628",
  "North West"    = "#f781bf",
  "Northern Cape" = "#999999",
  "Western Cape"  = "#1f78b4"
)

# Theme for static ggplots (also used inside ggplotly)
theme_mrc <- theme_bw(base_size = 12) +
  theme(
    panel.grid.minor  = element_blank(),
    strip.background  = element_rect(fill = "#e8edf2"),
    plot.title        = element_text(face = "bold", size = 13),
    plot.subtitle     = element_text(size = 10, colour = "grey40"),
    legend.position   = "bottom"
  )

# ── Shared plotly config: tidy toolbar, hi-res PNG export ──────────────────────
ply_config <- function(p) {
  config(
    p,
    displaylogo = FALSE,
    modeBarButtonsToRemove = c("lasso2d", "select2d", "autoScale2d",
                               "hoverClosestCartesian", "hoverCompareCartesian"),
    toImageButtonOptions = list(format = "png", scale = 2)
  )
}

# ── Small-multiples builder ───────────────────────────────────────────────────
# Faceted, free-y trend panels so EVERY indicator is interpretable at its own
# scale (large- and small-magnitude indicators no longer compete on one axis).
# `time_col` is either "year" (numeric) or "period_date" (Date).
make_small_multiples <- function(dt, time_col,
                                 date_axis = FALSE,
                                 line_col  = "#1f6eb5",
                                 ncol      = 3) {
  dt <- copy(dt)
  dt[, x_plot := get(time_col)]
  dt[, tt := paste0(
    short_label, "<br>",
    if (date_axis) format(x_plot, "%b %Y") else paste("Year:", x_plot),
    "<br>Count: ", format(value, big.mark = ",")
  )]

  g <- ggplot(dt, aes(x = x_plot, y = value, group = short_label, text = tt)) +
    geom_line(colour = line_col, linewidth = 0.6) +
    geom_point(colour = line_col, size = 1.1) +
    facet_wrap(~ short_label, scales = "free_y", ncol = ncol) +
    scale_y_continuous(labels = comma) +
    labs(x = NULL, y = "Count") +
    theme_mrc +
    theme(legend.position = "none",
          strip.text  = element_text(size = 8),
          axis.text.x = element_text(angle = 45, hjust = 1, size = 7))

  g <- if (date_axis) {
    g + scale_x_date(date_breaks = "6 months", date_labels = "%b %y")
  } else {
    g + scale_x_continuous(breaks = seq(2015, 2025, 2))
  }

  ggplotly(g, tooltip = "text") |> ply_config()
}

# ── Province explorer builder ─────────────────────────────────────────────────
# One indicator at a time (dropdown) across all provinces, with a Linear/Log
# y-axis toggle so small provinces remain interpretable against large ones.
# Works for annual or monthly series.
make_province_explorer <- function(dt, domain_label,
                                   x_col = "year",
                                   date_axis = FALSE) {
  dt        <- dt[!is.na(province)]
  dt[, x_plot := get(x_col)]
  dt[, hover_text := if (date_axis) {
    paste0("<b>", province, "</b><br>", short_label,
           "<br>", format(x_plot, "%b %Y"), ": ", format(value, big.mark = ","))
  } else {
    paste0("<b>", province, "</b><br>", short_label,
           "<br>", x_plot, ": ", format(value, big.mark = ","))
  }]
  indicators <- unique(dt$short_label)
  provs     <- sort(unique(dt$province))
  n_provs   <- length(provs)

  fig <- plot_ly()
  for (ind in indicators) {
    df_i <- dt[short_label == ind]
    for (pv in provs) {
      df_p <- df_i[province == pv]
      fig <- add_trace(
        fig,
        data        = df_p,
        x = ~x_plot, y = ~value,
        text        = ~hover_text,
        name        = pv,
        type = "scatter", mode = "lines+markers",
        visible     = (ind == indicators[1]),
        legendgroup = pv,
        showlegend  = (ind == indicators[1]),
        line   = list(color = unname(PROV_COLOURS[pv]), width = 2),
        marker = list(color = unname(PROV_COLOURS[pv]), size = 5),
        hovertemplate = "%{text}<extra></extra>"
      )
    }
  }

  # Indicator dropdown — toggles which block of province traces is visible
  ind_buttons <- lapply(seq_along(indicators), function(i) {
    vis <- rep(FALSE, length(indicators) * n_provs)
    vis[((i - 1) * n_provs + 1):(i * n_provs)] <- TRUE
    list(method = "update", label = indicators[i],
         args = list(list(visible = vis),
                     list(title = paste0(domain_label, ": ", indicators[i],
                                         " — by province"))))
  })

  # Linear / Log y-axis toggle
  scale_buttons <- list(
    list(method = "relayout", label = "Linear",
         args = list(list(`yaxis.type` = "linear",
                          `yaxis.title.text` = "Count"))),
    list(method = "relayout", label = "Log",
         args = list(list(`yaxis.type` = "log",
                          `yaxis.title.text` = "Count (log scale)")))
  )

  fig |>
    layout(
      title     = paste0(domain_label, ": ", indicators[1], " — by province"),
      hovermode = "x unified",
      xaxis     = if (date_axis) {
        list(title = "", tickformat = "%b %y", showspikes = TRUE,
             spikemode = "across", spikethickness = 1)
      } else {
        list(title = "", dtick = 1, showspikes = TRUE,
             spikemode = "across", spikethickness = 1)
      },
      yaxis     = list(title = "Count", tickformat = ","),
      legend    = list(orientation = "h", y = -0.22),
      margin    = list(t = 90),
      updatemenus = list(
        list(type = "dropdown", x = 0.0, xanchor = "left", y = 1.22,
             buttons = ind_buttons, showactive = TRUE,
             pad = list(t = 2, b = 2)),
        list(type = "buttons", direction = "right",
             x = 0.62, xanchor = "left", y = 1.22,
             buttons = scale_buttons, showactive = TRUE,
             pad = list(t = 2, b = 2))
      ),
      annotations = list(
        list(text = "Indicator:", x = 0.0, xref = "paper",
             y = 1.30, yref = "paper", xanchor = "left",
             showarrow = FALSE, font = list(size = 11, color = "grey40")),
        list(text = "y-axis:", x = 0.62, xref = "paper",
             y = 1.30, yref = "paper", xanchor = "left",
             showarrow = FALSE, font = list(size = 11, color = "grey40"))
      )
    ) |>
    ply_config()
}
```

::: {.callout-note}
**Exploratory only.** This page presents DHIS2 hospital data supplied by Pam Groenewald (NDoH). The data have not been quality-audited. Trends shown are descriptive; no causal or inferential claims should be drawn. For comparison with official MACOD data, see the dedicated comparison section (planned).
:::

## Data provenance & audit trail

```{r}
#| label: audit-table
#| echo: true

# Print the audit record so the page self-documents its inputs
cat("Source files:\n")
cat("  ", audit$source_annual,  "\n")
cat("  ", audit$source_monthly, "\n\n")
cat("Annual sheets read    :", paste(audit$annual_sheets, collapse = " | "), "\n")
cat("Monthly sheets read   :", paste(audit$monthly_sheets, collapse = " | "), "\n")
cat("Annual year range     :", audit$year_range_annual[1], "–", audit$year_range_annual[2], "\n")
cat("Monthly period range  :", format(audit$period_range_monthly[1]),
    "–", format(audit$period_range_monthly[2]), "\n")
cat("Wrangling run at      :", format(audit$run_timestamp), "\n\n")
cat("Dataset sizes (long rows):\n")
cat("  dhis_births :", nrow(dhis_births), "\n")
cat("  dhis_deaths :", nrow(dhis_deaths), "\n")
cat("  dhis_contra :", nrow(dhis_contra), "\n")
```

## Indicator reference

The 27 DHIS2 indicators are grouped into three domains. This table is the single source of truth used to split the data.

```{r}
#| label: indicator-table
datatable(
  indicator_meta[, .(domain, short_label, indicator)][order(domain, short_label)],
  colnames    = c("Domain", "Short label", "Full DHIS2 indicator name"),
  rownames    = FALSE,
  options     = list(pageLength = 30, dom = "ftp"),
  caption     = "Table 1. Indicator domain map"
)
```

---

## Reporting completeness {#sec-completeness}

Before interpreting any trend, inspect how many facilities were submitting data in each month. Spikes or drops in counts indicate data quality issues rather than true events.

```{r}
#| label: fig-completeness
#| fig-cap: "Figure 1. Monthly reporting completeness: the solid line connects observed monthly counts and the points mark each observed monthly count of distinct facility–province–district combinations submitting data for each domain (Apr 2024 – Mar 2026)."
#| fig-height: 5

fig_completeness <- plot_ly()
for (d in c("births", "deaths", "contra")) {
  dt <- agg_completeness_monthly[domain == d]
  fig_completeness <- fig_completeness |>
    add_trace(
      data = dt,
      x = ~period_date,
      y = ~n_facilities,
      type = "scatter",
      mode = "lines+markers",
      name = switch(d,
        births = "Births / deliveries / BCG",
        deaths = "Deaths",
        contra = "Contraception / reproductive"
      ),
      line = list(color = switch(d,
        births = "#2196F3",
        deaths = "#f44336",
        contra = "#4CAF50"
      ), width = 2),
      marker = list(color = switch(d,
        births = "#2196F3",
        deaths = "#f44336",
        contra = "#4CAF50"
      ), size = 6),
      hovertemplate = paste0(
        "<b>%{fullData.name}</b><br>",
        "Month: %{x|%b %Y}<br>",
        "Facilities reporting: %{y:,}<extra></extra>"
      )
    )
}

fig_completeness <- fig_completeness |>
  layout(
    xaxis = list(title = "", type = "date", tickformat = "%b %y"),
    yaxis = list(title = "Facilities reporting"),
    margin = list(b = 60),
    legend = list(orientation = "h", x = 0, y = 1.12)
  ) |>
  ply_config()

fig_completeness
```

This first figure reports the monthly number of distinct facility–province–district combinations that submitted at least one valid observation for each domain. The solid line traces the pattern through time and the points show the observed monthly counts themselves; a fall in either value suggests lower reporting completeness rather than necessarily a true change in service volume.

---

## Births, deliveries & BCG {#sec-births}

### National monthly trends (primary)

```{r}
#| label: fig-births-monthly
#| fig-cap: "Monthly national counts for births/delivery/BCG indicators (Apr 2024 – Mar 2026), independent y-axes."
#| fig-height: 7

make_small_multiples(
  agg_national_monthly[domain == "births"],
  time_col  = "period_date",
  date_axis = TRUE,
  line_col  = "#1f6eb5"
)
```

### By province (monthly)

```{r}
#| label: fig-births-province-monthly
#| fig-cap: "Provincial monthly births/delivery counts. Pick an indicator from the dropdown and toggle Linear / Log on the y-axis."

make_province_explorer(
  agg_prov_monthly[domain == "births"],
  domain_label = "Births / deliveries",
  x_col = "period_date",
  date_axis = TRUE
)
```

### Annual context

```{r}
#| label: fig-births-national
#| fig-cap: "National annual counts for each births/delivery/BCG indicator, 2015–2025. Each panel has its OWN y-axis (free scale) so small-volume indicators (e.g. teen deliveries) are interpretable alongside high-volume ones (e.g. live births). Hover for exact counts. 2026 is excluded (incomplete year)."
#| fig-height: 7

make_small_multiples(
  agg_national_annual[domain == "births"],
  time_col = "year",
  line_col = "#1f6eb5"
)
```

---

## Deaths {#sec-deaths}

### Monthly trend (Apr 2024 – Mar 2026, primary)

Monthly data allow finer temporal resolution over the most recent period. Each indicator is shown on its own scale so seasonality and short-term shifts are visible regardless of magnitude.

```{r}
#| label: fig-deaths-monthly
#| fig-cap: "Monthly national counts for each death indicator (Apr 2024 – Mar 2026), independent y-axes. Hover for month-level values."
#| fig-height: 7

make_small_multiples(
  agg_national_monthly[domain == "deaths"],
  time_col  = "period_date",
  date_axis = TRUE,
  line_col  = "#c0392b"
)
```

### By province (monthly)

```{r}
#| label: fig-deaths-province-monthly
#| fig-cap: "Provincial monthly death counts. Pick an indicator and toggle Linear / Log to compare high- and low-volume provinces."

make_province_explorer(
  agg_prov_monthly[domain == "deaths"],
  domain_label = "Deaths",
  x_col = "period_date",
  date_axis = TRUE
)
```

### Annual context

```{r}
#| label: fig-deaths-national
#| fig-cap: "National annual counts for each death indicator, 2015–2025, on independent (free) y-axes so low-count indicators (maternal death, 7–28 day deaths) are interpretable alongside the much larger total. 2026 is excluded (incomplete year). Note: 'Inpatient deaths (total)' is a separately reported aggregate and is NOT the sum of the age-specific panels."
#| fig-height: 7

make_small_multiples(
  agg_national_annual[domain == "deaths"],
  time_col = "year",
  line_col = "#c0392b"
)
```

---

## Contraception & reproductive {#sec-contra}

::: {.callout-note collapse="false"}
## Methods: facility counts, couple-years of protection, and monthly QA

**Facility counts.** A facility is counted as "reporting" for the contraception
domain in a given period (year or month) if it has at least one non-missing
service record for **any** contraceptive/reproductive indicator in that
period. Facilities are identified by the combination of `province + district
+ facility name` — the source files carry no separate facility ID/code, so
name clashes across districts are not resolved beyond this key. The same
definition underlies the [reporting completeness](#sec-completeness) panel
above and the monthly reporting-rate analysis below.

**Couple-years of protection (CYP).** Raw commodity/procedure counts are not
directly comparable across methods (e.g. one implant insertion protects for
years; one oral-pill cycle protects for a month). CYP converts each service
into a common annual-protection unit using standard USAID/FP2030 conversion
factors:

```{r}
#| label: tbl-cyp-factors
#| tbl-cap: "CYP conversion factors applied to each contraceptive service/procedure."

datatable(
  cyp_factors[, .(short_label,
                  cyp_factor = round(cyp_factor, 3),
                  interpretation = fifelse(
                    short_label == "Medroxyprogesterone", "4 injections (3-monthly) = 1 CYP",
                    fifelse(short_label == "Norethisterone", "6 injections (2-monthly) = 1 CYP",
                    fifelse(short_label == "Oral pill", "15 cycles = 1 CYP",
                    fifelse(short_label == "IUCD", "1 insertion = 4.6 CYP (avg use-life)",
                    fifelse(short_label == "Sub-dermal implant", "1 insertion = 3.5 CYP (avg use-life)",
                    fifelse(short_label %in% c("Male condom", "Female condom"),
                            "120 units distributed = 1 CYP (standard factor; not method-specific)",
                    "1 procedure = 10 CYP")))))))],
  colnames = c("Method", "CYP factor", "Interpretation"),
  rownames = FALSE,
  options  = list(pageLength = 10, dom = "t")
)
```

Termination of pregnancy is **not** a contraceptive method and is excluded
from CYP. Sterilisation and TOP counts remain shown separately in raw-count
form throughout this section.

**Condoms — separate source file, mixed reporting grain.** Male/female condom
distribution comes from a second DHIS2 export (`ANC_DHIS DATA 2015_2026 2.xls`)
with a data-quality issue of its own: reporting is at **facility level**
from 2015-2017, then collapses to a **single row per district/municipality**
from 2018 onwards (named `"<Municipality> Primary Condom Distribution
Site"` — not a real facility). Facility-level totals are therefore **not**
comparable across the 2017/2018 boundary, so condom volumes are summed to
province + district + year for the whole series (this collapse is safe —
aggregating real facilities to their district gives the same district total
as the later single row) and are excluded from the facility-count
completeness, outlier and heatmap checks elsewhere in this section, which
assume a stable facility identity over time. The standard USAID/FP2030 CYP
factor for condoms (120 units = 1 CYP) is applied to male and female counts
separately; no official method-specific factor exists for female condoms, so
this is a working assumption rather than a verified national standard.
Condom counts also measure *distribution*, not confirmed use — see the
[condom CYP sensitivity analysis](condom_sensitivity.qmd) for how assuming
lower real-world utilization (or excluding condoms entirely) affects total
CYP.

**Monthly data-quality checks.** Two checks are applied to the Apr 2024 – Mar
2026 monthly file: (1) a district-level **reporting rate**, comparing each
month's number of reporting facilities against that district's own
highest-observed facility count in the window (so district size doesn't bias
the comparison); and (2) a **within-facility outlier flag**, using a robust
(median/MAD-based) z-score computed separately for each facility × indicator
series, so a facility is only ever compared against its own history. Points
with $|z| > 3.5$ (the Iglewicz–Hoaglin threshold) are flagged as outliers.
:::

### Monthly trend (Apr 2024 – Mar 2026, primary)

```{r}
#| label: fig-contra-monthly
#| fig-cap: "Monthly national counts for each contraception indicator (Apr 2024 – Mar 2026), independent y-axes."
#| fig-height: 8

make_small_multiples(
  agg_national_monthly[domain == "contra"],
  time_col  = "period_date",
  date_axis = TRUE,
  line_col  = "#1e8449"
)
```

### By province (monthly)

```{r}
#| label: fig-contra-province-monthly
#| fig-cap: "Provincial monthly contraception counts. Pick an indicator and toggle Linear / Log on the y-axis."

make_province_explorer(
  agg_prov_monthly[domain == "contra"],
  domain_label = "Contraception",
  x_col = "period_date",
  date_axis = TRUE
)
```

### Annual context

```{r}
#| label: fig-contra-national
#| fig-cap: "National annual counts for each contraception / reproductive indicator, 2015–2025, on independent (free) y-axes so rarer procedures (e.g. male sterilisation, TOP) are interpretable alongside high-volume methods (oral pill, injectables). 2026 is excluded (incomplete year)."
#| fig-height: 8

make_small_multiples(
  agg_national_annual[domain == "contra"],
  time_col = "year",
  line_col = "#1e8449"
)
```

### Couple-years of protection (CYP)

National annual CYP, by method and combined. TOP and raw commodity counts are shown elsewhere in this section; this view expresses only methods with a defined CYP conversion factor (see Methods above).

```{r}
#| label: fig-contra-cyp-national
#| fig-cap: "National couple-years of protection (CYP) by method, 2015–2025. Each panel has its own y-axis."
#| fig-height: 6

cyp_plot_dt <- copy(agg_cyp_national_annual)
setnames(cyp_plot_dt, "cyp", "value")
make_small_multiples(cyp_plot_dt, time_col = "year", line_col = "#6a3d9a")
```

```{r}
#| label: fig-contra-cyp-total
#| fig-cap: "Total national CYP across all methods, 2015–2025."

cyp_total_plot <- ggplot(agg_cyp_national_annual_total, aes(x = year, y = cyp,
      text = paste0("Year: ", year, "<br>Total CYP: ", format(round(cyp), big.mark = ",")))) +
  geom_col(fill = "#6a3d9a") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(breaks = 2015:2025) +
  labs(x = NULL, y = "Couple-years of protection") +
  theme_mrc

ggplotly(cyp_total_plot, tooltip = "text") |> ply_config()
```

**Method mix.** The chart above shows the total volume of protection delivered, but not which methods drive it. The stacked view below shows each method's **share of total national CYP** per year — useful for spotting a shift towards/away from long-acting methods (implant, IUCD) versus short-acting ones (injectables, pill).

```{r}
#| label: fig-contra-cyp-share
#| fig-cap: "Share of total national CYP contributed by each contraceptive method, 2015–2025. Each bar sums to 100%."
#| fig-height: 6

CYP_METHOD_COLOURS <- c(
  "Medroxyprogesterone" = "#1b9e77", "Norethisterone"     = "#66c2a5",
  "Oral pill"           = "#a6d854", "IUCD"                = "#377eb8",
  "Sub-dermal implant"  = "#984ea3", "Sterilisation (F)"   = "#e6ab02",
  "Sterilisation (M)"   = "#d95f02", "Male condom"         = "#4d4d4d",
  "Female condom"       = "#bdbdbd"
)

cyp_share_dt <- copy(agg_cyp_national_annual)
cyp_share_dt[, share := cyp / sum(cyp), by = year]
cyp_share_dt[, short_label := factor(short_label, levels = names(CYP_METHOD_COLOURS))]

cyp_share_plot <- ggplot(cyp_share_dt, aes(x = year, y = share, fill = short_label,
      text = paste0(short_label, "<br>Year: ", year,
                    "<br>Share of total CYP: ", percent(share, accuracy = 0.1),
                    "<br>CYP: ", format(round(cyp), big.mark = ",")))) +
  geom_col(position = "fill", width = 0.8) +
  scale_y_continuous(labels = percent, expand = c(0, 0)) +
  scale_x_continuous(breaks = 2015:2025) +
  scale_fill_manual(values = CYP_METHOD_COLOURS, name = "Method") +
  labs(x = NULL, y = "Share of total CYP") +
  theme_mrc

ggplotly(cyp_share_plot, tooltip = "text") |>
  layout(legend = list(orientation = "h", y = -0.25)) |>
  ply_config()
```

```{r}
#| label: tbl-contra-cyp-share-latest
#| tbl-cap: "Method mix (share of total national CYP) for the most recent annual data year."

latest_year <- max(cyp_share_dt$year)
datatable(
  cyp_share_dt[year == latest_year][order(-share),
    .(short_label, cyp = round(cyp), share = percent(share, accuracy = 0.1))],
  colnames = c("Method", paste0("CYP (", latest_year, ")"), "Share of total CYP"),
  rownames = FALSE,
  options  = list(pageLength = 10, dom = "t")
)
```

**Full underlying data.** Raw DHIS counts (post-cleaning, national annual) alongside the CYP factor applied and the resulting CYP, for every method and year — exportable for colleagues to build on directly.

```{r}
#| label: tbl-cyp-raw-full
#| tbl-cap: "Raw DHIS national annual counts (post-cleaning) and derived CYP, by contraceptive method and year, 2015–2025."

raw_contra_dt <- agg_national_annual[
  domain == "contra" & short_label %in% cyp_factors$short_label,
  .(short_label, year, raw_value = value)
]
raw_condom_dt <- agg_condom_national_annual[, .(short_label, year, raw_value = value)]
raw_all_dt    <- rbindlist(list(raw_contra_dt, raw_condom_dt))

cyp_raw_full <- merge(raw_all_dt, agg_cyp_national_annual, by = c("short_label", "year"), all = TRUE)
cyp_raw_full <- merge(cyp_raw_full, cyp_factors, by = "short_label", all.x = TRUE)
setorder(cyp_raw_full, short_label, year)

datatable(
  cyp_raw_full[, .(short_label, year,
                   raw_value  = round(raw_value),
                   cyp_factor = round(cyp_factor, 4),
                   cyp        = round(cyp))],
  colnames   = c("Method", "Year", "Raw DHIS count", "CYP factor", "CYP"),
  rownames   = FALSE,
  filter     = "top",
  extensions = "Buttons",
  options    = list(pageLength = 15, dom = "Bfrtip", buttons = list("csv", "excel"), scrollX = TRUE)
) |>
  formatCurrency(c("raw_value", "cyp"), currency = "", interval = 3, mark = ",", digits = 0)
```

**Condom sensitivity check.** The table below shows the same national annual CYP total under alternative assumptions about real-world condom use, while holding the 120-units-per-CYP conversion factor fixed. This is a robustness check for colleagues; the default report keeps the standard 100% use assumption.

```{r}
#| label: tbl-condom-sensitivity-summary
#| tbl-cap: "National annual CYP under alternative condom-utilization assumptions, most recent year shown."

latest_sens_year <- max(agg_condom_sensitivity_annual$year)
sens_latest <- agg_condom_sensitivity_annual[year == latest_sens_year]
standard_total <- sens_latest[scenario == "Standard (100% used, current default)", total_cyp]
sens_latest[, pct_vs_standard := (total_cyp - standard_total) / standard_total]

datatable(
  sens_latest[order(scenario),
    .(scenario,
      condom_cyp = round(condom_cyp),
      total_cyp  = round(total_cyp),
      pct_vs_standard = percent(pct_vs_standard, accuracy = 0.1))],
  colnames = c("Scenario", "Condom CYP", "Total CYP", "% vs Standard"),
  rownames = FALSE,
  extensions = "Buttons",
  options = list(pageLength = 10, dom = "Bfrtip", buttons = list("csv", "excel"), scrollX = TRUE)
) |>
  formatCurrency(c("condom_cyp", "total_cyp"), currency = "", interval = 3, mark = ",", digits = 0)
```

### Monthly reporting quality (Apr 2024 – Mar 2026)

Districts are ranked by their **mean monthly reporting rate** (facilities reporting that month ÷ that district's own peak facility count in the window). None fall below the conventional 70% "poor reporting" threshold in this dataset; the table and chart below show the 15 lowest-ranked districts for transparency.

```{r}
#| label: tbl-contra-poor-reporting
#| tbl-cap: "Districts with the lowest mean monthly reporting rate for contraception indicators, Apr 2024 – Mar 2026 (worst 15 shown)."

datatable(
  head(agg_contra_district_poor_reporting, 15)[
    , .(province, district,
        mean_reporting_rate = percent(mean_reporting_rate, accuracy = 1),
        min_reporting_rate  = percent(min_reporting_rate, accuracy = 1),
        n_months)],
  colnames = c("Province", "District", "Mean reporting rate", "Min reporting rate", "Months observed"),
  rownames = FALSE,
  options  = list(pageLength = 15, dom = "ftp")
)
```

```{r}
#| label: fig-contra-district-reporting
#| fig-cap: "Monthly reporting rate for the 8 lowest-ranked districts (facilities reporting ÷ district's peak facility count). The dashed line marks the conventional 70% threshold — no district drops below it in this dataset, but relative dips are still visible."
#| fig-height: 6

worst_districts <- head(agg_contra_district_poor_reporting, 8)$district

rep_plot_dt <- agg_contra_completeness_district_monthly[district %in% worst_districts]

rep_plot <- ggplot(rep_plot_dt, aes(x = period_date, y = reporting_rate,
      colour = district, group = interaction(province, district),
      text = paste0(district, " (", province, ")<br>",
                    format(period_date, "%b %Y"), "<br>",
                    "Reporting rate: ", percent(reporting_rate, accuracy = 1)))) +
  geom_hline(yintercept = 0.7, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 0.5, alpha = 0.8) +
  scale_y_continuous(labels = percent) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %y") +
  labs(x = NULL, y = "Reporting rate", colour = NULL) +
  theme_mrc +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1))

ggplotly(rep_plot, tooltip = "text") |> ply_config()
```

### Reporting heatmaps

Heatmaps make gaps and outlier months visible at a glance, without pre-selecting "worst" districts.

**All districts, national overview.** Rows are sorted by mean reporting rate (worst at the top). Green = fully reporting that month relative to the district's own peak; red = a large share of facilities went silent.

```{r}
#| label: fig-contra-heatmap-district
#| fig-cap: "District x month reporting-rate heatmap, Apr 2024 – Mar 2026. Reporting rate = facilities reporting that month ÷ district's own peak facility count in the window."
#| fig-height: 9

dist_mat_dt <- dcast(agg_contra_heatmap_district, district_label ~ period_date,
                     value.var = "reporting_rate")
dist_labels <- dist_mat_dt$district_label
dist_mat    <- as.matrix(dist_mat_dt[, -1, with = FALSE])
rownames(dist_mat) <- dist_labels
month_labs  <- format(as.Date(colnames(dist_mat)), "%b %y")

plot_ly(
  z = dist_mat, x = month_labs, y = dist_labels, type = "heatmap",
  colorscale = list(c(0, "#b2182b"), c(0.5, "#f7f7f7"), c(1, "#1a9850")),
  zmin = 0, zmax = 1,
  hovertemplate = paste0("%{y}<br>%{x}<br>Reporting rate: %{z:.0%}<extra></extra>"),
  colorbar = list(title = "Reporting<br>rate", tickformat = ".0%")
) |>
  layout(
    xaxis  = list(title = NULL, tickangle = -45),
    yaxis  = list(title = NULL, tickfont = list(size = 8)),
    margin = list(l = 160)
  ) |>
  ply_config()
```

**Facility level, one district at a time.** Colour shows each facility's own robust z-score for total monthly contraceptive volume (so facilities are compared against their own history, not against each other). Blank cells are months with **no reported service at all** for that facility — the clearest signal of under-recording.

```{r}
#| label: fig-contra-heatmap-facility
#| fig-cap: "Facility x month heatmap for a selected district. Colour = robust z-score of total monthly contraceptive volume (own-facility baseline); blank = facility reported nothing that month. Use the dropdown to switch district."
#| fig-height: 7

fac_grid <- agg_contra_facility_monthly_grid
fac_grid[, district_prov := paste0(district, " (", province, ")")]
fac_grid[, facility_label := facility]

districts_avail <- fac_grid[, .(n_fac = uniqueN(facility)), by = district_prov][
  order(district_prov)
]$district_prov
month_labs_fac <- format(all_months <- sort(unique(fac_grid$period_date)), "%b %y")

fig_fac <- plot_ly()
for (i in seq_along(districts_avail)) {
  d <- districts_avail[i]
  sub <- fac_grid[district_prov == d]
  wide <- dcast(sub, facility_label ~ period_date, value.var = "z_robust")
  fac_labels <- wide$facility_label
  mat <- as.matrix(wide[, -1, with = FALSE])
  # Align columns to the full month range (some districts may be missing an
  # early/late month if no facility in that district reported at all then)
  mat_full <- matrix(NA_real_, nrow = nrow(mat), ncol = length(all_months),
                     dimnames = list(fac_labels, as.character(all_months)))
  mat_full[, colnames(mat)] <- mat

  fig_fac <- add_trace(
    fig_fac,
    z = mat_full, x = month_labs_fac, y = fac_labels, type = "heatmap",
    visible = (i == 1),
    colorscale = "RdBu", reversescale = TRUE, zmid = 0,
    zmin = -4, zmax = 4,
    hovertemplate = paste0("%{y}<br>%{x}<br>Robust z: %{z:.1f}<extra></extra>"),
    colorbar = list(title = "Robust z")
  )
}

dist_buttons <- lapply(seq_along(districts_avail), function(i) {
  vis <- rep(FALSE, length(districts_avail))
  vis[i] <- TRUE
  list(method = "update", label = districts_avail[i],
       args = list(list(visible = vis),
                   list(title = paste0("Facility reporting: ", districts_avail[i]))))
})

fig_fac |>
  layout(
    title  = paste0("Facility reporting: ", districts_avail[1]),
    xaxis  = list(title = NULL, tickangle = -45),
    yaxis  = list(title = NULL, tickfont = list(size = 8)),
    margin = list(l = 140, t = 80),
    updatemenus = list(
      list(type = "dropdown", x = 0, xanchor = "left", y = 1.15,
           buttons = dist_buttons, showactive = TRUE)
    )
  ) |>
  ply_config()
```

### CYP data-quality deep dive

The outlier checks above pool **every** contraception indicator (including TOP, which is not a contraceptive method) into one facility-level total, so a spike or dropout in a single method can be diluted by the other methods reported at the same facility. This deep dive restricts to only the **facility-based methods that actually feed the CYP calculation** (Medroxyprogesterone, Norethisterone, Oral pill, IUCD, Sub-dermal implant, Sterilisation F/M — condoms are excluded here as they have no monthly facility-level data; see the CYP methods note above) and keeps the **method dimension separate**, so an outlier in one method is never masked by the others.

**Which methods are noisiest to report?** Injectables (Norethisterone, Medroxyprogesterone) and the oral pill have the highest facility-month outlier rates — consistent with their high service frequency (monthly/bi-monthly re-supply means more opportunities for a data-entry slip) — while the two sterilisation procedures, which are rare, single-event procedures per facility, are the most stable.

```{r}
#| label: fig-cyp-outlier-rate-method
#| fig-cap: "Share of facility-months flagged as outliers (robust |z| > 3.5), by CYP-eligible method."

rate_plot_dt <- copy(agg_cyp_outlier_rate_method)
rate_plot_dt[, short_label := factor(short_label, levels = rev(short_label))]

rate_plot <- ggplot(rate_plot_dt, aes(x = short_label, y = outlier_rate,
      text = paste0(short_label, "<br>Outlier rate: ", percent(outlier_rate, accuracy = 0.1),
                    "<br>", n_outliers, " of ", format(n_facility_months, big.mark = ","),
                    " facility-months"))) +
  geom_col(fill = "#6a3d9a") +
  coord_flip() +
  scale_y_continuous(labels = percent) +
  labs(x = NULL, y = "Facility-months flagged as outliers") +
  theme_mrc

ggplotly(rate_plot, tooltip = "text") |> ply_config()
```

**Facility x month x method outlier table.** Every flagged facility-month for a CYP-eligible method, largest deviation first — the same underlying flags as @tbl-contra-outliers, filtered to CYP methods only. The single largest deviation (Sub-dermal implant, one facility, one month) is two orders of magnitude above that facility's own typical volume — almost certainly a data-entry error (e.g. a unit or decimal-place slip) rather than a genuine service surge, and a candidate for exclusion or correction before using that facility-month in any CYP total.

```{r}
#| label: tbl-cyp-outliers
#| tbl-cap: "Flagged facility-month outliers for CYP-eligible contraceptive methods only, Apr 2024 – Mar 2026 (largest deviations first)."

datatable(
  agg_cyp_outliers_monthly[, .(province, district, facility, short_label,
                                period = format(period_date, "%b %Y"),
                                value, z_robust = round(z_robust, 1))],
  colnames  = c("Province", "District", "Facility", "Method", "Period", "Value", "Robust z"),
  rownames  = FALSE,
  filter    = "top",
  options   = list(pageLength = 15, dom = "ftp", scrollX = TRUE)
)
```

**Facility x month heatmap, one method at a time.** Scoped to `r cyp_worst_district$district` (`r cyp_worst_district$province`) — the single district with the most flagged CYP outlier facility-months (`r cyp_worst_district$N`) — so the picture stays legible; the table above covers all districts. Colour is each facility's own robust z-score for that method (own-facility baseline); blank cells are months with no reported service for that method at that facility.

```{r}
#| label: fig-cyp-facility-method-heatmap
#| fig-cap: "Facility x month heatmap by contraceptive method, for the district with the most flagged CYP outliers. Colour = robust z-score (own-facility, own-method baseline); blank = no service reported that month. Use the dropdown to switch method."
#| fig-height: 7

cyp_grid <- agg_cyp_facility_method_monthly_grid
cyp_months <- sort(unique(cyp_grid$period_date))
cyp_month_labs <- format(cyp_months, "%b %y")

fig_cyp <- plot_ly()
for (i in seq_along(CYP_ELIGIBLE_METHODS)) {
  m <- CYP_ELIGIBLE_METHODS[i]
  sub <- cyp_grid[short_label == m]
  wide <- dcast(sub, facility ~ period_date, value.var = "z_robust")
  fac_labels_cyp <- wide$facility
  mat <- as.matrix(wide[, -1, with = FALSE])
  mat_full <- matrix(NA_real_, nrow = nrow(mat), ncol = length(cyp_months),
                     dimnames = list(fac_labels_cyp, as.character(cyp_months)))
  mat_full[, colnames(mat)] <- mat

  fig_cyp <- add_trace(
    fig_cyp,
    z = mat_full, x = cyp_month_labs, y = fac_labels_cyp, type = "heatmap",
    visible = (i == 1),
    colorscale = "RdBu", reversescale = TRUE, zmid = 0,
    zmin = -4, zmax = 4,
    hovertemplate = paste0("%{y}<br>%{x}<br>Robust z: %{z:.1f}<extra></extra>"),
    colorbar = list(title = "Robust z")
  )
}

method_buttons <- lapply(seq_along(CYP_ELIGIBLE_METHODS), function(i) {
  vis <- rep(FALSE, length(CYP_ELIGIBLE_METHODS))
  vis[i] <- TRUE
  list(method = "update", label = CYP_ELIGIBLE_METHODS[i],
       args = list(list(visible = vis),
                   list(title = paste0(CYP_ELIGIBLE_METHODS[i], " — ",
                                        cyp_worst_district$district))))
})

fig_cyp |>
  layout(
    title  = paste0(CYP_ELIGIBLE_METHODS[1], " — ", cyp_worst_district$district),
    xaxis  = list(title = NULL, tickangle = -45),
    yaxis  = list(title = NULL, tickfont = list(size = 8)),
    margin = list(l = 160, t = 80),
    updatemenus = list(
      list(type = "dropdown", x = 0, xanchor = "left", y = 1.15,
           buttons = method_buttons, showactive = TRUE)
    )
  ) |>
  ply_config()
```

### Monthly outlier flags

Facility-months where a contraceptive indicator deviates sharply (robust $|z| > 3.5$) from that same facility's own history — likely data-entry errors, stockouts, or reporting artefacts rather than true service-delivery shifts.

```{r}
#| label: tbl-contra-outliers
#| tbl-cap: "Flagged facility-month outliers for contraception indicators, Apr 2024 – Mar 2026 (largest deviations first)."

datatable(
  agg_contra_outliers_monthly[, .(province, district, facility, short_label,
                                   period = format(period_date, "%b %Y"),
                                   value, z_robust = round(z_robust, 1))],
  colnames  = c("Province", "District", "Facility", "Indicator", "Period", "Value", "Robust z"),
  rownames  = FALSE,
  filter    = "top",
  options   = list(pageLength = 15, dom = "ftp", scrollX = TRUE)
)
```

---

## Facility-level raw data browser {#sec-facility}

Search, sort and filter the full annual dataset at facility level.

```{r}
#| label: tbl-facility
#| tbl-cap: "Facility-level annual totals. Use the search box and column filters to explore specific provinces, districts or indicators."

datatable(
  agg_facility_totals[, .(domain, short_label, province, district, facility,
                           total_value, years_reported)],
  colnames  = c("Domain", "Indicator", "Province", "District", "Facility",
                "Total (all years)", "Years reporting"),
  rownames  = FALSE,
  filter    = "top",
  extensions = "Buttons",
  options   = list(
    pageLength = 20,
    dom        = "Bfrtip",
    buttons    = list("csv", "excel"),
    scrollX    = TRUE,
    columnDefs = list(list(className = "dt-right", targets = c(5, 6)))
  )
) |>
  formatCurrency("total_value", currency = "", interval = 3, mark = ",", digits = 0)
```

---

## Notes on methodology

### Data sources
- **Annual file**: `DHIS DATA 2015_2026 (2).xls` — six sheets covering 2015–2026 in wide format (one column per year). The 2022–2026 period is split across three sheets by indicator domain.
- **Monthly file**: `DHIS DATA APR 2024_MAR 2026_test.xls` — four sheets with monthly columns (e.g. "April 2024").

### Processing steps (see `wrangling.R`)
1. Each sheet read with `readxl::read_excel(..., col_types = "text")` to prevent type coercion errors.
2. Fixed columns 1–4 (Province, District, Sub-district, Facility) cleaned by stripping the two-letter DHIS2 province code prefix and trailing administrative suffixes.
3. Value columns melted to long format; values coerced to numeric after removing non-numeric characters.
4. Indicators classified into three domains (births, deaths, contra) via `indicator_meta` lookup table.
5. Monthly sheets de-duplicated on facility + indicator + period after binding.
6. **The 2026 calendar year is dropped from the annual series** (`year <= 2025`) because it is incomplete and would otherwise show a spurious partial-year drop. Monthly data, which is complete per-month through Mar 2026, is retained in the monthly figures.
7. No imputation of missing values; `NA` counts are excluded from aggregations.

### Limitations
- Annual counts in the main file are calendar-year aggregates; months within a year are not individually available before April 2024.
- The annual series ends in 2025; 2026 is excluded as incomplete. The monthly figures extend to Mar 2026.
- The `Inpatient deaths - total` indicator is a separately reported aggregate that may not equal the sum of age-specific death indicators — they are shown together but not summed.
- Facility coverage varies by year and month (see [Reporting completeness](#sec-completeness)). Declining counts are often incomplete reporting, not true declines.
- These data have not been validated against the official MACOD or StatsSA vital registration series.
